---
title: "In-class Exercise 6"
author: "Nor Hendra"
date: "3 October 2026"
date-modified: "last-modified"
format: html
editor: visual
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: true
---

# 1. Getting Started
## 1.1. Installing and Loading R Packages
```{r}
pacman::p_load(sf, sfdep, tmap, 
               plotly, tidyverse, 
               Kendall)
```

## 1.2 The Data

There are two data sets in this use case, they are:

1. Hunan, a geospatial data set in ESRI shapefile format, and
2. Hunan_GDPPC, an attribute data set in csv format.

### 1.2.1 Importing geospatial data
```{r}
hunan <- st_read(
  dsn = "data/geospatial", 
  layer = "Hunan") %>% 
  st_transform(crs = 32650)
```
### 1.2.2 Importing attribute table
```{r}
GDPPC <- read_csv(
  "data/aspatial/Hunan_GDPPC.csv")
```
# 2. Creating a Time Series Cube
```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

Next, `is_spacetime_cube()` of sfdep package will be used to verify if GDPPC_st is indeed an space-time cube object.
```{r}
is_spacetime_cube(GDPPC_st)
```

# 3. Computing Gi*
## 3.1. Deriving the spatial weights
```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wt = st_inverse_distance(nb, 
                             geometry, 
                             scale = 1,
                             alpha = 1),
    .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```
## 3.2. Computing Gi* (unnest)

We can use these new columns to manually calculate the local Gi* for each location. We can do this by grouping by Year and using local_gstar_perm() of sfdep package. After which, we use `unnest()` to unnest gi_star column of the newly created gi_starts data.frame.

Don't need to include `tidyr` and just type in unnest.

```{r}
gi_stars <- GDPPC_nb %>% 
  group_by(Year) %>% 
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>% 
  tidyr::unnest(gi_star)
```

# 4. Mann-Kendall Test
## 4.1. Mann-Kendall Test on Gi*

Below code chunk uses Changsa county
```{r}
cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```

Next, plotting the result by using ggplot2 functions.

```{r}
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```
## 4.2. Interactive Mann-Kendall Plot
Using `ggplotly()` of plotly package
```{r}
p <- ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```
## 4.3. Printing Mann-Kendall test report
```{r}
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>% 
  tidyr::unnest_wider(mk)
```
In the above result, sl is the p-value. With reference to the results, we will reject the hypothesis null and infer that a slight upward trend.

## 4.4. Mann-Kendall test data.frame
Replace this for each location using `group_by()` of dplyr package.
```{r}
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
head(ehsa)
```
### 4.4.1. Show Significant Emerging Hot/Cold Spots
```{r}
emerging <- ehsa %>% 
  arrange(sl, abs(tau)) %>% 
  slice(1:10)
head(emerging)
```
# 5. Performing Emerging Hotspot Analysis
```{r}
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, 
  .var = "GDPPC", 
  k = 1, 
  nsim = 99
)
```

## 5.1. Visualising the distribution of EHSA classes
```{r}
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()
```
## 5.2. Visualising EHSA

Joining hunan and ehsa together:
```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

`tmap` functions:
```{r}
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(fill_alpha = 0.4)
```