---
title: "Hands-on Exercise 1B: Thematic Mapping and GeoVisualisation with R"
author: "Nor Hendra"
date: "23 August 2026"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 5
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
editor: visual  
---

# 1. Overview

Thematic mapping involves the use of map symbols to visualise selected properties of geographic features that are not naturally visible, such as population, temperature, crime rate, and property prices, to name just a few.

Geovisualisation is the process of using visual representations and cartographic techniques to explore, analyse, and communicate geospatial data. It combines elements of cartography, computer science, and information visualisation to enhance spatial understanding and knowledge discovery.

In this hands-on exercise, I learned how to plot functional and truthful choropleth maps by using an R package called **tmap**.

::: callout-note
## Survival tip

I made it a habit to read the functional description of each function before using it. That mattered more than usual here, because tmap version 4 renamed a lot of arguments from version 3, and much of the older material online still uses the version 3 forms.
:::

::: callout-important
## Note on data formats used in this write-up

The reference material uses the Master Plan 2019 Subzone Boundary as a **KML** file, which is why it spends a whole subsection scraping attributes out of an embedded HTML table. What I downloaded from data.gov.sg came as **GeoJSON** instead, and this GeoJSON already carries `SUBZONE_N`, `PLN_AREA_N`, `REGION_N` and `SUBZONE_C` as proper columns. That means the entire HTML-scraping step the reference material performs is unnecessary for my file, so I dropped it. I flagged every place the format changes something.
:::

# 2. Getting Started

The key R package for this exercise is [**tmap**](https://cran.r-project.org/web/packages/tmap/). Alongside it, five other packages are used:

- [**readr**](https://readr.tidyverse.org/) for importing delimited text files,
- [**tidyr**](https://tidyr.tidyverse.org/) for tidying data,
- [**dplyr**](https://dplyr.tidyverse.org/) for wrangling data,
- [**sf**](https://cran.r-project.org/web/packages/sf/index.html) for handling geospatial data, and
- [**rvest**](https://rvest.tidyverse.org/) for scraping (harvesting) data from web pages.

Among these, **readr**, **tidyr** and **dplyr** are part of **tidyverse**, so only tidyverse needs installing rather than each one individually.

```{r}
pacman::p_load(sf, tmap, tidyverse, rvest)
```

I also check which version of tmap is loaded, because the syntax below is written for version 4 and will not run correctly on version 3.

```{r}
packageVersion("tmap")
```

::: callout-warning
tmap version 4 introduced breaking changes to the argument names used throughout this exercise. Under version 3, the equivalents were `tm_polygons(col = ..., style = ..., palette = ...)`. Under version 4 these became `tm_polygons(fill = ..., fill.scale = tm_scale_intervals(style = ..., values = ...))`. If a code chunk below throws an "unused argument" error, a version 3 installation is the first thing to check.
:::

# 3. Importing Data into R

## 3.1 The Data

Two data sets are used to create the choropleth maps:

- **Master Plan 2019 Subzone Boundary (No Sea)**, downloadable at [data.gov.sg](https://data.gov.sg/). This is geospatial data consisting of the geographical boundary of Singapore at planning subzone level, based on the URA Master Plan 2019. I have it as GeoJSON rather than the KML the reference material uses.
- **Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2024** in csv format (`respopagesextod2024.csv`), downloadable from the [Department of Statistics, Singapore](https://www.singstat.gov.sg/). This is aspatial data. It contains no coordinate values, but its `PA` and `SZ` fields serve as unique identifiers to georeference against the Master Plan 2019 Subzone Boundary.

```         
Hands-on_Ex01/
└── data/
    ├── geospatial/
    │   └── MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson
    └── aspatial/
        └── respopagesextod2024.csv
```

::: callout-important
## Correction to the reference material

Note the change of boundary vintage from Hands-on Exercise 1A. That exercise used the Master Plan **2014** boundary. This one uses the Master Plan **2019** boundary.
:::

## 3.2 Importing Geospatial Data into R

`st_read()` of **sf** imports the boundary into R as a simple feature data frame called `mpsz`, with `st_transform()` reprojecting it to svy21 (EPSG:3414) in the same pipe.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson") %>%
  st_transform(crs = 3414)
```

::: callout-note
Unlike the shapefile in Hands-on Exercise 1A, transforming at import is appropriate here. There is no subsequent projection lesson that depends on the layer staying in wgs84, and every downstream operation needs a projected CRS.
:::

### 3.2.1 Checking the attribute columns

First I check what columns came through, so I know whether any tidying is needed.

```{r}
names(mpsz)
```

::: callout-important
## Why the reference material's whole "tidying" step is skipped here

The reference chapter's KML stores its attributes inside an HTML table buried in a single `Description` field, so it has to write a helper function with **rvest** to scrape `REGION_N`, `PLN_AREA_N`, `SUBZONE_N` and `SUBZONE_C` back out as columns:

``` r
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  if (length(value) == 0) NA_character_ else value
}
```

The GeoJSON does not need any of that. As the `names(mpsz)` output shows, it already carries `SUBZONE_N`, `PLN_AREA_N`, `REGION_N` and `SUBZONE_C` as ordinary columns, so there is nothing to scrape. I keep **rvest** loaded in Section 2 only for completeness against the reference material; it is not actually exercised in my version.

The one small schema difference worth noting: my file names its area and length fields `SHAPE.AREA` and `SHAPE.LEN` (the reference material's shapefile used `SHAPE_Leng` / `SHAPE_Area`). I don't use either in this exercise, so it has no effect on the analysis.
:::

Examining the content:

```{r}
mpsz
```

Only the first ten features are displayed. This is sf's default print behaviour for an sf object, which shows a header block of geometry metadata followed by the first ten records rather than the whole table. My file has 332 subzone features, matching the expected MP2019 count.

## 3.3 Importing Attribute Data into R

`read_csv()` of **readr** imports `respopagesextod2024.csv` into a tibble called `popdata`.

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
```

## 3.4 Data Preparation

Before a thematic map can be prepared, I need a data table containing the variables `PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, and `DEPENDENCY`:

- **YOUNG**: age group 0 to 4 through age group 20 to 24,
- **ECONOMY ACTIVE**: age group 25 to 29 through age group 60 to 64,
- **AGED**: age group 65 and above,
- **TOTAL**: all age groups, and
- **DEPENDENCY**: the ratio of young plus aged against the economy active group.

::: callout-important
## Correction to the reference material

The reference chapter instructs the reader to "prepare a data table with year **2020** values," but the file it imports is `respopagesextod2024.csv` and every object it creates is named `popdata2024`. The reference to 2020 is a leftover from an earlier edition. The data being prepared is June **2024**.
:::

### 3.4.1 Data wrangling

The functions used are `pivot_wider()` of **tidyr**, and `mutate()`, `filter()`, `group_by()` and `select()` of **dplyr**.

::: callout-important
## The most serious correction in this exercise

The reference chapter builds `YOUNG`, `ECONOMY ACTIVE` and `AGED` using **positional column indexing** on the pivoted table:

``` r
mutate(YOUNG = rowSums(.[3:6]) + rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11]) + rowSums(.[13:15])) %>%
mutate(`AGED` = rowSums(.[16:21]))
```

Those hard-coded positions are only correct if `pivot_wider()` lays the age-group columns out in one specific order. It does not for my file, and the result is silently wrong.

The problem is that `pivot_wider()` orders the new columns by the **alphabetical** sort of the `AG` values, not their numerical age order. When I actually inspect the pivoted column order (next chunk), `5_to_9` lands at **column 14**, sorted after `45_to_49` and `50_to_54`, because "5" sorts after "4". That single displacement throws off the reference formula in two places:

- `YOUNG` is meant to be ages 0 to 24. The formula `.[3:6] + .[12]` captures `0_to_4, 10_to_14, 15_to_19, 20_to_24` and then, via column 12, `50_to_54`. It **misses `5_to_9` entirely and wrongly folds in 50-to-54-year-olds**.
- `ECONOMY ACTIVE` is meant to be ages 25 to 64. Its ranges pull in column 14, which is `5_to_9`, so **primary-school-age children get counted as working adults**.

Only `AGED` (cols 16 to 21) happens to land correctly. The upshot is that the dependency ratio, the single variable this whole exercise maps, comes out wrong for every subzone.

I fix this by selecting age groups **by name** rather than by position, so it is correct regardless of how `pivot_wider()` happens to order the columns.
:::

First, to make the problem visible, here is the actual column order `pivot_wider()` produces:

```{r}
popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(POP = sum(Pop), .groups = "drop") %>%
  pivot_wider(names_from = AG, values_from = POP) %>%
  names()
```

Note where `5_to_9` sits in that list. It is not adjacent to `0_to_4`.

Now the corrected, name-based wrangling:

```{r}
young_ages <- c("0_to_4", "5_to_9", "10_to_14", "15_to_19", "20_to_24")

econ_ages <- c("25_to_29", "30_to_34", "35_to_39", "40_to_44",
               "45_to_49", "50_to_54", "55_to_59", "60_to_64")

aged_ages <- c("65_to_69", "70_to_74", "75_to_79", "80_to_84",
               "85_to_89", "90_and_over")

popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(POP = sum(Pop), .groups = "drop") %>%
  pivot_wider(names_from = AG, values_from = POP) %>%
  mutate(
    YOUNG            = rowSums(across(all_of(young_ages))),
    `ECONOMY ACTIVE` = rowSums(across(all_of(econ_ages))),
    AGED             = rowSums(across(all_of(aged_ages)))
  ) %>%
  mutate(TOTAL = YOUNG + `ECONOMY ACTIVE` + AGED) %>%
  mutate(DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`) %>%
  select(PA, SZ, YOUNG, `ECONOMY ACTIVE`, AGED, TOTAL, DEPENDENCY)
```

::: callout-note
Two smaller points on top of the main fix:

- `across(all_of(...))` is the current tidyverse for summing a named set of columns. If any named column is missing, it would fail which is exactly the behaviour I want.
- The reference chapter's `summarise()` omits the `.groups` argument, which prints a grouping message and leaves the tibble grouped, so it then calls `ungroup()` on the next line. Passing `.groups = "drop"` does the same thing in one step and suppresses the message.
:::

### 3.4.2 Joining the attribute data and geospatial data

Before the georelational join, the values in `PA` and `SZ` must be converted to uppercase, because those fields contain a mix of upper and lowercase while `SUBZONE_N` and `PLN_AREA_N` are entirely uppercase.

```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ), 
            .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, `left_join()` of **dplyr** joins the geographical data and attribute table using planning subzone name (`SUBZONE_N` and `SZ`) as the common identifier.

```{r}
mpsz_pop2024 <- left_join(mpsz, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

`left_join()` is used with the `mpsz` simple feature data frame as the **left** table specifically so that the output stays a simple features data frame. Reversing the order would return a plain tibble and every subsequent `tm_shape()` call would fail.

Before going on I check how well the join actually matched, since a name-based join across two independently maintained data sets rarely matches perfectly:

```{r}
sum(is.na(mpsz_pop2024$DEPENDENCY))
```

::: callout-important
## An unexplained gap in the reference material

The reference chapter's own `summary()` output later reports NA values in `DEPENDENCY` but never explains where they come from or what to do about them.

They arise from two distinct causes worth separating: subzones that genuinely have no resident population (industrial estates, water catchment areas, port land, military zones), and subzones whose names differ between the URA boundary file and the SingStat population file so the join simply fails to match.

This matters for map interpretation. tmap shades missing values in grey by default, and a reader looking at a grey subzone cannot tell whether it means "nobody lives here" or "the join broke." I identify which subzones are affected in the Extras.
:::

```{r}
#| eval: false
write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```

# 4. Choropleth Mapping Geospatial Data Using tmap

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colours.

::: callout-note
## From the lesson slides

A choropleth map shades areas in proportion to a statistical variable representing an **aggregate summary** of a geographic characteristic within each area. The critical design consequence is that the variable being mapped must be **normalised**, meaning a rate, ratio, or density, not a raw count. Mapping raw counts across areas of unequal size produces a map of area size rather than of the phenomenon. `DEPENDENCY` is already a ratio, which is why it is the appropriate variable to map here.
:::

Two approaches can be used to prepare a thematic map using tmap: plotting quickly with `qtm()`, or plotting a highly customisable map using tmap elements.

## 4.1 Plotting a choropleth map quickly by using qtm()

The quickest way to draw a choropleth map with tmap is `qtm()`. It is concise and provides good default visualisation in many cases.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2024, 
    fill = "DEPENDENCY")
```

`tmap_mode()` with the `"plot"` option produces a static map. `"view"` produces an interactive one. The `fill` argument maps the attribute to be shaded.

## 4.2 Creating a choropleth map by using tmap's elements

The disadvantage of `qtm()` is that it makes the aesthetics of individual layers harder to control. For a high quality cartographic choropleth map, tmap's drawing elements should be used.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(col_alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

::: callout-important
## Correction to the reference material

The reference chapter writes this as `tm_grid(alpha = 0.2)`. In tmap version 4 the bare `alpha` argument was split into role-specific forms, `col_alpha` for line or border colour transparency and `fill_alpha` for fill transparency. Passing `alpha` to `tm_grid()` under version 4 will either warn or be silently ignored depending on the point release. I used `col_alpha`, which is the version 4 equivalent for grid lines.

The same chapter also spells the URA as "Urban Redevelopment Authorithy" in the credits string. Corrected above.
:::

### 4.2.1 Drawing a base map

The basic building block of tmap is `tm_shape()` followed by one or more layer elements such as `tm_polygons()`, `tm_symbols()`, `tm_lines()`, `tm_raster()` and `tm_text()`.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

By default, polygon areas are plotted in light grey (`gray85`) and polygon borders in a slightly darker grey (`gray25`).

### 4.2.2 Drawing a choropleth map using tm_polygons()

To shade by a selected variable, that variable is assigned to `tm_polygons()`.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY")
```

Things to note about `tm_polygons()`:

- The default interval binning is called "pretty".
- The default colour scheme is `blues3` of ColorBrewer.
- Missing values are shaded in grey by default.

### 4.2.3 Drawing a choropleth map using tm_fill() and tm_borders()

`tm_polygons()` is a wrapper of `tm_fill()` and `tm_borders()`. `tm_fill()` shades the polygons using the default colour scheme. `tm_borders()` adds the borders of the polygon features.

Using `tm_fill()` alone:

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY")
```

The subzones are shaded according to their dependency values, but with no boundaries drawn. Adding `tm_borders()`:

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders()
```

Light-grey border lines have now been added. The `fill_alpha` argument defines transparency between 0 (fully transparent) and 1 (opaque). Besides `fill_alpha`, `tm_borders()` takes three other arguments:

- `col` for border colour,
- `lwd` for border line width, default 1, and
- `lty` for border line type, default `"solid"`.

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60", 
             lwd = 0.1, 
             lty = "dashed")
```

## 4.3 Data classification methods of tmap

Most choropleth maps employ some method of data classification. The point of classification is to take a large number of observations and group them into data ranges or classes. tmap provides ten data classification methods: `fixed`, `sd`, `equal`, `pretty` (default), `quantile`, `kmeans`, `hclust`, `bclust`, `fisher`, and `jenks`.

::: callout-note
## From the lesson slides

The choice of classification method is not cosmetic. It changes which spatial pattern the reader sees.

- **Quantile** puts the same number of features in each class. Good for showing rank, but can group very different values together.
- **Equal interval** divides the value range into equally sized classes. The slides caution against this when data is skewed or has large outliers, because outliers produce empty classes.
- **Jenks (natural breaks)** finds natural groupings by statistical optimisation.
- **Standard deviation** is appropriate when the distribution approximates normal.

The slides also note there is no ideal number of classes. More classes means less generalisation but worse legibility, and 4 to 7 is the usual working range.
:::

### 4.3.1 Plotting choropleth maps with built-in classification methods

Quantile classification with 5 classes:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

Equal interval classification:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

The quantile classification is far more evenly distributed than the equal interval one. The reason is visible in the summary statistics further below: `DEPENDENCY` has a much larger maximum than its median, so equal interval spends almost all of its classes on an empty upper range and crushes nearly every subzone into the bottom class. This is exactly the skew case the lesson slides warn about.

### 4.3.2 Plotting a choropleth map with custom breaks

For all built-in styles, the category breaks are computed internally. To override these defaults, breakpoints can be set explicitly with the `breaks` argument of `tm_scale_intervals()`. In tmap, breaks include a minimum and maximum, so to end up with n categories, n+1 elements must be specified in increasing order.

It is good practice to get descriptive statistics on the variable before setting break points.

```{r}
summary(mpsz_pop2024$DEPENDENCY)
```

::: callout-important
## Correction to the reference material

The reference chapter says: "we set break point at 0.60, 0.70, 0.80, and 0.90. In addition, we also need to include a minimum and maximum, which we set at 0 and 100. Our breaks vector is thus c(0, 0.60, 0.70, 0.80, 0.90, 1.00)."

There are two problems here.

**First, the text contradicts itself.** It says the maximum is set at 100, then gives a vector ending at 1.00.

**Second, and more seriously, 1.00 is almost certainly the wrong maximum.** The summary output shows `DEPENDENCY` reaching well above 1.00. Any subzone with a dependency ratio above 1.00 falls outside the break range entirely and is rendered as a missing value, shaded grey, indistinguishable from the subzones that are genuinely NA.

So the reference map silently hides real data and mislabels it as missing. The fix is to close the upper break at or above the observed maximum. I use 100 below, which matches the chapter's stated intent and comfortably covers the real maximum.
:::

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 100))) +
  tm_borders(fill_alpha = 0.5)
```

To make the consequence explicit, here are the two versions side by side. The map on the left uses the reference chapter's break vector, the one on the right uses the corrected version.

```{r}
map_broken <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Upper break at 1.00 (values above are dropped)")

map_fixed <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 100))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Upper break at 100 (all values retained)")

tmap_arrange(map_broken, map_fixed, asp = 1, ncol = 2)
```

Counting exactly how many subzones the reference version loses:

```{r}
sum(mpsz_pop2024$DEPENDENCY > 1.00, na.rm = TRUE)
```

## 4.4 Colour Scheme

tmap supports colour ramps either defined by the user or drawn from the predefined ramps of the **RColorBrewer** package.

### 4.4.1 Using a ColorBrewer palette

The preferred palette is assigned to the `values` argument of `tm_scale_intervals()`.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

To reverse the colour shading, a `-` prefix is added:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

::: callout-note
The reference chapter describes this as assigning "the preferred colour to palette argument of values," which conflates two things. In tmap version 3 the argument was `palette`. In version 4 it is `values`. There is no `palette` argument inside `tm_scale_intervals()` any more, only `values`.
:::

### 4.4.2 Cartographic furniture

Besides map style, tmap provides arguments to draw map furniture such as a compass, scale bar, and grid lines.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, col_alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat", 
             position = c("left", "bottom"))
```

## 4.5 Map Layout

Map layout refers to the combination of all map elements into a cohesive whole: background, frame, typography, scale, aspect ratio, margins, and more.

### 4.5.1 Map legend

tmap provides several legend options to change placement, format and appearance.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5),
              fill.legend = tm_legend(
                title = "Dependency ratio",
                position = tm_pos_in("left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, col_alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat", 
             position = c("left", "bottom"))
```

::: callout-important
## Correction to the reference material

The reference chapter adds `tm_pos_auto_in()` as a standalone map element in the chain:

``` r
tm_shape(mpsz_pop2024) +
  tm_polygons(...) +
  tm_pos_auto_in() +          # <- this line
  tm_borders(fill_alpha = 0.5) +
  ...
```

This is a misuse of the function. `tm_pos_auto_in()` is a **position specifier**, not a map layer. It is meant to be passed to the `position` argument of a component such as `tm_legend()`, not added to the map with `+`. Used as written, it either errors or is silently discarded, and the legend position never changes, which defeats the point of a section titled "Map Legend".

I moved the positioning into `tm_legend(position = ...)` above, where it actually takes effect.
:::

### 4.5.2 Map style

tmap allows a wide variety of layout settings to be changed globally via `tmap_style()`.

```{r}
tmap_style("natural")

tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

Resetting to the default style:

```{r}
tmap_style("white")
```

::: callout-important
## Correction to the reference material

The reference chapter appends `tmap_style("natural")` to the end of the map chain with a `+`:

``` r
tm_shape(mpsz_pop2024) +
  tm_polygons(...) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")        # <- this line
```

`tmap_style()` is **not a tmap element**. It is a global options setter that returns the previous style invisibly. It cannot be added to a map object with `+`. Written this way it either errors, or worse, sets the style as a side effect *after* the map has already been composed, so the style applies to the *next* map rather than this one.

The correct usage is to call it as a standalone statement **before** the map, as I did above, and to reset it afterwards so it does not silently leak into every subsequent chunk in the document. The same error appears again in the reference chapter's small-multiples section.
:::

# 5. Drawing Small Multiple Choropleth Maps

Small multiple maps, also called facet maps, are composed of many maps arranged side by side and sometimes stacked vertically. They allow visualisation of how spatial relationships change with respect to another variable, such as time.

In tmap, small multiples can be plotted in three ways: by assigning multiple values to an aesthetic argument, by creating multiple stand-alone maps with `tmap_arrange()`, or by defining a group-by variable in `tm_facets()`.

## 5.1 By assigning multiple values to an aesthetic argument

```{r}
#| fig-width: 12
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = c("YOUNG", "AGED"),
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom")),
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5)
```

## 5.2 By arranging multiple choropleth maps in a grid layout

```{r}
youngmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"),
                item.height = 0.8),
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of young population")

agedmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"),
                item.height = 0.8),
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of aged population")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

::: callout-warning
Both maps above use `style = "quantile"` computed independently, which means the two legends have **different break values**. They therefore cannot be compared directly by colour. A dark blue on the left map does not represent the same population as a dark blue on the right. This is a well-known trap with small multiples. I address it in the Extras by fixing a shared break vector across both.
:::

## 5.3 By defining a group-by variable in tm_facets()

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile",
            values = "brewer.blues")) +
  tm_facets(by = "REGION_N", 
            nrow = 2, 
            ncol = 3,
            free.coords = TRUE, 
            drop.units = TRUE) +
  tm_borders(fill_alpha = 0.5)
```

::: callout-important
## Correction to the reference material

Two issues in the reference chapter's version of this chunk.

**First**, it passes `ncols = 3` to `tm_facets()`. The argument is `ncol`, singular, consistent with `nrow` immediately above it and with `tmap_arrange()`. `ncols` is not a recognised argument and the intended 2 by 3 grid never materialises.

**Second**, it passes title arguments to `tm_layout()`:

``` r
tm_layout(legend.show = TRUE,
          title.position = c("center", "center"),
          title.size = 20)
```

In tmap version 4, title handling moved out of `tm_layout()` into the dedicated `tm_title()` component. These `title.*` arguments are version 3 holdovers. There is also no title text supplied anywhere in the chunk, so `title.position` and `title.size` are positioning something that does not exist. I dropped them rather than reproduce a no-op.
:::

# 6. Mapping Spatial Objects Meeting a Selection Criterion

Instead of small multiples, `filter()` of **dplyr** can select a geographical area of interest so the choropleth map focuses only on the selected region.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.greens"),
              fill.legend = tm_legend()) +
  tm_borders(fill_alpha = 0.5)
```

# 7. Complementing a Thematic Map with a Statistical Chart

Maps and statistical charts complement each other by representing different aspects of the same data. Maps excel at showing spatial relationships and geographical patterns. Charts effectively display numerical distributions, trends, and comparisons. Combining both supports a more insightful reading of spatial data that also has quantifiable characteristics.

With tmap, a statistical chart can be incorporated into the map visualisation using the `fill.chart` argument of map layers.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.greens"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

::: callout-note
The reference chapter calls this "the `fill.chat` argument." The correct argument name is `fill.chart`.
:::

The visual representation can be improved further by highlighting and labelling the outliers on the choropleth map.

```{r}
mpsz_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- mpsz_selected[mpsz_selected$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.blues"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```

::: callout-warning
The `%in%` matching on floating-point values in `mpsz_selected$DEPENDENCY %in% outlier_vals` is fragile. Floating-point equality is not guaranteed to hold even for values that came from the same source vector. It happens to work here because `outlier_vals` is extracted from that same vector, so the bit patterns are identical, but it is not a pattern to reuse on values computed by a different route. Matching on a subzone identifier would be safer.
:::

# 8. Creating an Interactive Map

Interactive maps let users actively explore data by zooming, panning, and clicking locations for more information. tmap switches between static and interactive maps with `tmap_mode()`.

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")

tm_shape(region_selected, bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c("Name" = "SUBZONE_N",
                         "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2)
```

The interactive map above gives users a lot of freedom to zoom in and out, which risks them losing spatial context entirely. The `set_zoom_limits` argument of `tm_view()` constrains the map extent.

```{r}
tm_shape(region_selected, bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c("Name" = "SUBZONE_N",
                         "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12, 14))
```

```{r}
tmap_mode("plot")
```

::: callout-note
It is important to switch back to `"plot"` mode at the end. `tmap_mode()` sets a global option, so leaving it in `"view"` mode means every subsequent map in the document renders as an interactive Leaflet widget, which bloats the rendered HTML considerably and breaks any static export.
:::

# 9. Extras

## 9.1 Diagnosing the missing values

Section 3.4.2 flagged that the reference chapter reports NA values in `DEPENDENCY` without explaining them. Since tmap renders missing values in the same grey regardless of cause, it is worth knowing which subzones are affected and why.

```{r}
missing_subzones <- mpsz_pop2024 %>%
  st_drop_geometry() %>%
  filter(is.na(DEPENDENCY)) %>%
  select(REGION_N, PLN_AREA_N, SUBZONE_N)

nrow(missing_subzones)
```

```{r}
missing_subzones %>%
  count(PLN_AREA_N, sort = TRUE) %>%
  head(15)
```

The planning areas at the top of this list are recognisable as non-residential land: port, airport, military, catchment, and industrial estates. That is reassuring, because it means the bulk of the NAs are genuine "nobody lives here" cases rather than broken joins.

I can make the distinction explicit on the map by giving missing values their own labelled colour rather than leaving them as ambiguous grey:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues",
                value.na = "#E8E4D9",
                label.na = "No resident population"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_title("Dependency ratio, with missing values labelled explicitly") +
  tm_layout(frame = FALSE)
```

## 9.2 Comparing all classification methods side by side

The lesson slides stress that classification method choice changes what the reader sees. Rather than take that on faith, I plot four methods on identical data and compare.

```{r}
#| fig-height: 9
make_map <- function(style_name, label) {
  tm_shape(mpsz_pop2024) +
    tm_polygons(fill = "DEPENDENCY",
                fill.scale = tm_scale_intervals(
                  style = style_name,
                  n = 5,
                  values = "brewer.yl_or_rd"),
                fill.legend = tm_legend(
                  position = tm_pos_in("left", "bottom"),
                  text.size = 0.5)) +
    tm_borders(col = "white", lwd = 0.4) +
    tm_title(label, size = 0.9)
}

tmap_arrange(
  make_map("quantile", "Quantile"),
  make_map("equal",    "Equal interval"),
  make_map("jenks",    "Jenks / natural breaks"),
  make_map("sd",       "Standard deviation"),
  ncol = 2, nrow = 2
)
```

The four maps use the same variable, the same palette, and the same number of classes, yet they tell noticeably different stories. Equal interval and standard deviation are both dominated by the outlier subzones and flatten almost everything else into one class.

## 9.3 Fixing the shared-scale problem in small multiples

Section 5.2 flagged that `youngmap` and `agedmap` use independently computed quantile breaks, so their colours are not comparable. The fix is a fixed break vector shared by both maps.

```{r}
shared_breaks <- c(0, 2000, 5000, 10000, 20000, 60000)

youngmap_fixed <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG",
              fill.scale = tm_scale_intervals(
                breaks = shared_breaks,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"),
                text.size = 0.5)) +
  tm_borders(col = "white", lwd = 0.4) +
  tm_title("Young population", size = 0.9)

agedmap_fixed <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED",
              fill.scale = tm_scale_intervals(
                breaks = shared_breaks,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"),
                text.size = 0.5)) +
  tm_borders(col = "white", lwd = 0.4) +
  tm_title("Aged population", size = 0.9)

tmap_arrange(youngmap_fixed, agedmap_fixed, asp = 1, ncol = 2)
```

Now a given shade means the same headcount on both maps, and the comparison is legitimate.

## 9.4 Unclassed (continuous) choropleth

The lesson slides distinguish classed from unclassed choropleth maps. tmap supports the latter through `tm_scale_continuous()`, which drops binning entirely and maps values onto a continuous colour ramp.

::: panel-tabset
## Classed

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.pu_bu_gn"),
              fill.legend = tm_legend(title = "Dependency")) +
  tm_borders(col = "white", lwd = 0.4) +
  tm_title("Classed: 5 quantile classes")
```

## Unclassed

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_continuous(
                limits = c(0, 2),
                outliers.trunc = c(TRUE, TRUE),
                values = "brewer.pu_bu_gn"),
              fill.legend = tm_legend(title = "Dependency")) +
  tm_borders(col = "white", lwd = 0.4) +
  tm_title("Unclassed: continuous ramp, truncated at 2.0")
```
:::

The unclassed version preserves every value distinction but is harder to read off precisely, since a reader cannot tell a 0.83 from a 0.87. The classed version generalises but is legible. Which is preferable depends on whether the map is for exploration or for communication.

## 9.5 A custom palette instead of ColorBrewer

`values` accepts an arbitrary vector of colours, not only named ColorBrewer ramps. Below is a custom sequential ramp built from a muted teal-to-ochre progression, which reads more distinctly on a light background than the default blues.

```{r}
custom_ramp <- c("#EDF2F0", "#B8D4CD", "#7FB0A4", "#4A8577", "#2E6F63")

tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "jenks",
                n = 5,
                values = custom_ramp,
                value.na = "#F0EEE8",
                label.na = "No resident population"),
              fill.legend = tm_legend(
                title = "Dependency ratio",
                position = tm_pos_in("left", "bottom"))) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_title("Dependency ratio by planning subzone, custom palette") +
  tm_credits("Source: URA Master Plan 2019 & Singstat June 2024",
             position = c("left", "bottom")) +
  tm_layout(frame = FALSE)
```
