---
title: "Hands-on Exercise 6: Geographical Segmentation with Spatially Constrained Clustering Techniques"
author: "Nor Hendra"
date: "27 September 2026"
date-modified: "last-modified"
format: html
editor: visual
freeze: true
execute:
  eval: true
  echo: true
  warning: false
---

# 1. Learning Outcome

The previous exercises tested whether values cluster in space. This one builds the clusters. The task is to take 55 townships of Shan State, Myanmar, six ICT ownership measures from the 2014 census, and carve the state into six homogeneous regions. Three methods get used on the same data: ordinary hierarchical clustering with `hclust()`, SKATER from **spdep**, and the ClustGeo algorithm, the last two of which force the resulting regions to respect geography.

The interesting question is not which method produces the prettiest map. It is what a spatial constraint costs. A clustering algorithm that ignores geography will always find more homogeneous groups than one that cannot split a region into disconnected pieces, so the comparison worth making is how much within-cluster homogeneity gets traded for how much spatial contiguity. I have tried to answer that with numbers at the end rather than by eye.

# 2. The Data

Two files, in the folder layout I have used for every exercise so far:

- `myanmar_township_boundaries`, an ESRI shapefile of township polygons for the whole of Myanmar, in `data/geospatial/`. It carries 330 townships and 14 attribute fields.
- `Shan-ICT.csv`, an extract of the [2014 Myanmar Population and Housing Census](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at township level, in `data/aspatial/`. 55 rows, 11 fields.

Both come from the [Myanmar Information Management Unit](http://themimu.info/).

# 3. Installing and Loading the R Packages

The reference lists **rgdal** under spatial data handling. That package was archived from CRAN in October 2023 and is not in the `p_load()` call anyway, so nothing needs loading in its place. **sf** and **spdep** cover everything it used to do here.

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

# 4. Data Import and Preparation

## 4.1 Importing geospatial data into R environment

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

The full shapefile reads in as 330 features with 14 fields, WGS 84, geometry type MULTIPOLYGON. After the filter and the column selection I am left with 55 features and 6 fields.

::: callout-note
### Checking the filter before trusting it

`filter(ST %in% ...)` only works if those three strings match the file exactly. The `ST` field holds 18 distinct values, and the three Shan entries are spelled `Shan (East)`, `Shan (North)` and `Shan (South)`, with the spacing and capitalisation the filter expects. They contribute 10, 24 and 21 townships, which sums to the 55 the census file has rows for.
:::

```{r}
shan_sf
```

```{r}
glimpse(shan_sf)
```

## 4.2 Importing aspatial data into R environment

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
```

```{r}
summary(ict)
```

Eleven fields, 55 observations. The spread in `Total households` is the thing to note: from 3,318 in Matman to 82,604 in Taunggyi, a factor of 25. Any raw count in this table is going to be driven by that.

## 4.3 Derive new variables using **dplyr** package

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

```{r}
summary(ict_derived)
```

Six new fields land in positions 12 to 17, which is what the correlation step later relies on. Penetration rates per thousand households now range from 21 to 485 for radio, 116 to 843 for television, and 3 to 92 for computers.

# 5. Exploratory Data Analysis (EDA)

## 5.1 EDA using statistical graphics

```{r, fig.width= 4, fig.height=3}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r, fig.width= 4, fig.height=3}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Raw radio counts are strongly right skewed, skewness 2.59, with three townships beyond the upper whisker: Taunggyi at 30,176, Lashio at 15,307 and Nyaungshwe at 13,801. Those are the three largest townships by household count, which is the whole problem the penetration rate is meant to fix.

```{r, fig.width= 4, fig.height=3}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r, fig.width= 4, fig.height=3}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Dividing by household count changes the shape completely. Skewness falls from 2.59 to 0.48, the distribution is close to symmetric, and only one township sits outside the whiskers: Ywangan at 484.5 radios per thousand households. Ywangan reappears later as a cluster of one, so it is worth remembering.

```{r}
#| fig-width: 10
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

Radio, television and mobile phone are roughly symmetric. Land line, computer and internet are not: their skewness values are 1.4, 1.8 and 1.4, with long right tails driven by a handful of urban townships. That asymmetry matters for the Z-score standardisation later, which assumes something closer to normality than these three offer.

## 5.2 EDA using choropleth map

### 5.2.1 Joining geospatial data with aspatial data

```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, by=c("TS_PCODE"="TS_PCODE"))
  
write_rds(shan_sf, "data/rds/shan_sf.rds")
```

```{r}
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

::: callout-note
### What the join actually produced

All 55 township codes match on both sides, no `NA` values appear in any of the derived columns, and no township name disagrees with its code across the two files. Clean one to one join.

The join also creates duplicate column names, since both tables carry `DT`, `DT_PCODE` and `TS`. **dplyr** suffixes them, so the result has `DT.x` and `DT.y`, `TS.x` and `TS.y`. That is why section 7.1 selects `TS.x` and not `TS`. The full object is 55 rows by 23 columns, with the five clustering variables sitting at positions 17 to 21 and `INTERNET_PR` at 22. Section 10.2 indexes those positions directly.

One practical note: `write_rds()` will not create the folder for you. `data/rds/` has to exist before this chunk runs.
:::

### 5.2.2 Preparing a choropleth map

```{r}
qtm(shan_sf, fill = "RADIO_PR")
```

```{r}
#| fig-width: 10
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_polygons(fill = "TT_HOUSEHOLDS",
              fill.scale = tm_scale_intervals(style = "jenks", n = 5),
              fill.legend = tm_legend(title = "Total households")) + 
  tm_borders(col_alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_polygons(fill = "RADIO",
              fill.scale = tm_scale_intervals(style = "jenks", n = 5),
              fill.legend = tm_legend(title = "Number Radio")) + 
  tm_borders(col_alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

::: callout-tip
### Translating the tmap code

The reference writes these maps in tmap v3 syntax: `tm_fill(col = , n = , style = , title = )`, `tm_borders(alpha = )`, and a standalone `tm_legend()` layer. My environment runs tmap 4, where the classification arguments moved inside `tm_scale_intervals()`, legends are configured through `fill.legend`, and border transparency is `col_alpha` because borders are lines and have no fill. This is the same translation I applied in Exercises 5A and 5B.

`qtm()` survives unchanged except that the variable now goes in explicitly as `fill =`.
:::

The reference reads these two maps as showing that townships with more households also own more radios. That is easy to confirm numerically instead of visually: the correlation between `TT_HOUSEHOLDS` and `RADIO` is 0.935. Against `RADIO_PR` it drops to 0.339. The penetration rate removes most, though not all, of the size effect.

```{r}
#| fig-width: 10
tm_shape(shan_sf) +
    tm_polygons(fill = c("TT_HOUSEHOLDS", "RADIO_PR"),
                fill.scale = tm_scale_intervals(style = "jenks")) +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_layout(outer.margins = 0)
```

The two panels barely resemble each other. Household counts concentrate in a few large townships in the south and along the Lashio corridor. Radio penetration is highest in the western uplands, which are not the populous ones.

# 6. Correlation Analysis

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

`COMPUTER_PR` and `INTERNET_PR` correlate at 0.870, the highest pair in the matrix by a clear margin. The next highest is `MPHONE_PR` with `COMPUTER_PR` at 0.737, then `TV_PR` with `MPHONE_PR` at 0.722. Dropping one of the computer and internet pair is justified.

Radio is the odd variable out. It correlates at -0.310 with land line, -0.069 with mobile phone and 0.080 with computer. Every other variable in the set moves together as a rough proxy for urbanisation. Radio moves against it, or not at all, which means radio ownership is going to do a lot of work in separating the rural clusters later.

# 7. Hierarchy Cluster Analysis

## 7.1 Extracting clustering variables

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

`INTERNET_PR` is out, on the strength of that 0.870 correlation with `COMPUTER_PR`.

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

```{r}
write_rds(shan_ict, "data/rds/shan_ict.rds")
```

```{r}
shan_ict <- read_rds("data/rds/shan_ict.rds")
```

## 7.2 Data Standardisation

### 7.2.1 Min-Max standardisation

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

### 7.2.2 Z-score standardisation

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

Means of 0 and standard deviations of 1, as expected. The skewness column is the useful part: `LLPHONE_PR` at 1.37 and `COMPUTER_PR` at 1.80 are not remotely normal, and the warning in the reference about Z-scores assuming normality applies to exactly those two.

### 7.2.3 Visualising the standardised clustering variables

```{r}
#| fig-width: 12
#| fig-height: 4
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

```{r}
#| fig-width: 12
#| fig-height: 4
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

The shape of the distribution does not change across the three panels. Only the axis does. Both standardisations are linear transformations, so they can move a variable's scale but not its skew.

## 7.3 Computing proximity matrix

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
```

Distances run from 38.5 to 995.7 in penetration-rate units.

```{r}
#| eval: false
proxmat
```

## 7.4 Computing hierarchical clustering

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

```{r}
#| fig-width: 10
#| fig-height: 6
plot(hclust_ward, cex = 0.6)
```

## 7.5 Selecting the optimal clustering algorithm

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

Ward comes out highest at 0.943, then complete at 0.895, average at 0.813 and single at 0.663. The reference reads this as Ward giving the strongest clustering structure, and on this metric it does. Section 11.3 checks the same question with a second diagnostic that disagrees.

## 7.6 Determining Optimal Clusters

### 7.6.1 Gap Statistic Method

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

```{r}
fviz_gap_stat(gap_stat)
```

::: callout-warning
### The 6-cluster recommendation is thinner than it looks

`firstmax` returns 1, which is not a usable answer. The Hands-on reference then picks 6 on the grounds that it gives the largest gap statistic among the remaining options. Checking the printed table, the gaps for k = 5 and k = 6 are 0.22156 and 0.22218. The difference is 0.0006. The standard error on each of those estimates is about 0.039.

So 6 beats 5 by roughly one sixtieth of a standard error. It is the maximum, technically, but the gap statistic is not the thing making the decision here.
:::

## 7.7 Interpreting the dendrograms

```{r}
#| fig-width: 10
#| fig-height: 6
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

## 7.8 Visually-driven hierarchical clustering analysis

### 7.8.1 Transforming the data frame into a matrix

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

### 7.8.2 Plotting interactive cluster heatmap using *heatmaply()*

```{r}
#| fig-width: 8
#| fig-height: 6
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

I notice that this heatmap normalises the matrix first, unlike the dendrogram above it, so the row ordering here comes from a different distance matrix than `proxmat`.

## 7.9 Mapping the clusters formed

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

```{r}
qtm(shan_sf_cluster, fill = "CLUSTER")
```

Cluster sizes are 20, 4, 13, 6, 7 and 5.Counting properly against the contiguity graph built in section 8.2, those six clusters break into 24 spatially disconnected pieces: 4, 4, 5, 5, 5 and 1. The four-township cluster sits in four separate places, one township each. Only cluster 6 forms a single block. That is the number the constrained methods have to beat.

# 8. Spatially Constrained Clustering: SKATER approach

## 8.1 Converting into SpatialPolygonsDataFrame

```{r}
shan_sp <- as_Spatial(shan_sf)
```

## 8.2 Computing Neighbour List

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

55 regions, 264 non-zero links, 8.73% of the full matrix, an average of 4.8 neighbours per township. Five townships have only 2 neighbours and one has 9.

```{r}
coords <- st_coordinates(
  st_centroid(st_geometry(shan_sf)))
```

```{r}
plot(st_geometry(shan_sf), 
     border=grey(.5))
plot(shan.nb,
     coords, 
     col="blue", 
     add=TRUE)
```

## 8.3 Computing minimum spanning tree

### 8.3.1 Calculating edge costs

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
```

Edge costs range from 47.8 to 721.9. They are attribute distances, not geographic ones: two adjacent townships with similar ICT profiles get a cheap edge regardless of how far apart their centroids sit.

```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

### 8.3.2 Computing minimum spanning tree

```{r}
shan.mst <- mstree(shan.w)
```

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

54 rows and 3 columns. One edge fewer than the 55 nodes, as a spanning tree requires, with the third column holding the cost. The tree's total cost is 9,686.6, down from the 264 edges the contiguity graph started with.

```{r}
head(shan.mst)
```

```{r}
plot(st_geometry(shan_sf), 
                 border=gray(.5))
plot.mst(shan.mst, 
         coords, 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

## 8.4 Computing spatially constrained clusters using SKATER method

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

```{r}
str(clust6)
```

```{r}
ccs6 <- clust6$groups
ccs6
```

```{r}
table(ccs6)
```

::: callout-warning
### Two clusters of one township

The group sizes come out as 22, 18, 11, 2, 1 and 1. The reference describes the first cluster as having 12 observations, which doesn't match what my data produces. I couldn't tell whether that came from a different underlying dataset, an older package version, or something else in the setup perhaps.
:::

`ssto` is 12,612.5 and the within-group sum of squares falls to 8,735.2 after five cuts.

```{r}
#| fig-width: 4
#| fig-height: 6
plot(st_geometry(shan_sf), 
     border=gray(.5))
plot(clust6, 
     coords, 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005, 
     add=TRUE)
```

Two small things about this chunk, both harmless. `groups.colors` is given five colours for six groups, and `add = TRUE` is not an argument of `plot.skater()` at all, so it falls through to `plot()` and raises "add is not a graphical parameter". I surpressed the warning as warning: false and the pruned tree still draws over the township boundaries correctly.

## 8.5 Visualising the clusters in choropleth map

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, fill = "SP_CLUSTER")
```

```{r}
#| fig-width: 10
#| fig-height: 6
hclust.map <- qtm(shan_sf_cluster,
                  fill = "CLUSTER") + 
  tm_borders(col_alpha = 0.5) 

shclust.map <- qtm(shan_sf_spatialcluster,
                   fill = "SP_CLUSTER") + 
  tm_borders(col_alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```

Side by side, the difference is exactly what the constraint promises. The hierarchical map scatters each cluster across the state in 24 separate pieces. The SKATER map has 6 clusters in 6 pieces, one contiguous block each, by construction.

# 9. Spatially Constrained Clustering: ClustGeo Method

## 9.1 A short note about ClustGeo package

ClustGeo works from two dissimilarity matrices, D0 in the attribute space and D1 in the constraint space, combined at each merge step through a mixing parameter alpha between 0 and 1 (Chavent et al., 2018). At alpha = 0 geography is ignored entirely. At alpha = 1 only geography counts. The aim is to find a value that buys spatial coherence without giving away too much attribute homogeneity, which `choicealpha()` helps with.

Unlike SKATER, this is a soft constraint. Nothing in the algorithm guarantees the resulting clusters are contiguous, it only makes distant townships expensive to merge.

## 9.2 Ward-like hierarchical clustering: ClustGeo

```{r}
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, 
            k = 6, 
            border = 2:5)
```

### 9.2.1 Mapping the clusters formed

```{r}
groups <- as.factor(cutree(nongeo_cluster, k=6))
```

```{r}
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_ngeo_cluster, fill = "CLUSTER")
```

Still fragmented, and slightly more so than section 7.9: 28 disconnected pieces across the six clusters, against 24 for the ward.D solution.

## 9.3 Spatially Constrained Hierarchical Clustering

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

`st_distance()` on a geographic CRS returns great circle distances in metres through **s2**, not degrees, so the constraint space is in sensible units without reprojecting. The largest township-to-township distance in the state is 444 km. `as.dist()` strips the units class and returns a plain `dist` object, which is what `choicealpha()` needs.

```{r}
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K=6, graph = TRUE)
```

::: callout-tip
### Reading the alpha plots as numbers

The two curves are easier to judge from `cr$Q` and `cr$Qnorm` than from the plot. At alpha = 0, the proportion of explained pseudo-inertia is 0.757 in the attribute space and 0.131 in the constraint space. At alpha = 1 the constraint space reaches 0.941.

At alpha = 0.2, attribute homogeneity drops to 0.647 and geographic homogeneity rises to 0.626. That is a loss of 0.11 and a gain of 0.50, which matches the figures quoted in the Lesson 6 slides for this dataset. Pushing to alpha = 0.3 costs another 0.02 of attribute quality for 0.07 more geographic coherence, so the returns are already flattening. 0.2 sits just past the crossover of the two curves and is a defensible pick.

One curiosity in the table: alpha = 0.5 and alpha = 0.6 produce identical Q values. The criterion changes continuously with alpha but the resulting tree does not, so a range of alpha values can give the same partition.
:::

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
```

```{r}
groups <- as.factor(cutree(clustG, k=6))
```

```{r}
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_Gcluster, fill = "CLUSTER")
```

The sizes are the most balanced of the three methods: 10, 9, 7, 9, 12 and 8. No singletons, no cluster holding a third of the state. The map is visibly more regionalised than section 9.2.1 without being forced into single blocks, and counting pieces confirms the middle position: 18 in total, against 28 unconstrained and 6 for SKATER. Two of the six are single blocks, two more are in two pieces, and one still scatters across seven. The soft constraint does what it claims, and it does not pretend to guarantee contiguity.

# 10. Visual Interpretation of Clusters

## 10.1 Visualising individual clustering variable

```{r}
ggplot(data = shan_sf_ngeo_cluster,
       aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

Ranking the clusters by mean radio penetration gives 3 \> 2 \> 1 \> 4 \> 5 \> 6. The hands-on showed ranking as 3, 2, 1, 4, 6, 5. The first four agree, and the last two are the other way round in my run: cluster 6 averages 98.6 radios per thousand and cluster 5 averages 124.0.

## 10.2 Multivariate Visualisation

```{r}
#| fig-width: 12
ggparcoord(data = shan_sf_ngeo_cluster, 
           columns = c(17:21), 
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30))
```

Positions 17 to 21 are `RADIO_PR`, `TV_PR`, `LLPHONE_PR`, `MPHONE_PR` and `COMPUTER_PR`, which I checked back in section 5.2.1. Hard-coded column indices are fragile, and this one only holds because the join produced exactly the column order it did.

```{r}
shan_sf_ngeo_cluster %>% 
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```

# 11. Extras

## 11.1 The two methods the chapter names but never runs

Section 7.6 lists elbow, average silhouette and gap statistic as the three common ways to choose k, but only runs the gap statistic. NbClust gets loaded in section 3 and mentioned in a note, but never actually called. Since the choice of six clusters drives every map in this document, I wanted to see what the other two methods say.

```{r}
fviz_nbclust(shan_ict, hcut, method = "wss", k.max = 10)
```

```{r}
fviz_nbclust(shan_ict, hcut, method = "silhouette", k.max = 10)
```

```{r}
#| results: hide
#| fig-show: hide
set.seed(12345)
nbc <- NbClust(shan_ict, distance = "euclidean",
               min.nc = 2, max.nc = 10, method = "ward.D", index = "all")
```

```{r}
table(nbc$Best.nc["Number_clusters",])
```

The elbow curve bends hardest between k = 2 and k = 3, where within-cluster sum of squares falls from 3.75 million to 2.12 million and then 1.67 million. From k = 5 onward each extra cluster buys between 70,000 and 150,000, and there is no clear elbow at 6.

Average silhouette width peaks at k = 2 with 0.347 and never recovers. At k = 6 it is 0.253, the lowest value across the whole range tested. NbClust runs 26 indices and votes 9 for k = 2, 6 for k = 4, 3 for k = 5, 2 each for k = 3 and k = 10, and exactly 1 for k = 6, with three indices returning 0 or 1 and abstaining in effect.

## 11.2 Would standardising have changed the answer?

Section 7.2 raised this: the chapter standardises two ways and then clusters the raw values, where mobile phone and television carry 81% of the variance and computer carries 0.6%. Running the same ward.D clustering on each version answers how much that matters.

```{r}
g_raw <- cutree(hclust(dist(shan_ict), method = "ward.D"), 6)
g_z   <- cutree(hclust(dist(scale(shan_ict)), method = "ward.D"), 6)
g_mm  <- cutree(hclust(dist(normalize(shan_ict)), method = "ward.D"), 6)

rbind(raw = table(g_raw), z_score = table(g_z), min_max = table(g_mm))
```

```{r}
table(raw = g_raw, z_score = g_z)
```

```{r}
adj_rand <- function(a, b) {
  t <- table(a, b); n <- sum(t)
  s1 <- sum(choose(t, 2))
  s2 <- sum(choose(rowSums(t), 2)); s3 <- sum(choose(colSums(t), 2))
  expected <- s2 * s3 / choose(n, 2)
  (s1 - expected) / ((s2 + s3) / 2 - expected)
}
c(raw_vs_z = adj_rand(g_raw, g_z),
  raw_vs_minmax = adj_rand(g_raw, g_mm),
  z_vs_minmax = adj_rand(g_z, g_mm))
```

The adjusted Rand index between the raw and Z-score solutions is 0.30. For reference, 0 is what two independent random partitions score and 1 is identical. Raw against min-max is 0.42. The two standardised solutions agree with each other at 0.79, which is the useful part: they disagree with the raw solution far more than they disagree with each other.

The sizes tell the same story. Raw gives 20, 4, 13, 6, 7, 5 with one cluster holding over a third of the state. Z-scores give 11, 12, 9, 8, 9, 6, much more even, because land line and computer stop being rounding errors in the distance calculation and start separating townships.

## 11.3 What the spatial constraint actually costs

The reference compares the hierarchical and SKATER maps visually and says one is fragmented. Both halves of that trade-off can be measured: within-cluster sum of squares for homogeneity, and the number of connected pieces each cluster breaks into for contiguity.

```{r}
g_hclust   <- cutree(hclust_ward, 6)
g_ngeo     <- cutree(nongeo_cluster, 6)
g_skater   <- clust6$groups
g_clustgeo <- cutree(clustG, 6)

pieces <- function(g) {
  sum(sapply(unique(g), function(k)
    n.comp.nb(subset.nb(shan.nb, seq_along(g) %in% which(g == k)))$nc))
}

wss <- function(g, dat) {
  sum(sapply(unique(g), function(k) {
    x <- dat[g == k, , drop = FALSE]
    if (nrow(x) < 2) return(0)
    sum(scale(x, scale = FALSE)^2)
  }))
}

X <- as.matrix(shan_ict)
total_ss <- sum(scale(X, scale = FALSE)^2)

data.frame(
  solution = c("hclust ward.D", "hclustgeo alpha=0", "SKATER", "ClustGeo alpha=0.2"),
  sizes = sapply(list(g_hclust, g_ngeo, g_skater, g_clustgeo),
                 function(g) paste(sort(table(g), decreasing = TRUE), collapse = "/")),
  wss_pct = round(100 * sapply(list(g_hclust, g_ngeo, g_skater, g_clustgeo),
                               wss, dat = X) / total_ss, 1),
  spatial_pieces = sapply(list(g_hclust, g_ngeo, g_skater, g_clustgeo), pieces)
)
```

| Solution           | Sizes          | Within-cluster SS | Pieces |
|--------------------|----------------|-------------------|--------|
| hclust ward.D      | 20/13/7/6/5/4  | 23.9%             | 24     |
| hclustgeo alpha=0  | 16/13/8/7/6/5  | 24.3%             | 28     |
| SKATER             | 22/18/11/2/1/1 | 50.0%             | 6      |
| ClustGeo alpha=0.2 | 12/10/9/9/8/7  | 35.3%             | 18     |

SKATER is the expensive one. Perfect contiguity, 6 pieces for 6 clusters, at the cost of doubling unexplained within-cluster variation from 24% to 50%. Part of that is the two singleton clusters, which contribute nothing to homogeneity while using up two of the six slots.

ClustGeo at alpha = 0.2 lands in between on both axes: 35% within-cluster sum of squares and 18 pieces. It is the reasonable middle position, and it is worth being clear that its 18 pieces mean it has not delivered regions in the strict sense. Six contiguous regions is SKATER's guarantee and nobody else's.

The choice follows from what the output is for. A policy map that has to name contiguous administrative regions needs SKATER or something like it, and should accept the homogeneity cost.

## 11.4 Fixing SKATER's singletons with `crit`

Section 8.4 produced clusters of 22, 18, 11, 2, 1 and 1. `skater()` takes a `crit` argument that sets a minimum group size and refuses cuts that violate it, which the reference never mentions. The question is what enforcing a floor costs.

```{r}
for (cr in c(0, 3, 5, 8)) {
  cl <- if (cr == 0) {
    spdep::skater(shan.mst[,1:2], shan_ict, method = "euclidean", ncuts = 5)
  } else {
    spdep::skater(shan.mst[,1:2], shan_ict, method = "euclidean", ncuts = 5, crit = cr)
  }
  cat("crit =", cr,
      "| clusters:", length(unique(cl$groups)),
      "| sizes:", paste(sort(table(cl$groups), decreasing = TRUE), collapse = "/"),
      "| final ssw:", round(tail(cl$ssw, 1), 1), "\n")
}
```

Almost nothing, as it turns out. A floor of 3 gives 15, 12, 12, 8, 4, 4 with a final within-group sum of squares of 8,772.7 against 8,735.2 unconstrained, which is 0.4% worse. A floor of 5 gives 14, 12, 11, 8, 5, 5 at 8,828.8, or 1.1% worse. Both still produce six fully contiguous regions.

A floor of 8 breaks it: only five groups come out, because after four cuts no remaining edge can be removed without leaving a subgraph under the minimum.

For a segmentation that is meant to support planning, trading 0.4% of homogeneity to avoid two regions of one township each looks like a good deal maybe?

# Summary of Adjustments

| Location in reference | Issue found | Correction applied |
|----|----|----|
| 12.5.2.2, 12.8.6 (all `tmap` calls) | Written in tmap v3 syntax: `tm_fill(col=, n=, style=, title=)`, `tm_borders(alpha=)`, standalone `tm_legend()`, `qtm()` with a positional variable | Translated to tmap v4: `tm_polygons(fill=, fill.scale=tm_scale_intervals(), fill.legend=tm_legend())`, `tm_borders(col_alpha=)`, `qtm(fill=)`. Border transparency is `col_alpha` because borders are lines. Same translation applied in Exercises 5A and 5B |
| 12.8.5 (`plot(clust6, ..., add = TRUE)`) | `add` is not an argument of `plot.skater()`, so it passes through to `plot()` and raises "add is not a graphical parameter". `groups.colors` also supplies five colours for six groups | Left the code as written. The figure still draws over the boundaries correctly and the warning is suppressed by the YAML, noted so it is not a surprise when run interactively |
| 12.8.5 (SKATER cluster sizes) | Prose describes the first cluster as having 12 observations. This data gives 22, 18, 11, 2, 1, 1, including two single-township clusters | Reported the sizes this data produces, identified the singletons as Ywangan and Narphan, and explored `skater(crit=)` as a fix in section 11.5 |
| 12.10.1 (boxplot reading) | Cluster ranking by mean radio penetration given as 3, 2, 1, 4, 6, 5 | On this data the order is 3, 2, 1, 4, 5, 6. |

# References

- Assunção, R. M., Neves, M. C., Câmara, G., & Costa Freitas, C. D. (2006). Efficient regionalization techniques for socio-economic geographical units using minimum spanning trees. *International Journal of Geographical Information Science*, 20(7), 797-811.
- Chavent, M., Kuentz-Simonet, V., Labenne, A., & Saracco, J. (2018). ClustGeo: an R package for hierarchical clustering with spatial constraints. *Computational Statistics*, 33(4), 1799-1822.
- Kam, T. S. (2026). *ISSS626 Lesson 6: Geographic Segmentation with Spatial Clustering* \[Lecture slides\]. Singapore Management University.
