---
title: "Hands-on Exercise 5A: Global Measures of Spatial Autocorrelation"
author: "Nor Hendra"
date: "20 September 2026"
date-modified: "last-modified"
format: html
editor: visual
freeze: true
execute:
  eval: true
  echo: true
  warning: false
---

# 1. Learning Outcome

This exercise moves from spatial weights, which I built in the previous chapter, into actually using them. The question here is simple to state and harder to answer well: is GDP per capita spread evenly across Hunan's counties, or does it cluster? I am using **spdep** to compute Moran's I and Geary's C, both as analytical tests and as Monte Carlo permutations, and then a spatial correlogram to see how the clustering behaves as distance between counties increases.

By the end of this I should be able to say, with a statistically defensible answer, whether Hunan's GDPPC in 2012 shows spatial randomness or spatial structure, and if structure, roughly how far that structure reaches.

# 2. The Data

Two files, same as Hands-on Exercise 4:

- Hunan county-level administrative boundary, ESRI shapefile format.
- `Hunan_2012.csv`, containing GDPPC and other development indicators for 2012.

# 3. Installing and Loading the R Packages

Four packages this time: `sf` for the geospatial import, `spdep` for the weights and autocorrelation statistics, `tmap` for the choropleth, `tidyverse` for everything else.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

# 4. Getting the Data Into R

## 4.1 Importing the shapefile

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```

88 features, 7 fields, WGS 84. Matches what the reference reports for the same file.

## 4.2 Importing the csv

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

88 rows, one per county, with `GDPPC` sitting as the ninth column alongside wages, deposits, fixed asset investment and the rest.

## 4.3 Joining the two

```{r}
hunan <- left_join(hunan, hunan2012) %>%
  select(1:4, 7, 15)
```

## 4.4 Visualising GDPPC

```{r}
#| fig-width: 12
#| fig-height: 8
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(legend.position = c("left", "bottom"),
            main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_layout(legend.position = c("left", "bottom"),
            main.title = "Quantile classification")

tmap_arrange(equal,
             quantile,
             asp = 1,
             ncol = 2)
```
::: callout-note
### My Findings
The quantile map spreads colour more evenly across the province since each bucket holds the same number of counties. The equal-interval map is more honest about the actual GDPPC gap: most counties sit in the lowest band and a handful of outliers, Changsha and its immediate neighbours, pull the breaks wide, so those outliers end up carrying most of the visual weight.
:::

# 5. Global Measures of Spatial Autocorrelation

## 5.1 Computing Contiguity Spatial Weights

```{r}
wm_q <- poly2nb(hunan,
                queen = TRUE)
summary(wm_q)
```

88 regions, 448 nonzero links, average of just over 5 neighbours per county. Two counties (index 30 and 65, Jinshi and Shaoshan by position) have only one neighbour each, and one county has 11. Matches the reference exactly.

## 5.2 Row-Standardised Weights Matrix

```{r}
rswm_q <- nb2listw(wm_q,
                   style = "W",
                   zero.policy = TRUE)
rswm_q
```

::: callout-note
### What this actually does

`nb2listw()` needs an `nb` object as input. `style = "W"` row-standardises, meaning each neighbour gets weight 1/(number of neighbours), so every county's row of weights sums to 1. The alternative, `style = "B"`, just uses binary 0/1 weights without that normalisation. `zero.policy = TRUE` matters here because it lets counties with no neighbours (none in this data, but worth knowing) get a zero-length weight vector instead of throwing an error.
:::

## 5.3 Moran's I

### 5.3.1 The test

```{r}
moran.test(hunan$GDPPC,
           listw = rswm_q,
           zero.policy = TRUE,
           na.action = na.omit)
```

Moran's I comes out at 0.301, well above the expected value of -0.011 under spatial randomness, with a standardised deviate of 4.74 and a p-value under 0.0001. That's a clear rejection of the null. GDPPC in Hunan is not scattered at random across counties, it clusters.

### 5.3.2 Monte Carlo Moran's I

An analytical test assumes normality or randomisation conditions that may not hold perfectly for 88 irregular polygons, so it's worth checking the same question with a permutation approach that doesn't lean on that assumption.

```{r}
set.seed(1234)
bperm <- moran.mc(hunan$GDPPC,
                listw = rswm_q,
                nsim = 999,
                zero.policy = TRUE,
                na.action = na.omit)
bperm
```

::: callout-note
### What can I conclude?
Same conclusion from a different angle. The observed statistic of 0.301 ranks 1000th out of 1000 (999 simulations plus the observed value), giving a permutation p-value of 0.001, the smallest this test can report at this simulation count. The two methods agree.
:::

### 5.3.3 Visualising the Monte Carlo distribution

```{r}
mean(bperm$res[1:999])
var(bperm$res[1:999])
summary(bperm$res[1:999])
hist(bperm$res,
     freq = TRUE,
     breaks = 20,
     xlab = "Simulated Moran's I")
abline(v = 0,
       col = "red")
```

::: callout-note

### Observation
The simulated values centre near zero (mean -0.015), as they should under the null of no spatial pattern, and the observed 0.301 sits well out past the right tail, clear of the whole simulated distribution.
::: 

::: callout-tip
The reference leaves plotting this in `ggplot2` as a reader challenge instead of doing it. I've picked that up properly in the Extras section rather than half-answering it here.
:::

## 5.4 Geary's C

### 5.4.1 The test

```{r}
geary.test(hunan$GDPPC, listw = rswm_q)
```

::: callout-note
### Conclusion
Geary's C runs on a different scale to Moran's I: 1 means no autocorrelation, values below 1 indicate positive autocorrelation (similar values cluster together), above 1 indicates negative autocorrelation. Here C = 0.691, meaningfully below 1, with a standardised deviate of 3.61 and p = 0.00015. Same conclusion as Moran's I: clustering, not randomness.
:::

### 5.4.2 Monte Carlo Geary's C

```{r}
set.seed(1234)
bperm = geary.mc(hunan$GDPPC,
               listw = rswm_q,
               nsim = 999)
bperm
```

::: callout-note
### Conclusion
Observed rank 1 out of 1000 this time, since Geary's C reads in the opposite direction to Moran's I (low values are the interesting tail here, not high ones). Same p-value of 0.001, same conclusion.
:::

### 5.4.3 Visualising the Monte Carlo distribution

```{r}
mean(bperm$res[1:999])
var(bperm$res[1:999])
summary(bperm$res[1:999])
hist(bperm$res, freq = TRUE, breaks = 20, xlab = "Simulated Geary c")
abline(v = 1, col = "red")
```

::: callout-note 
### Conclusion
Simulated values centre close to 1 (mean 1.004), and the observed 0.691 sits well below that whole distribution. Consistent with the analytical test above.
:::

## 5.5 Spatial Correlogram

A single Moran's I or Geary's C number tells me whether there's clustering overall, but not how far it reaches. A correlogram answers that by recomputing the statistic at increasing neighbour lags, so lag 1 is immediate neighbours, lag 2 is neighbours-of-neighbours not already counted at lag 1, and so on.

### 5.5.1 Moran's I correlogram

```{r}
MI_corr <- sp.correlogram(wm_q,
                          hunan$GDPPC,
                          order = 6,
                          method = "I",
                          style = "W")
plot(MI_corr)
```

```{r}
print(MI_corr)
```

::: callout-note
### Conclusion
Positive and significant at lags 1 and 2, fading through lag 3 and 4, then flipping to significantly negative at lags 5 and 6. That flip is the signature of a province with a rich core (Changsha and its neighbours) surrounded by a poorer periphery: counties close to the core resemble each other, counties far from it resemble each other, and the two groups pull apart from each other at that distance.
:::

### 5.5.2 Geary's C correlogram

```{r}
GC_corr <- sp.correlogram(wm_q,
                          hunan$GDPPC,
                          order = 6,
                          method = "C",
                          style = "W")
plot(GC_corr)
```

```{r}
print(GC_corr)
```

::: callout-note
### Conclusion
Reads the same story from the other statistic: significant clustering (C below 1) at lags 1 and 2, and a significant swing toward dissimilarity (C above 1) at lag 5. Lags 3, 4 and 6 don't clear significance either way. Moran's I and Geary's C don't always land on exactly the same lags as significant, which makes sense given they weight deviations differently, but the overall shape they trace is the same.
::: 

# 6. Extras

## 6.1 Answering the reference's own ggplot2 challenge

Section 9.5.3 of the reference poses this as a challenge rather than doing it: plot the simulated Moran's I values with `ggplot2` instead of base graphics.

```{r}
set.seed(1234)
bperm_gg <- moran.mc(hunan$GDPPC, listw = rswm_q, nsim = 999,
                      zero.policy = TRUE, na.action = na.omit)

ggplot(data.frame(sim = bperm_gg$res[1:999]), aes(x = sim)) +
  geom_histogram(bins = 20, fill = "grey70", color = "white") +
  geom_vline(xintercept = 0, color = "red") +
  labs(x = "Simulated Moran's I", y = "Count",
       title = "Monte Carlo simulation of Moran's I (ggplot2)") +
  theme_minimal()
```

Same distribution, same conclusion, just cleaner axis handling and no need to call `abline()` as a separate step.

## 6.2 Row-standardised versus binary weights: does the coding scheme change the verdict?

The reference mentions in passing that `style = "B"` (binary weights) is available as a more robust alternative to `style = "W"`. I'm checking whether the choice changes anything that matters.

```{r}
wm_B <- nb2listw(wm_q, style = "B", zero.policy = TRUE)
moran.test(hunan$GDPPC, listw = wm_B, zero.policy = TRUE, na.action = na.omit)
```

I comes out at 0.309 under binary weighting versus 0.301 under row standardisation, and the standardised deviate actually goes up slightly, 5.12 versus 4.74. Direction and significance both hold either way. For this data set, the coding scheme shifts the exact number a little but doesn't change the conclusion. That's reassuring, since it means the earlier result isn't an artefact of picking "W" for convenience.

## 6.3 Queen versus rook contiguity: how much does the neighbour rule matter?

`poly2nb()` defaults to queen contiguity (shared edge or corner counts as neighbouring). Rook contiguity (`queen = FALSE`) requires a shared edge, corners alone don't count. Given how irregular Hunan's county boundaries are, I wanted to see how much that distinction actually changes.

```{r}
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```

```{r}
rswm_r <- nb2listw(wm_r, style = "W", zero.policy = TRUE)
moran.test(hunan$GDPPC, listw = rswm_r, zero.policy = TRUE, na.action = na.omit)
```

Rook contiguity drops the link count from 448 to 440, eight fewer neighbour pairs that only touched at a corner. Moran's I moves from 0.301 to 0.308, again in the same direction with comparable significance. The two contiguity rules are near-substitutes here, which isn't guaranteed in general, some geographies have enough corner-touching polygons that the rule genuinely matters, but for Hunan's counties it doesn't swing the result.

## 6.4 Pushing the correlogram past six lags

The reference stops the correlogram at order 6 without saying why that particular cutoff. I wanted to see what happens further out, and whether the sample size behind each lag holds up.

```{r}
MI_corr10 <- sp.correlogram(wm_q, hunan$GDPPC, order = 10, method = "I",
                             style = "W", zero.policy = TRUE)
plot(MI_corr10)
```

```{r}
print(MI_corr10)
```

The number of counties actually contributing to each lag (shown in parentheses next to the lag number) drops off fast: 88 at lag 6, down to 67 at lag 8 and just 19 by lag 10. That's not a flaw in the method, it's simply how many county pairs are still that many steps apart in a network with only 88 nodes and an average of 5 neighbours each. The error bars widen accordingly and neither lag 9 nor lag 10 reaches significance. Six lags turns out to be a reasonable place to stop, not an arbitrary one: past that point the correlogram is estimating from a shrinking, noisier subset of the province rather than telling me anything new about the spatial structure.



