---
title: "In-class Exercise 5"
author: "Nor Hendra"
date: "26 September 2026"
date-modified: "last-modified"
format: html
editor: visual
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: true
---

# 1. Installing and Loading the R Packages

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

# 2. Preparing the Data

::: panel-tabset
## 2.1 Loading the Data

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

## 2.2 Importing the csv file

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## 2.3 Performing the relational join

```{r}
hunan_GDPCC <- left_join(hunan, hunan2012) %>%
  dplyr::select(1:4, 7, 15, 16)
```
:::

# 3. Global Measures of Spatial Association 
## 3.1 Step 1: Deriving Queen's continuity weights: sfdep methods

```{r}
wm_q <- hunan_GDPCC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)
```

## 3.2 Performing Moran's I

### 3.2.1 Step 2A: Computing Global Moran's I

`global_moran()`  function is used to compute the Moran's I value. Different from sdep package, the output is a tibble data.frame. 

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
glimpse(moranI)
```
### 3.2.2 Step 2B: Performing Global Moran's I test

Moran's I test will be performed instead of just computing the Moran's I statistics. With sfdep package, Moran's I test can be performed by using `global_moran_test()` as shown in the code chunk below. 

```{r}
global_moran_test(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
```

### 3.2.3 Step 2C: Performing Global Moran's I permutation test

Monte Carlo simulation should be used to perform the statistical test. The number of simulations is always equal nsim + 1.

```{r}
set.seed(1234)

global_moran_perm(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt,
                  nsim = 99)
```
## 3.3 Step 3: Computing local Moran's I

```{r}
lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(local_moran)
```

### 3.3.1 Plotting LISA map

```{r}
lisa_sig <- lisa %>%
  filter(p_ii_sim < 0.05)
tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons() + 
  tm_borders(fill_alpha = 0.5) +
tm_shape(lisa_sig) +
tm_polygons(fill = "mean",
            fill.legend = tm_legend(
              text.size = 1.20,
              width = 6)) +
  tm_borders(fill_alpha = 0.4) +
  tm_layout(legend.title.size = 1.10)
```

