---
title: "Take-home_Ex01"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
bibliography: references.bib
csl: apa.csl
#draft: true
---

## Introduction and Study Context

Forest and peatland fires in Indonesian Kalimantan are a recurring seasonal hazard with substantial public-health, economic, and environmental consequences, driven in large part by historically drained peatland that burns far more readily than intact peat swamp forest. Central Kalimantan's Kapuas Regency carries a particularly well-documented version of this risk: most of the roughly one-million-hectare area affected by the mid-1990s Peatland Development Project (the Mega Rice Project) falls within it, and existing research has linked the resulting drained peat to fire activity concentrated along roads and rivers, driven substantially by human land-clearing activity [@thoha2019characteristics]. Satellite-based fire detection systems such as NASA's FIRMS now make it possible to observe this activity at fine spatial and temporal resolution across an entire fire season, but a raw point map of detections does not, on its own, answer whether that activity is meaningfully clustered, at what spatial scales, or what that clustering implies for where and when a response agency should concentrate resources.

This report applies point pattern analysis to FIRMS MODIS fire detections in Kapuas Regency across the 2026 observation window (1 January-11 September) to address exactly this gap. Kapuas was not chosen a priori: it was selected through a data-driven screening process across all 58 Kalimantan regencies (Regency Selection), on the basis of both a substantial fire count and a spread ratio indicating genuinely dispersed rather than corner-concentrated activity -- a selection subsequently corroborated by the literature identifying Kapuas as an area of documented, infrastructurally-driven fire risk. The analysis proceeds through first-order methods (where is intensity elevated, and by how much) and second-order methods (is that concentration statistically distinguishable from chance, and does genuine point-to-point clustering persist once known geography is accounted for), before translating these findings into concrete public-safety implications.

## Research Question and Objectives

**Research question.** Is Kapuas Regency's 2026 fire activity spatially clustered beyond what its already-known fire-prone geography would predict, and if so, at what spatial scales -- and what does this imply for public-safety planning and resource allocation?

This question has two parts that this report addresses in sequence. The first is essentially descriptive: where is fire intensity highest, and is that concentration statistically real rather than an artefact of small-sample variation (First-Order Spatial Point Pattern Analysis)? The second is more demanding: even after accounting for the fact that some parts of Kapuas are already known to be more fire-prone than others, does a genuine, unexplained clustering signal remain -- evidence of fires interacting with, or responding to, the same underlying drivers beyond mere geography (Second-Order Spatial Point Pattern Analysis)? Both standard and intensity-adjusted second-order methods are applied specifically because the first-order analysis confirms Kapuas's intensity is strongly non-uniform, making a flat-CSR-only comparison insufficient on its own.

**Objectives.**

1. Prepare and quality-assess FIRMS fire-detection data for Kapuas Regency, resolving duplicate and near-duplicate detections, missing or invalid values, and coordinate reference system requirements (Data Preparation and Quality Assessment).
2. Characterise the first-order spatial intensity of fire activity via kernel density estimation, and formally test this intensity for departure from complete spatial randomness (First-Order Spatial Point Pattern Analysis).
3. Characterise second-order spatial structure -- nearest-neighbour spacing, empty-space distribution, and clustering across distance scales -- against both homogeneous and intensity-adjusted reference models, to distinguish genuine point interaction from intensity variation alone (Second-Order Spatial Point Pattern Analysis).
4. Synthesise first- and second-order findings into an integrated interpretation, including plausible alternative explanations and a stated account of assumptions and limitations (Integrated Findings and Discussion).
5. Communicate the analytical findings through a curated set of geovisualisations designed for clarity rather than exhaustiveness (Geovisualisation and Geocommunication).
6. Translate evidence-supported findings into concrete public-safety and resource-planning implications, explicitly distinguishing what the analysis directly supports from what it only motivates pending further data (Public-Safety Planning and Management Implications).

## Data and Study Design

### Point Event Definition

For this study, one point event is defined as one fire detection at one location on one calendar day. The following data-quality issues were identified in the FIRMS fire-detection data and addressed before point events were finalised; full inspection code, diagnostic output, and cleaning code are reported in Data Preparation and Quality Assessment below.

- **Duplicate observations.** A literal row-duplicate check found zero exact duplicates across the combined 64,044-record dataset.
- **Repeated observations or detections.** MODIS fire pixels are approximately 1km, and the same physical fire is frequently caught more than once — by Terra and Aqua on the same day, or by the same satellite's separate day/night overpasses. 6,747 location/date groups (rounded to \~1km precision) involved 14,985 individual detections — 23.4% of all records — reflecting genuine satellite revisit geometry rather than independent events. Each group was collapsed to a single representative record (the highest-confidence detection), which is what drives the bulk of the cleaning reduction reported below.
- **Missing coordinates.** None present — `latitude` and `longitude` had zero missing values in the combined dataset.
- **Invalid coordinates or geometries.** None present in the fire records: no values outside the physical lat/lon range, and no "null island" (0,0) artifacts.
- **Observations outside the study area.** Fire records were filtered to those falling within a Kalimantan regency polygon via an inner spatial join (Regency Selection), then filtered further to Kapuas specifically and reprojected to a planar CRS (Data Preparation, CRS Transformation and Kapuas Subsetting).
- **Observations outside the selected temporal period.** None — the archive and NRT products' native coverage windows already matched the intended 1 Jan – 11 Sep 2026 observation period exactly (confirmed by `range(fires_raw$acq_date)`), so no additional temporal filter was required.
- **Other data-quality issues.** 3 records with non-positive fire radiative power (`frp` ≤ 0) — physically implausible for a genuine detection — were removed during cleaning.

### Spatial Study Window

Kapuas Regency, Central Kalimantan, was selected as the study area through a data-driven screening process rather than chosen first and justified after the fact. All 58 Kalimantan regencies were profiled on cleaned fire-point count and spatial spread (convex-hull area relative to regency area); 44 regencies met a minimum-count floor of 50 points. Kapuas and Sintang stood out with both a substantial point count and a spread ratio above 1.3. Candidate maps of the leading regencies showed Kapuas's fires forming a distinctive linear, corridor-like concentration rather than diffuse scatter or single-corner clustering — a pattern independently corroborated by existing literature on Kapuas identifying peatland drainage from the mid-1990s Peatland Development Project (PLG / Mega Rice Project) as a documented driver of fire risk concentrated along roads and rivers [@thoha2019characteristics]. The full screening metrics, candidate maps, and selection reasoning are documented in Data Preparation and Quality Assessment, Regency Selection.

The spatial study window, denoted $W \subset \mathbb{R}^2$, is defined as the administrative boundary polygon of Kapuas Regency. Unless otherwise noted, all maps in this report are oriented with north at the top; individual compass indicators are therefore omitted from subsequent maps, and a scale bar is retained wherever spatial scale is analytically relevant. The administrative boundary was used, rather than a bounding box or a data-driven extent (e.g., the convex hull of the observed fire points), for two reasons. First, in spatial point process analysis the observation window should represent the area within which events could in principle occur, independent of where they happened to be observed in this particular period; an administrative boundary satisfies this, while a hull constructed from the observed points would define the window circularly, using the outcome to define the space in which the outcome is measured. Second, several covariates relevant to later analysis — including distance to the nearest Regional Disaster Management Agency (*Badan Penanggulangan Bencana Daerah*, BPBD) facility, an approach applied to Kalimantan-wide fire clustering by @kindhi2026spatial — are meaningful at the administrative-jurisdiction level, since BPBD response responsibility and resourcing are themselves organised by regency. Using the administrative boundary keeps the observation window consistent with the boundary at which such covariates and any resulting mitigation planning would actually apply.

The boundary geometry itself required two corrections before it could serve as this window: a spurious Z-dimension (a known characteristic of the BIG source data) was stripped, and one invalid geometry was corrected with `st_make_valid()`. Full detail is in Administrative Boundary Preparation below.

### Temporal Study Window

The temporal study window spans 1 January – 11 September 2026, matching the full observation period covered by the combined FIRMS archive and NRT products (see Data Acquisition below). This window was fixed by data availability at the time of access rather than being a chosen analytical boundary — FIRMS data becomes available only up to the date of download, so no post-window fire activity could be included even if it existed.

No temporal aggregation (e.g., binning into months or fire-season periods) has yet been applied to the point events; the full window's cleaned point events for Kapuas (1,849, per the screening table) are treated as one pooled cross-section for the first-order and second-order spatial analyses in Sections 5–6. Whether a temporal aggregation or a genuinely spatio-temporal analysis is warranted was assessed once the first-order intensity pattern was examined: fire activity in Kapuas is strongly concentrated in the final six weeks of the observation window (First-order Spatial Point Pattern Analysis, Temporal Profile of Fire Activity), coinciding with the documented onset of the 2026 dry season, rather than spread evenly across Jan–Sep. This pooled cross-sectional analysis proceeds with that late-season concentration noted as a scope limitation; a spatio-temporal extension splitting the window into quiet and surge periods is attempted separately as this report's Bonus/Value-Added component.
### Analytical Assumptions

Assumptions are specific to each analytical method rather than to the point pattern in general, so they are not enumerated here in the abstract. Each method introduced in First-order Spatial Point Patterns and Second-order Spatial Point Patterns (Sections 5–6) states its own assumptions immediately before it is applied — for example, kernel density estimation's assumption of a smoothly varying underlying intensity, or the K-function's assumption of stationarity under the reference null model — together with a judgement of whether that assumption is reasonable for Kapuas's fire pattern. This keeps each assumption tied to the specific method and result it constrains, rather than listed generically before any method has been chosen.

### Methodological Decisions

The coordinate reference system used throughout data preparation so far is WGS84 (EPSG:4326), which is adequate for the operations already performed — bounding-box cropping, point-in-polygon joins, and area computation via `st_area()`, which handles geographic coordinates correctly. It is not suitable, however, for the planar operations `spatstat` requires (kernel density bandwidths, K-function distances), which need a projected CRS with linear units. Kapuas was transformed to EPSG:32750 (WGS 84 / UTM zone 50S), the appropriate UTM zone for its longitude; Kapuas's extent straddles the 49/50 zone boundary at 114°E, a minor edge effect addressed in full in Data Preparation, CRS Transformation and Kapuas Subsetting below. Restricting the data to the Kapuas subset is the remaining Data Preparation step, addressed next; the resulting projected data is then used to construct the `spatstat` `ppp` and `owin` objects at the start of First-order Spatial Point Pattern Analysis.

Bandwidth, distance range, edge correction, and simulation settings are method-specific and are documented alongside each method as it is introduced in Sections 5–6.

## Data Preparation and Quality Assessment

### Data Acquisition

This study uses two categories of source data.

**Fire event data** were obtained from NASA's Fire Information for Resource Management System (FIRMS), MODIS Collection 61 product, downloaded on 11 September 2026. FIRMS distributes fire detections across two products with different processing latency: the *standard/archive* product, which is quality-controlled but has a multi-month processing lag, and the *near-real-time (NRT)* product, which is available within hours of satellite overpass but has not yet been reprocessed to standard quality. Because the standard product's lag meant it only covered detections up to 30 April 2026, both products were downloaded and combined to cover the full observation window of 1 January -- 11 September 2026:

- `fire_archive_M-C61_803864.csv` -- standard product, 1 Jan -- 30 Apr 2026 (4,122 records)
- `fire_nrt_M-C61_803864.csv` -- NRT product, 1 May -- 11 Sep 2026 (59,922 records)

Both files cover all of Indonesia; neither is pre-clipped to Kalimantan.

**Administrative boundary data** were obtained from Indonesia's Geospatial Information Agency (BIG), 2023 village/kelurahan-level administrative boundary dataset (`Batas_Wilayah_KelurahanDesa_10K_AR`), at 1:10,000 scale, national coverage, accessed on 11 September 2026. 

```{r}
#| label: setup
#| message: false
#| code-fold: true
#| code-summary: "Show package imports"

library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(patchwork)
library(ggspatial)
library(ggiraph)
```

### Administrative Boundary Preparation

The national boundary dataset contains all of Indonesia at village level. Rather than reading the full file and filtering afterward -- which would require parsing geometry for tens of thousands of polygons only to discard most of them -- the province filter was pushed down to the OGR driver at read time via an SQL `WHERE` clause, so only Kalimantan's village polygons were parsed.

```{r}
#| label: extract-kalimantan-boundary
#| code-fold: true
#| code-summary: "Show boundary extraction code"

kalimantan_provinces <- c("Kalimantan Barat", "Kalimantan Tengah",
                           "Kalimantan Selatan", "Kalimantan Timur",
                           "Kalimantan Utara")

where_clause <- paste0(
  "WADMPR IN ('", paste(kalimantan_provinces, collapse = "','"), "')"
)

desa_kalimantan <- st_read(
  "data/Batas_Wilayah_KelurahanDesa_10K_AR.shp",
  query = paste("SELECT * FROM Batas_Wilayah_KelurahanDesa_10K_AR WHERE", where_clause)
)

nrow(desa_kalimantan)
```

The extraction returned 7,234 village-level polygons across the five Kalimantan provinces.

Two data-quality issues were identified and corrected before any spatial analysis. First, the source geometry carried a spurious Z-dimension (elevation-like values, meaningless for a 2D administrative boundary) -- a known characteristic of this BIG source. Second, one invalid geometry was present after the Z-dimension strip, corrected with `st_make_valid()`.

Village-level polygons were then dissolved to regency level (`WADMKK`), since the screening question -- which regency has the most, or most widely spread, fire activity -- operates at that granularity. Regency area was computed at this stage, since it is needed later for the screening metrics.

```{r}
#| label: clean-and-dissolve-boundary
#| code-fold: true
#| code-summary: "Show cleaning and dissolve code"

desa_kalimantan <- st_zm(desa_kalimantan)
sum(!st_is_valid(desa_kalimantan))
desa_kalimantan <- st_make_valid(desa_kalimantan)

regency_kalimantan <- desa_kalimantan |>
  group_by(WADMPR, WADMKK) |>
  summarise(.groups = "drop") |>
  st_make_valid()

regency_kalimantan$area_km2 <- as.numeric(st_area(regency_kalimantan)) / 1e6

regency_areas <- regency_kalimantan |>
  st_drop_geometry() |>
  select(WADMPR, WADMKK, area_km2)

nrow(regency_kalimantan)

st_write(regency_kalimantan, "data/regency_kalimantan.gpkg", delete_dsn = TRUE)
```

Dissolving produced 58 regency polygons, matching the expected regency count across the five provinces.

### Fire Event Data Preparation

#### Import and integration

The archive and NRT products required one explicit harmonisation decision. The archive product includes a `type` field (classifying each detection as vegetation fire, active volcano, other static land source, or offshore) that the NRT product does not carry, since that classification is only assigned during reprocessing to standard quality. Rather than silently dropping the column or letting a type mismatch cause an opaque row-binding failure, NRT records were explicitly assigned `NA` for `type`, preserving the archive product's classification where available and documenting its absence for NRT records rather than imputing it. A similar type inconsistency was found and resolved for the `version` field.

```{r}
#| label: import-fire-data
#| message: false
#| code-fold: true
#| code-summary: "Show import and integration code"

archive <- read_csv("data/fire_archive_M-C61_803864.csv", show_col_types = FALSE) |>
  mutate(type = as.character(type), version = as.character(version))

nrt <- read_csv("data/fire_nrt_M-C61_803864.csv", show_col_types = FALSE) |>
  mutate(type = NA_character_)

fires_raw <- bind_rows(archive, nrt)

nrow(fires_raw)
range(fires_raw$acq_date)
sum(duplicated(fires_raw))
```

Combining both products yielded 64,044 records. The combined date range matches the intended 1 January -- 11 September 2026 observation window exactly, confirming no records fall outside the selected temporal period -- each source file only ever covered its own designated window, so no explicit temporal filter was required beyond the acquisition choice itself. Zero exact-duplicate rows were found; this check identifies only literal full-row duplicates, and near-duplicate detections are addressed separately below.

#### Data inspection

Before any cleaning decisions were made, the combined dataset was inspected for missingness, coordinate validity, and value-range plausibility.

```{r}
#| label: fires-data-inspection

glimpse(fires_raw)

# Missingness per column
colSums(is.na(fires_raw))

# Coordinate validity - valid lat/lon ranges, and the classic "null island" check
sum(fires_raw$latitude < -90 | fires_raw$latitude > 90, na.rm = TRUE)
sum(fires_raw$longitude < -180 | fires_raw$longitude > 180, na.rm = TRUE)
sum(fires_raw$latitude == 0 & fires_raw$longitude == 0, na.rm = TRUE)

# Confidence field distribution
summary(fires_raw$confidence)

# brightness/frp sanity check - both should be strictly positive for a genuine fire detection
sum(fires_raw$brightness <= 0, na.rm = TRUE)
sum(fires_raw$frp <= 0, na.rm = TRUE)
```

The only missing values found were in `type` (59,922 NAs), matching the expected NRT-product gap already documented above -- no other column contained missing values. Coordinates were fully valid: no values outside the physical latitude/longitude range, and no records at (0,0) ("null island," a common sentinel-value artifact in geospatial data, absent here). MODIS confidence for this product runs 0--100 (not 0--99 as sometimes assumed), with a median of 69. `brightness` contained no non-positive values. `frp` (fire radiative power) contained **3 records with a value of 0 or below**, which is physically implausible for a genuine fire detection and was treated as an invalid-observation issue, resolved during cleaning below.

#### Near-duplicate detections

A literal row-duplicate check (above) found none, but MODIS fire pixels are approximately 1km, and the same physical fire is frequently caught more than once -- by Terra and Aqua on the same day, or by the same satellite's separate day/night overpasses. These are not exact duplicates in the raw data (coordinates differ slightly, acquisition time differs) but represent the same underlying point event if left unaddressed.

```{r}
#| label: near-duplicate-check

# Round coordinates to ~1km precision (matching MODIS pixel size) and
# check same-day, same-rounded-location detections
near_dupes <- fires_raw |>
  mutate(lat_round = round(latitude, 2), lon_round = round(longitude, 2)) |>
  count(lat_round, lon_round, acq_date, name = "n_detections") |>
  filter(n_detections > 1)

nrow(near_dupes)
sum(near_dupes$n_detections)
```

6,747 location/date groups involved a total of 14,985 individual detections -- 23.4% of all records. Before deciding how to handle this, the underlying mechanism was checked rather than assumed.

```{r}
#| label: inspect-near-duplicates

near_dupe_detail <- fires_raw |>
  mutate(lat_round = round(latitude, 2), lon_round = round(longitude, 2)) |>
  add_count(lat_round, lon_round, acq_date, name = "n_detections") |>
  filter(n_detections > 1)

# Group size distribution
table(near_dupe_detail$n_detections) |> head(10)

# For exact pairs, are they different satellites (expected) or the same satellite (also plausible)?
pairs_same_satellite <- near_dupe_detail |>
  filter(n_detections == 2) |>
  group_by(lat_round, lon_round, acq_date) |>
  summarise(n_satellites = n_distinct(satellite), .groups = "drop") |>
  count(n_satellites)

pairs_same_satellite
```

Group sizes ranged from 2 to 6 detections at the same location/day (11,068 pairs, 2,925 triples, 812 groups of four, 150 of five, 30 of six). Among the 5,534 exact pairs, 3,298 (60%) involved two different satellites -- the expected pattern from Terra and Aqua both overpassing the same active fire within a day -- and the remaining 2,236 (40%) involved the same satellite appearing twice, consistent with a single MODIS instrument's separate day and night overpasses catching a fire still burning across both. Neither case reflects a data error; both reflect genuine satellite revisit geometry rather than one independent point event per detection.

Collapsing each of the 6,747 groups to a single representative record removes 8,238 excess detections (14,985 minus the 6,747 retained) -- an approximately 12.9% reduction, which is what the cleaning step below actually applies (the near-duplicate group counts above are computed on the full 64,044 raw records, ahead of the 3-record `frp` filter applied in cleaning; the resulting discrepancy is negligible relative to the overall reduction).

#### Cleaning to point events

Applying the point event definition stated in Data and Study Design: where a location/day group contained multiple detections, the record with the highest confidence value was retained as representative. The 3 records with non-positive `frp` identified during inspection were also removed at this stage, as physically invalid.

```{r}
#| label: clean-fires-data
#| code-fold: true
#| code-summary: "Show cleaning code"

fires_clean <- fires_raw |>
  filter(frp > 0) |>
  mutate(lat_round = round(latitude, 2), lon_round = round(longitude, 2)) |>
  group_by(lat_round, lon_round, acq_date) |>
  slice_max(confidence, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(-lat_round, -lon_round)

nrow(fires_raw)
nrow(fires_clean)
```

Cleaning reduced the dataset from 64,044 to 55,804 point events (a 13.0% reduction), consistent with the near-duplicate reduction rate identified above.

### Regency Selection

The brief requires selecting a single Kalimantan regency as the study area but does not prescribe how. Rather than choosing a regency first and hoping it contained an analytically useful fire pattern, the selection was treated as a data-driven screening step: all Kalimantan regencies were profiled on fire count and spatial distribution, using the cleaned point-event data, before any one regency was chosen -- so the eventual choice can be defended against stated criteria rather than justified after the fact. The outcome of this screening (Kapuas) is stated in Data and Study Design, Spatial Study Window above; this section documents the full workflow: screening all regencies on two criteria, shortlisting five for closer inspection, mapping the shortlist both nationally and individually, and finally selecting Kapuas from among the five.

Two criteria were defined in advance:

1.  **Sufficient point count.** Second-order spatial statistics (K-function, pair correlation function) become unstable with very few points; a minimum of 50 detections was set as a floor for a regency to be considered.
2.  **Spatial spread relative to regency size.** Raw fire count alone conflates fire activity with regency size -- a large regency with fires confined to one corner could out-count a smaller regency where fires are genuinely dispersed, without being the more interesting subject for point-pattern analysis. A spread ratio (convex hull area of the fire points, divided by regency area) captures this: values near 1 indicate fires distributed across most of the regency; low values indicate spatial concentration in one part of it.

::: callout-note
## Why a fixed count floor, and what it does not guarantee

The 50-point floor and the spread-ratio comparison are practical decision rules, not statistical guarantees. Setting the floor at 50 rather than, say, 30 or 100 is itself a judgement call: a lower floor would let more regencies compete on spread ratio but at the cost of noisier convex-hull estimates from very small samples; a higher floor would give more stable spread estimates but exclude some regencies from consideration entirely, regardless of how genuinely dispersed their (small number of) fires might actually be. Consistent with the brief's own guidance that "the objective is not to eliminate all assumptions... \[but\] to demonstrate that you understand them and recognise their implications," this floor is treated here as a stated, defensible threshold rather than a search for one "correct" cutoff. Its main implication: regencies with fewer than 50 detections are excluded from the *dispersion* comparison specifically because their spread ratio cannot be reliably estimated -- not because their fire activity is assumed uninteresting.
:::

#### Spatial filtering to Kalimantan

Cleaned fire coordinates were converted to spatial point geometry in WGS84 (EPSG:4326, matching the FIRMS coordinate reference) and filtered to Kalimantan in two stages: an inexpensive bounding-box crop first, followed by a true point-in-polygon join against the regency boundaries. The bounding-box step alone is insufficient, since a rectangular extent includes area between and around the irregular regency boundaries and would therefore wrongly retain observations outside the actual study area; the polygon join, using an inner join (`left = FALSE`), is what correctly excludes any point falling outside every regency polygon.

```{r}
#| label: spatial-filter-fires
#| code-fold: true
#| code-summary: "Show spatial filtering code"

fires_sf <- st_as_sf(fires_clean, coords = c("longitude", "latitude"),
                      crs = 4326, remove = FALSE)

kalimantan_bbox <- st_bbox(regency_kalimantan)
fires_kalimantan_bbox <- st_crop(fires_sf, kalimantan_bbox)

fires_regency <- st_join(fires_kalimantan_bbox, regency_kalimantan,
                          join = st_intersects, left = FALSE)

nrow(fires_regency)
```

31,586 cleaned point events were confirmed within a Kalimantan regency polygon.

#### Screening metrics

```{r}
#| label: regency-screening-metrics

regency_diag <- fires_regency |>
  st_drop_geometry() |>
  count(WADMPR, WADMKK, name = "n") |>
  filter(n >= 50)

hull_sf <- fires_regency |>
  semi_join(regency_diag, by = c("WADMPR", "WADMKK")) |>
  group_by(WADMPR, WADMKK) |>
  summarise(.groups = "drop") |>
  st_convex_hull()

hull_sf$hull_area_km2 <- as.numeric(st_area(hull_sf)) / 1e6

hull_areas <- hull_sf |>
  st_drop_geometry() |>
  select(WADMPR, WADMKK, hull_area_km2)

screening <- regency_diag |>
  left_join(regency_areas, by = c("WADMPR", "WADMKK")) |>
  left_join(hull_areas, by = c("WADMPR", "WADMKK")) |>
  mutate(
    density_per_1000km2 = round(n / area_km2 * 1000, 2),
    spread_ratio = round(hull_area_km2 / area_km2, 3)
  ) |>
  arrange(desc(n))

write_csv(screening, "data/regency_screening.csv")

screening
```

Applying the 50-point floor left 44 of the 58 regencies eligible. Kapuas and Sintang stand out among candidates with both a substantial point count and a spread ratio above 1.3, indicating fires distributed across a footprint comparable to or larger than a simple corner-concentration pattern.

#### Selecting the five candidates for closer inspection

Rather than mapping all 44 eligible regencies, five were shortlisted to represent both ends of the two screening criteria, since a regency can score well on one without the other.

```{r}
#| label: candidate-selection-rationale

# Four most spatially dispersed regencies (highest spread ratio)
screening |> slice_max(spread_ratio, n = 4) |> select(WADMKK, n, spread_ratio)

# Two highest-count regencies, for comparison against the dispersion leaders
screening |> slice_max(n, n = 2) |> select(WADMKK, n, spread_ratio)

candidates <- c("Seruyan", "Kayong Utara", "Sintang", "Kapuas",
                "Kotawaringin Timur")
```

The four most spatially dispersed regencies -- Seruyan (spread ratio 1.554), Kayong Utara (1.531), Sintang (1.314), and Kapuas (1.308) -- were carried forward directly, since these represent the most genuinely dispersed fire activity in the dataset rather than counts inflated by a small number of outlying points.

To this set, one high-count regency was added, to give the shortlist a contrasting, concentrated case rather than five dispersion-leaning regencies alone. Ketapang records the highest count nationally (5,496) but a spread ratio of 1.099 -- close enough to the neutral value of 1 that it offers neither a clearly dispersed nor a clearly concentrated pattern, making it visually redundant next to the four dispersion leaders already selected. Kotawaringin Timur, the second-highest-count regency (1,967), was selected instead: its spread ratio of 0.858 is clearly below 1, giving the shortlist a genuinely concentrated-activity case to contrast against the four dispersed ones.

::: callout-warning
## A density- or count-only shortlist would have been misleading

Kota Palangkaraya records the highest fire *density* nationally (323.6 per 1,000km²) simply because it is a very small, largely urban regency -- a modest count of 926 divided by an area of only 2,861km². Ketapang records the highest *absolute count* (5,496) but, as noted above, a spread ratio too close to neutral to be analytically interesting on its own. Neither regency appears in the five shortlisted below: selecting by density or count alone, without also checking spread, would have highlighted regencies whose apparent "extremeness" is a size artefact rather than evidence of a genuinely interesting spatial pattern. This mirrors the brief's broader caution that an observed concentration of events should not automatically be read as meaningful without considering what produced the observation.
:::

#### Kalimantan-wide overview

```{r}
#| label: national-fire-density
#| code-fold: true
#| code-summary: "Show density calculation code"

regency_counts_all <- fires_regency |>
  st_drop_geometry() |>
  count(WADMPR, WADMKK, name = "n")

regency_density <- regency_areas |>
  left_join(regency_counts_all, by = c("WADMPR", "WADMKK")) |>
  mutate(
    n = tidyr::replace_na(n, 0),
    density_per_1000km2 = round(n / area_km2 * 1000, 2)
  )

regency_density_sf <- regency_kalimantan |>
  left_join(
    regency_density |> select(WADMPR, WADMKK, n, density_per_1000km2),
    by = c("WADMPR", "WADMKK")
  )

summary(regency_density_sf$density_per_1000km2)
sum(regency_density_sf$n == 0)
```

```{r}
#| label: national-overview-map-interactive
#| fig-width: 9
#| fig-height: 7
#| code-fold: true
#| code-summary: "Show plotting code"

candidate_boundaries <- regency_density_sf |> filter(WADMKK %in% candidates)

p_national <- ggplot() +
  geom_sf_interactive(
    data = regency_density_sf,
    aes(fill = density_per_1000km2,
        tooltip = paste0(WADMKK, ", ", WADMPR,
                          "\nFires: ", n,
                          "\nDensity: ", density_per_1000km2, " per 1,000km\u00b2"),
        data_id = WADMKK),
    colour = "white", linewidth = 0.15
  ) +
  geom_sf(data = fires_regency, shape = 21, fill = "white", colour = "black",
          size = 0.35, stroke = 0.15, alpha = 0.55) +
  geom_sf(data = candidate_boundaries, fill = NA, colour = "red",
          linewidth = 0.6) +
  scale_fill_viridis_c(
    name = "Fires per\n1,000 km²",
    option = "viridis",
    trans = "sqrt",
    breaks = c(0, 25, 50, 100, 200, 300)
  ) +
  annotation_scale(location = "bl", width_hint = 0.25) +
  
  theme_void() +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5, margin = margin(b = 4)),
    plot.subtitle = element_text(size = 10, hjust = 0.5, colour = "grey30", margin = margin(b = 10)),
    legend.position = "right"
  ) +
  labs(
    title = "Fire Detection Density Across Kalimantan Regencies",
    subtitle = "Red outlines mark the five shortlisted regencies."
  )

ggsave("data/kalimantan_overview_density.png", plot = p_national,
       width = 9, height = 7, dpi = 150)

girafe(
  ggobj = p_national,
  width_svg = 9, height_svg = 7,
  options = list(
    opts_hover(css = "stroke:black;stroke-width:2.5px;"),
    opts_tooltip(css = "background-color:white;color:black;padding:6px 10px;
                  border-radius:4px;font-family:sans-serif;font-size:12px;
                  box-shadow:2px 2px 6px rgba(0,0,0,0.3);"),
    opts_sizing(rescale = TRUE)
  )
)
```

::: callout-important
## 🚨 Interactive Map

**Hover your mouse over a regency for its name, fire count, and density above**
:::

The five shortlisted regencies (outlined in red, selected above) sit in the moderate-to-high density band rather than at either extreme -- consistent with the reasoning in Selecting the five candidates for closer inspection, which deliberately avoided regencies whose extreme count or density is a small-area artefact. Kayong Utara's split across two separated landmasses is also visible here, foreshadowing the observation-window concern raised for it in the candidate maps below. Hovering over any regency reveals its exact count and density, useful for checking the visual pattern against the full screening table without switching views.

#### Candidate regency maps

To support the selection decision at a scale where individual fire locations are actually legible, the five shortlisted regencies were mapped individually.

```{r}
#| label: candidate-plot-function
#| code-fold: true
#| code-summary: "Show plotting function and PNG export code"

make_candidate_plot <- function(regency_name) {
  boundary <- regency_kalimantan |> filter(WADMKK == regency_name)
  fires <- fires_regency |> filter(WADMKK == regency_name)

  ggplot() +
    geom_sf(data = boundary, fill = "grey95", colour = "grey40") +
    geom_sf(data = fires, colour = "firebrick", size = 0.6, alpha = 0.5) +
    annotation_scale(location = "bl", width_hint = 0.3) +
    theme_void() +
    theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5,
                                margin = margin(b = 8))
    ) +
    labs(title = regency_name)
}

plots <- setNames(lapply(candidates, make_candidate_plot), candidates)

for (name in names(plots)) {
  ggsave(
    filename = paste0("data/candidate_", gsub(" ", "_", name), ".png"),
    plot = plots[[name]],
    width = 7, height = 6, dpi = 150
  )
}
```

::: panel-tabset
#### Seruyan

```{r}
#| label: candidate-tab-seruyan
#| fig-width: 7
#| fig-height: 6
#| echo: false

plots[["Seruyan"]]
```

#### Kayong Utara

```{r}
#| label: candidate-tab-kayong-utara
#| fig-width: 7
#| fig-height: 6
#| echo: false

plots[["Kayong Utara"]]
```

#### Sintang

```{r}
#| label: candidate-tab-sintang
#| fig-width: 7
#| fig-height: 6
#| echo: false

plots[["Sintang"]]
```

#### Kapuas

```{r}
#| label: candidate-tab-kapuas
#| fig-width: 7
#| fig-height: 6
#| echo: false

plots[["Kapuas"]]
```

#### Kotawaringin Timur

```{r}
#| label: candidate-tab-kotawaringin-timur
#| fig-width: 7
#| fig-height: 6
#| echo: false

plots[["Kotawaringin Timur"]]
```
:::

Seruyan and Kotawaringin Timur, despite reasonable spread ratios, show fires visibly concentrated in each regency's southern portion rather than genuinely dispersed across the territory, indicating their spread-ratio scores were inflated by a small number of outlying points rather than broad-based activity. Kayong Utara is split across two landmasses separated by open water, complicating the definition of a single coherent observation window. Sintang shows fires distributed across two connected lobes of the regency with minimal concentration in any single area. Kapuas shows a distinctive linear, corridor-like concentration of fire events rather than diffuse scatter, consistent with activity tracking an accessibility feature such as a river or road network rather than being uniformly distributed across the landscape.

#### Selection and justification

Kapuas was selected as the study area over the other four shortlisted regencies (and over Ketapang, addressed above), for two reasons beyond the screening metrics alone.

First, Kapuas Regency has been identified in the literature as an area of extreme fire risk within Central Kalimantan, on the basis that most of the roughly one-million-hectare area affected by the mid-1990s Peatland Development Project (*Pengembangan Lahan Gambut*, PLG) -- also referred to as the Mega Rice Project -- falls within the regency [@thoha2019characteristics]. The PLG project constructed thousands of kilometres of drainage canals across peat swamp forest with the intention of converting it to rice cultivation; the resulting drainage lowered the peat water table across the affected area, leaving a landscape substantially more prone to fire than undrained peat swamp forest. This gives Kapuas's fire pattern a documented infrastructural and anthropogenic driver that can be examined analytically, rather than an unexplained spatial pattern.

Second, existing spatial research conducted specifically on Kapuas found high fire activity concentrated in peatland and swamp shrub areas close to roads and rivers and far from villages, with human activities such as land clearing and deliberate burning identified as the predominant cause [@thoha2019characteristics]. This corresponds closely to the corridor-like concentration observed in the candidate map above, giving the subsequent point-pattern analysis a concrete, literature-grounded pattern to test rather than treating any observed clustering as unexplained.

::: callout-important
## Final selection

Kapuas Regency was selected as the study area: it is one of only two regencies with both a substantial fire count and a spread ratio above 1.3, its fires form a distinctive corridor-like pattern rather than diffuse scatter, and that pattern is independently corroborated by prior research identifying peatland drainage from the Mega Rice Project as a documented driver of fire risk in Kapuas specifically [@thoha2019characteristics].
:::

### CRS Transformation and Kapuas Subsetting

```{r}
#| label: kapuas-extent-check
#| code-fold: true
#| code-summary: "Show extent check code"

kapuas_boundary <- regency_kalimantan |> filter(WADMKK == "Kapuas")

st_bbox(kapuas_boundary)
```

```{r}
#| label: crs-transform-kapuas
#| code-fold: true
#| code-summary: "Show CRS transformation and Kapuas subsetting code"

kapuas_fires_sf <- fires_regency |>
  filter(WADMKK == "Kapuas") |>
  st_transform(crs = 32750)

kapuas_boundary_utm <- kapuas_boundary |>
  st_transform(crs = 32750)

st_crs(kapuas_fires_sf)$epsg
nrow(kapuas_fires_sf)
```

Kapuas's longitudinal extent (113.68°E–114.80°E) straddles the UTM zone 49/50 boundary at 114°E. EPSG:32750 (zone 50S) was used for the entire regency regardless, since the majority of both the boundary area and the fire points themselves fall east of the boundary (median fire longitude 114.3°E, per the `ppp` summary in First-order Spatial Point Pattern Analysis below); a small minority of points near the western edge incur minor positional distortion under this choice. Splitting the analysis across two CRS zones was judged not worth the added complexity for this degree of edge effect. The transformed fire count (1,849) matches the Kapuas row in the screening table exactly, confirming no records were lost in reprojection.

## First-Order Spatial Point Pattern Analysis

### Constructing the Point Pattern and Observation Window

```{r}
#| label: setup-spatstat
#| message: false
#| code-fold: true
#| code-summary: "Show additional package imports for spatstat analysis"

pacman::p_load(spatstat, terra, tmap)
```

```{r}
#| label: construct-ppp-owin
#| code-fold: true
#| code-summary: "Show ppp and owin construction code"

kapuas_owin <- as.owin(kapuas_boundary_utm)
class(kapuas_owin)

# Check window structure -- is it a single polygon or multiple pieces?
length(kapuas_owin$bdry)
plot(kapuas_owin, main = "Kapuas owin -- checking window structure")

# Keep only marks relevant to later analysis; drop redundant coordinate
# duplicates, constant regency-level fields, and the mostly-missing `type`
# field (documented in Point Event Definition as an NRT-product gap)
kapuas_fires_for_ppp <- kapuas_fires_sf |>
  select(acq_date, confidence, frp, daynight)

kapuas_ppp <- as.ppp(kapuas_fires_for_ppp)
class(kapuas_ppp)

kapuas_ppp <- kapuas_ppp[kapuas_owin]
kapuas_ppp
summary(kapuas_ppp)
```

```{r}
#| label: verify-sliver-polygon
#| code-fold: true
#| code-summary: "Show sliver-polygon verification code"

kapuas_boundary_parts <- kapuas_boundary_utm |>
  st_cast("POLYGON")

kapuas_boundary_parts$part_area_km2 <- as.numeric(st_area(kapuas_boundary_parts)) / 1e6

kapuas_boundary_parts |> st_drop_geometry() |> select(part_area_km2)

small_part <- kapuas_boundary_parts |> slice_min(part_area_km2, n = 1)

st_join(kapuas_fires_sf, small_part, join = st_intersects, left = FALSE) |> nrow()
```

The `owin` object comprises two disjoint polygons rather than one: a primary polygon covering 17,065.8km² (99.999% of the total window area) and a second fragment of only 0.1457km² -- smaller than a single MODIS detection pixel (\~1km²). The verification above confirms this fragment contains zero fire points, consistent with it being a topological sliver introduced during the village-to-regency dissolve (Administrative Boundary Preparation) rather than a genuine, analytically meaningful exclave of Kapuas. Its presence does not affect the point pattern itself, but it does trivially reduce the usable window area below the boundary's true geodetic area, contributing to the small residual gap in the intensity cross-check below.

The average intensity reported above (1.083 × 10⁻⁷ points/m²) converts to approximately 108.34 fires per 1,000km², closely matching the 108.02 figure computed independently in the screening table (Regency Selection). The small residual difference stems from the projected UTM window area (17,065.9km², combining both polygons above) being marginally smaller than the geodetic area used in the original screening computation (17,117.52km²) -- an expected consequence of projection distortion and the sliver-polygon effect just described, rather than an inconsistency in the underlying point data.

```{r}
#| label: kapuas-boundary-map
#| fig-width: 6
#| fig-height: 10

tm_shape(kapuas_boundary_utm) +
  tm_polygons(fill = "grey95", col = "grey40") +
  tm_shape(kapuas_fires_sf) +
  tm_dots(fill = "firebrick", size = 0.3, fill_alpha = 0.6) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_title("Kapuas Regency: Fire Detections (Projected, UTM 50S)") +
  tm_layout(inner.margins = c(bottom = 0.06, top = 0.02, left = 0.02, right = 0.02))
```

### Kernel Density Estimation

Bandwidth selection is the central tuning decision in KDE: a **fixed bandwidth** applies one smoothing radius uniformly across the study area, while an **adaptive bandwidth** allows the radius to shrink in dense areas and widen in sparse ones. `spatstat` provides several automatic fixed-bandwidth selectors; `bw.ppl()` is recommended when a pattern consists predominantly of tight clusters, while `bw.diggle()` is tuned for detecting a single tight cluster within otherwise random noise [@baddeley2015spatial]. Point event coordinates were rescaled from metres to kilometres before bandwidth selection, purely so the resulting bandwidth values are legible (a bandwidth in metres for a regency spanning \~340km north-south would be an unwieldy number). A Gaussian kernel was used throughout; per Baddeley et al. (2015), kernel shape has substantially less influence on the resulting density surface than bandwidth choice, so this was not treated as a parameter requiring separate justification or comparison.

```{r}
#| label: rescale-ppp-km
#| code-fold: true
#| code-summary: "Show rescaling code"

kapuas_ppp_km <- rescale.ppp(kapuas_ppp, 1000, "km")
kapuas_ppp_km
```

```{r}
#| label: bandwidth-comparison

bw_ppl_val <- bw.ppl(kapuas_ppp_km)
bw_diggle_val <- bw.diggle(kapuas_ppp_km)

bw_ppl_val
bw_diggle_val
```

`bw.ppl()` and `bw.diggle()` returned 1.113km and 0.550km respectively -- roughly a factor of two apart. This disagreement is itself informative: it suggests Kapuas's fire pattern does not cleanly match either selector's ideal use case (a single tight cluster in noise, or a pattern dominated by uniformly tight clusters) [@baddeley2015spatial], consistent with a sustained linear corridor rather than either textbook shape. `bw.ppl()` was carried forward as the primary fixed-bandwidth choice, being the better-justified fit for a pattern of sustained clustering rather than an isolated hotspot.

Fixed and adaptive density surfaces were computed for comparison. The adaptive surface used `spatstat`'s default nearest-neighbour fraction (f = 0.1, left untuned since fixed bandwidth was ultimately preferred as the primary result).

```{r}
#| label: kde-fixed-vs-adaptive
#| code-fold: true
#| code-summary: "Show KDE computation code"

kde_kapuas_fixed <- density(kapuas_ppp_km,
                            sigma = bw_ppl_val,
                            edge = TRUE,
                            kernel = "gaussian")

kde_kapuas_adaptive <- adaptive.density(kapuas_ppp_km, f = 0.1, method = "kernel")
```

```{r}
#| label: kde-negative-value-check
#| code-fold: true
#| code-summary: "Show negative-value diagnostic code"

range(kde_kapuas_fixed$v, na.rm = TRUE)
sum(kde_kapuas_fixed$v < 0, na.rm = TRUE)

range(kde_kapuas_adaptive$v, na.rm = TRUE)
sum(kde_kapuas_adaptive$v < 0, na.rm = TRUE)
```

::: callout-note
## Why the density surfaces contain values marginally below zero

Both surfaces contain a small number of pixels with values marginally below zero (minimum -3.4 × 10⁻¹⁶ for fixed bandwidth, affecting 457 of the grid's pixels; -1.4 × 10⁻¹⁶ for adaptive, affecting 15), several orders of magnitude smaller than the surfaces' maximum values (6.59 and 8.37 respectively). This is floating-point noise from the edge-correction calculation near the boundary, not a genuine negative density estimate -- density cannot be negative by definition. These values were clamped to 0 before the square-root transform used below, applied purely for visualisation given the surfaces' right-skewed values (the same transform already used on the national choropleth in Regency Selection) -- the transform does not alter the underlying density estimates.
:::

::: panel-tabset
#### Fixed Bandwidth (bw.ppl)

```{r}
#| label: kde-fixed-sqrt-tab
#| fig-width: 7
#| fig-height: 9
#| code-fold: true
#| code-summary: "Show plotting code"

kde_fixed_clamped <- eval.im(pmax(kde_kapuas_fixed, 0))
kde_fixed_sqrt <- sqrt(kde_fixed_clamped)

plot(kde_fixed_sqrt, main = "Fixed bandwidth (sqrt scale, bw.ppl = 1.113km)")
```

#### Adaptive Bandwidth

```{r}
#| label: kde-adaptive-sqrt-tab
#| fig-width: 7
#| fig-height: 9
#| code-fold: true
#| code-summary: "Show plotting code"

kde_adaptive_clamped <- eval.im(pmax(kde_kapuas_adaptive, 0))
kde_adaptive_sqrt <- sqrt(kde_adaptive_clamped)

plot(kde_adaptive_sqrt, main = "Adaptive bandwidth (sqrt scale, f = 0.1)")
```
:::

Fixed bandwidth was chosen as the primary density estimate over the adaptive alternative. While adaptive bandwidth captures finer local variation in point density, its output here is visibly more fragmented -- small, sharply localized hotspots rather than continuous elevated zones -- which risks implying a level of spatial precision the underlying MODIS detection data cannot genuinely support, and obscures the sustained corridor structure that is Kapuas's defining spatial signature (Regency Selection). This sensitivity to bandwidth method is treated as a stated limitation of the fixed-bandwidth approach rather than grounds for abandoning it.

::: callout-important
## Primary first-order method

Fixed bandwidth via `bw.ppl()` (σ = 1.113km) is used as the primary kernel density estimate for Kapuas going forward. The adaptive alternative was tested and rejected: its finer local variation fragments the sustained corridor structure into disconnected hotspots, implying a precision the underlying detection data does not support. The `bw.ppl()`/`bw.diggle()` disagreement (1.113km vs. 0.550km) is retained as a documented limitation of this choice.
:::

The bandwidth and density surface computed above (`kde_kapuas_fixed`) used point coordinates rescaled to kilometres, purely so the bandwidth value itself was legible. Converting that surface directly into a georeferenced raster would carry over a subtle problem: the raster's coordinate values would still be in kilometres, but assigning a metre-based CRS (EPSG:32750) to it as metadata would not rescale those values to match -- the raster would report a correct-looking CRS while its actual coordinates were numerically wrong by a factor of 1,000. This would go unnoticed on a simple plot, but would produce visibly incorrect graticule labels and scale bar distances on the cartographic map below, since both depend on the CRS being numerically accurate, not just present as a label.

::: callout-note
## Why the density surface is recomputed here, rather than reused

The surface used for cartographic output below is recomputed on the original metre-based point pattern (`kapuas_ppp`), with the bandwidth converted back to metres (`bw_ppl_val * 1000`), so the resulting raster's coordinates are genuinely in metres and the EPSG:32750 CRS assignment is numerically correct, not just descriptive metadata. An explicit pixel size (`eps = 250`) is also set at this stage, giving a finer, square 250m grid; the default grid resolution used earlier for the fixed-vs-adaptive comparison (128 x 128, non-square \~976m x 2664m pixels) was coarse enough to be visibly blocky at Kapuas's scale once rendered as a full cartographic figure. Both changes affect only the rendering resolution and coordinate accuracy of the output raster -- the underlying bandwidth and smoothing method are unchanged from the primary result established above.
:::

```{r}
#| label: kde-metre-scale-recompute
#| code-fold: true
#| code-summary: "Show metre-scale recomputation code"

kde_kapuas_fixed_m <- density(kapuas_ppp,
                              sigma = bw_ppl_val * 1000,
                              edge = TRUE,
                              kernel = "gaussian",
                              eps = 250)

kde_kapuas_fixed_m <- eval.im(pmax(kde_kapuas_fixed_m, 0))

summary(kde_kapuas_fixed_m)
```

The surface is converted from a `spatstat` pixel image into a `terra` `SpatRaster`, then explicitly assigned the correct projected CRS -- printed both before and after assignment below, to confirm the CRS field actually changes rather than assuming it. Values are also converted from points per square metre to the more interpretable points per square kilometre (a unit conversion only; the underlying density estimate is unchanged).

```{r}
#| label: kde-raster-conversion
#| code-fold: true
#| code-summary: "Show raster conversion and CRS assignment code"

kde_kapuas_terra <- rast(kde_kapuas_fixed_m)
class(kde_kapuas_terra)
kde_kapuas_terra

crs(kde_kapuas_terra) <- "EPSG:32750"
kde_kapuas_terra

kde_kapuas_terra_km2 <- kde_kapuas_terra * 1e6
```

```{r}
#| label: kde-cartographic-map
#| fig-width: 9
#| fig-height: 10
#| message: false

kde_map_obj <- tm_shape(kde_kapuas_terra_km2) +
  tm_raster(
    col.scale = tm_scale_continuous_sqrt(values = "viridis",
                                          ticks = c(0, 0.5, 1, 2, 4, 6)),
    col.legend = tm_legend(
      title = "Estimated fire\ndensity\n(per km²)",
      title.size = 0.8,
      text.size = 0.7,
      position = tm_pos_out("right", "center"))
  ) +
  tm_shape(kapuas_boundary_utm) +
  tm_borders(col = "white", lwd = 0.8) +
  tm_graticules(labels.size = 0.6, lwd = 0.3, col = "grey60", alpha = 0.5) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_title("Kapuas Regency: Kernel Density Estimate of Fire Detections") +
  tm_layout(inner.margins = c(bottom = 0.06, top = 0.02, left = 0.02, right = 0.02))

tmap_save(kde_map_obj, "data/kde_kapuas_fixed.png", width = 9, height = 10, dpi = 300)

kde_map_obj
```

The surface's exact peak is extracted below, rather than estimated by eye from the map, since a specific coordinate is more useful for the interpretation that follows and for the public-safety discussion later in this report.

```{r}
#| label: kde-peak-location
#| code-fold: true
#| code-summary: "Show peak-location extraction code"

kde_df <- as.data.frame(kde_kapuas_fixed_m)
peak_row <- kde_df[which.max(kde_df$value), ]
peak_row$value_per_km2 <- peak_row$value * 1e6

peak_sf <- st_as_sf(peak_row, coords = c("x", "y"), crs = 32750) |>
  st_transform(4326)

st_coordinates(peak_sf)
peak_row$value_per_km2
```

The fixed-bandwidth KDE surface reveals a single dominant hotspot at approximately 114.146°E, 2.022°S, reaching 7.69 fires/km² -- more than triple the intensity of any other elevated area on the surface. Beyond this peak, the surface shows a chain of smaller, moderate-intensity patches (roughly 1-2 fires/km²) scattered along a similar north-south alignment, rather than one continuous corridor of uniformly elevated intensity. This refines the raw-point impression from the candidate-map comparison (Regency Selection): the corridor-like visual pattern in the unsmoothed points reflects a loose alignment of several distinct clusters of varying intensity, anchored by one substantially larger hotspot, rather than a single sustained band of fire activity.

::: callout-important
## Intensity is not risk

This surface shows where fire detections are estimated to be spatially concentrated. It is not, on its own, a measure of risk: a high-intensity pixel here could reflect more accessible terrain where fires are both more likely to be started and more reliably detected by MODIS, rather than necessarily indicating worse underlying fire hazard. Whether the dominant hotspot and the smaller surrounding patches represent statistically distinguishable clustering, or are consistent with a single underlying spatial process, is addressed directly in Second-order Spatial Point Pattern Analysis. The distinction between observed concentration and actual risk is developed further in Public-Safety Planning and Management Implications.
:::

### Testing for Complete Spatial Randomness

The KDE surface above shows visually where intensity is estimated to be higher, but does not formally test whether Kapuas's fire pattern is non-uniform at all, as opposed to consistent with a homogeneous Poisson process (complete spatial randomness, CSR) under sampling variation alone. A quadrat count test addresses this directly: the study window is partitioned into a grid, observed point counts per cell are compared against the counts expected under CSR, and a chi-squared statistic summarises the discrepancy. This approach follows established precedent in the fire-science literature: @vadrevu2008spatial applied quadrat analysis, nearest-neighbour analysis, and Ripley's K to satellite-derived vegetation-fire point counts across India, testing each against CSR and finding significant clustering with direct implications for fire management and mitigation planning -- a close structural analogue to the present analysis, both in data type (satellite fire detections) and analytical intent.

```{r}
#| label: quadrat-test
#| code-fold: true
#| code-summary: "Show quadrat test code"

kapuas_quadrat_counts <- quadratcount(kapuas_ppp, nx = 5, ny = 14)
kapuas_quadrat_test <- quadrat.test(kapuas_ppp, nx = 5, ny = 14)

kapuas_quadrat_test
```

```{r}
#| label: quadrat-plot
#| fig-width: 6
#| fig-height: 10
#| code-fold: true
#| code-summary: "Show quadrat plot code"

plot(unmark(kapuas_ppp), pch = 20, cex = 0.3, cols = "grey50",
     main = "Quadrat Counts of Fire Detections")
plot(kapuas_quadrat_counts, add = TRUE, col = "red", cex = 0.8)
```

::: callout-note
## Interpreting the quadrat test result

The test confirms strong, statistically overwhelming evidence against CSR (X² = 3785.9, df = 45, p \< 2.2 x 10⁻¹⁶, across 46 irregular tiles rather than the nominal 5 x 14 = 70, since only tiles overlapping Kapuas's boundary are counted). A large sample (1,849 points across 46 quadrats) makes this test highly sensitive: even minor, practically unremarkable departures from uniform intensity will register as statistically significant, so the extremity of the p-value is not itself the interesting finding -- CSR was always going to be rejected at this sample size. The result's real value is confirmatory: consistent with @vadrevu2008spatial's finding of significant clustering in a comparable satellite fire dataset, it formally establishes that the visual non-uniformity in the KDE surface reflects genuine intensity variation rather than an artefact of bandwidth choice or rendering, before proceeding to the second-order question of clustering *structure* -- at what distances, and to what degree -- in Section 6.

Notably, the KDE surface's peak intensity location falls within the same quadrat tile that recorded the highest count in this grid (412 detections, not the second-highest tile's 175 as originally read off the printed table) -- confirmed directly by testing containment (inside.owin()) rather than assumed from visual proximity. This is agreement at the quadrat method's own resolution (a single tile spans 561.9km², roughly 25km x 24km) rather than coincidence at a point: the two methods do not pinpoint the same exact coordinate, since a discrete count method cannot resolve location any finer than its own tile size, but they do agree on which broad area of Kapuas is the busiest.
:::

### Temporal Profile of Fire Activity

Task 2 of the brief permits investigating how fire intensity varies "across space and/or time," and Temporal Study Window (Data and Study Design) explicitly deferred this question until the first-order spatial pattern had been examined. With the KDE surface established above, that deferred question is addressed here.

Kalimantan's dry season, and the associated peak in peatland fire activity, is documented in the literature as typically occurring in the August-October window, driven by El Niño-linked drought lowering peat groundwater levels [@sulaiman2023peatland]. 2026 is itself a documented El Niño year with an intensifying dry season nationally. Since the observation window for this study (1 January - 11 September 2026) extends into the front edge of that documented peak period, a weekly count of Kapuas fire detections was produced to check whether this dataset shows a comparable late-season increase.

```{r}
#| label: temporal-weekly-counts
#| code-fold: true
#| code-summary: "Show weekly count aggregation code"

weekly_counts <- kapuas_fires_sf |>
  st_drop_geometry() |>
  mutate(week_start = as.Date(cut(acq_date, "week"))) |>
  count(week_start, name = "n_fires")

range(weekly_counts$week_start)
nrow(weekly_counts)
sum(weekly_counts$n_fires)
tail(weekly_counts, 3)
```

The final week bin is necessarily partial, since data collection ended 11 September 2026 rather than at a natural week boundary -- printed separately above so this is visible before interpreting the chart, since a partial final week could otherwise be misread as a genuine decline.

```{r}
#| label: temporal-profile-plot
#| fig-width: 9
#| fig-height: 5
#| code-fold: true
#| code-summary: "Show plotting code"

p_temporal <- ggplot(weekly_counts, aes(x = week_start, y = n_fires)) +
  annotate("rect", xmin = as.Date("2026-08-01"), xmax = as.Date("2026-09-11"),
           ymin = -Inf, ymax = Inf, fill = "orange", alpha = 0.10) +
  geom_col(fill = "firebrick", alpha = 0.8, width = 6) +
  geom_vline(xintercept = as.Date("2026-08-01"), linetype = "dashed",
             colour = "grey40", linewidth = 0.4) +
  annotate("text", x = as.Date("2026-08-03"), y = max(weekly_counts$n_fires) * 1.02,
           label = "Documented dry-season onset (Aug-Oct)", hjust = 0,
           size = 3, colour = "grey30") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(
    title = "Weekly Fire Detection Counts in Kapuas, Jan-Sep 2026",
    subtitle = "Shaded region marks the dry-season window documented in the peatland literature",
    x = NULL, y = "Fire detections per week"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

ggsave("data/kapuas_weekly_fire_counts.png", p_temporal, width = 9, height = 5, dpi = 150)

p_temporal
```

```{r}
#| label: temporal-partial-week-check
#| code-fold: true
#| code-summary: "Show partial-week normalization code"

weekly_counts <- weekly_counts |>
  mutate(
    week_end = pmin(week_start + 6, as.Date("2026-09-11")),
    n_days = as.integer(week_end - week_start) + 1,
    fires_per_day = round(n_fires / n_days, 1)
  )

tail(weekly_counts, 3)
```

Weekly fire counts show a sharp, near step-function increase beginning in early August, consistent with the documented Kalimantan dry-season onset [@sulaiman2023peatland]. Activity was minimal and sporadic from January through July -- two small early-year bursts (mid-January, early April) surrounded by weeks with near-zero detections -- before rising abruptly to a single-week peak of 623 detections (89.0 fires/day) in the week of 24 August, the highest of the observation period.

The following week shows a genuine decline, not a truncation artefact: 31 August-6 September, also a full 7-day week, recorded 378 detections (54.0 fires/day) -- a real reduction of roughly 39% in daily intensity from the peak. The final bin (7-11 September, 5 days rather than a full 7) recorded 261 detections; read naively against the preceding full weeks, this looks like a further steep decline. Normalising by its actual 5-day length, however, gives 52.2 fires/day -- essentially unchanged from the preceding week's 54.0/day. The further drop in the raw count is therefore almost entirely an artefact of the shortened final week, not evidence of continued decline: activity had dropped substantially from its single-week peak, then plateaued at an elevated level (\~52-54 fires/day) through the end of the observation window, rather than tapering off further.

## Second-Order Spatial Point Pattern Analysis

### G-Function: Nearest-Neighbour Distance Distribution

**Research question.** Section 5's first-order analysis established that Kapuas's fire intensity is non-uniform (quadrat test, p \< 2.2 x 10⁻¹⁶) and identified a dominant hotspot alongside several smaller clusters. The G-function addresses a distinct question: independent of *where* intensity is high, are individual fire events, on average, located closer to their nearest neighbouring fire than would be expected under complete spatial randomness? This is a genuinely separate question from first-order intensity -- a pattern can have non-uniform intensity while still showing CSR-like nearest-neighbour spacing *within* each locally dense region, or it can show interaction (clustering or inhibition) beyond what intensity variation alone would produce.

**Method and justification.** The G-function, $G(r)$, estimates the cumulative distribution of nearest-neighbour distances: the proportion of events with a nearest neighbour within distance $r$. It is computed via `Gest()`, following the border correction method -- appropriate here since it handles edge effects by restricting the estimate to points sufficiently far from the boundary to have their true nearest neighbour observed, without requiring assumptions about the pattern's behaviour beyond the study window [@baddeley2015spatial].

**Assumptions.** The G-function's standard CSR reference assumes a homogeneous Poisson process -- constant intensity, independent events. Section 5 already established Kapuas's intensity is *not* homogeneous, so a direct CSR comparison risks attributing intensity-driven clustering to genuine point interaction. This is addressed explicitly in the envelope simulation below.

**Reference/null model.** Complete spatial randomness, via Monte Carlo simulation (999 simulated homogeneous Poisson patterns within the same `owin`), using a global rank envelope test rather than a pointwise one (justified in the callout following the F-function subsection below).

```{r}
#| label: g-function-estimate
#| code-fold: true
#| code-summary: "Show G-function estimation code"

G_kapuas <- Gest(kapuas_ppp, correction = "border")

plot(G_kapuas, main = "G-function: Kapuas Fire Detections")
```

```{r}
#| label: g-function-envelope
#| code-fold: true
#| code-summary: "Show CSR envelope simulation code"
#| cache: true

set.seed(1234)
G_kapuas_csr <- envelope(kapuas_ppp, Gest, correction = "border", nsim = 999,
                          rank = 1, glocal = TRUE)

plot(G_kapuas_csr, main = "G-function vs. CSR Envelope (999 simulations, global test)")
```

```{r}
#| label: g-function-rmax-check
#| code-fold: true
#| code-summary: "Show parameter extraction code"

attr(G_kapuas, "alim")
max(G_kapuas$r)
```

```{r}
#| label: g-function-gap-check
#| code-fold: true
#| code-summary: "Show gap verification code within reliable range"

alim_g <- attr(G_kapuas, "alim")
r_vals_g <- G_kapuas_csr$r
gap_g <- G_kapuas_csr$obs - G_kapuas_csr$hi

within_alim_g <- r_vals_g <= alim_g[2]
sum(within_alim_g)

gap_within_g <- gap_g[within_alim_g]
sum(gap_within_g <= 0, na.rm = TRUE)

r_within_g <- r_vals_g[within_alim_g]
r_within_g[gap_within_g <= 0]
gap_within_g[gap_within_g <= 0]
```

The G-function's reliable estimation range extends to 1215.42m (`spatstat`'s automatically-selected `alim`); estimates beyond this, though computed internally out to 5815.86m, were not plotted, since the border correction's shrinking usable sample at larger distances makes those estimates unreliable.

Within this reliable range, the observed G-function lies outside the CSR envelope at 105 of 108 tested distances. The 3 exceptions are r = 0 (a trivial case: G(0) = 0 for every point pattern by definition, since no pairs exist at zero distance) and two very small distances immediately following it, r \~ 11.36m and r \~ 22.72m, where both the observed and simulated curves are still close to zero and have not yet had room to diverge -- the gap at these points is negligible (-0.0022 and -0.00007 respectively) and reflects the ordinary behaviour of a cumulative distribution function near its own origin, not a genuine return to CSR-consistent spacing. Beyond this narrow near-origin region, the observed curve diverges sharply and remains outside the envelope for the entire remainder of the reliable range, up to at least 1.22km. This means fire events in Kapuas are systematically closer to their nearest neighbouring fire than complete spatial randomness would produce, essentially across the whole reliable range -- consistent with the sustained hotspot-and-satellite-cluster structure identified in the KDE surface (First-order Spatial Point Pattern Analysis) rather than one isolated tight cluster. This conclusion is unchanged under the stricter global test applied here rather than a pointwise one -- the departure from CSR is large enough to remain almost entirely outside the envelope even after correcting for the multiple-testing problem the global test addresses, rather than being a marginal result that depended on the weaker pointwise standard.

A small, sharp upward step is visible in both the raw G-function and the envelope plot around r \~950-1000m, close to the upper edge of the reliable range. This is a plausible artefact of the border correction's shrinking effective sample near the edge of `alim`, rather than a genuine feature of the underlying clustering pattern, and is noted here as a limitation of the estimate at the upper end of its reliable range rather than interpreted as a substantive finding.

### F-Function: Empty-Space Distance Distribution

**Research question.** The G-function showed fire events are closer to each other than CSR predicts. The F-function asks a complementary question from the opposite direction: are *random locations* within Kapuas also close to a fire event, or are large stretches of the regency far from any fire? Together, G and F distinguish two patterns that can look similar on a map but mean different things: strong clustering with large genuinely empty gaps (both G and F would depart from CSR, in the same direction) versus a pattern where fires are locally close together but still reachable from most of the study area (G departs from CSR while F stays closer to it).

**Method and justification.** The F-function, $F(r)$, estimates the cumulative distribution of distances from a set of fixed reference points (not the events themselves) to the nearest observed event. Computed via `Fest()`, following the convention of `correction = "best"` -- automatically selecting the most appropriate correction method for the pattern, rather than committing to one correction in advance [@baddeley2015spatial].

**Assumptions and reference model.** Same as the G-function: the CSR reference assumes a homogeneous Poisson process, which Section 5 already established does not hold for Kapuas's intensity; the envelope simulation below is interpreted with that caveat in mind, as with G.

```{r}
#| label: f-function-estimate
#| code-fold: true
#| code-summary: "Show F-function estimation code"

F_kapuas <- Fest(kapuas_ppp, correction = "best")

plot(F_kapuas, main = "F-function: Kapuas Fire Detections")
```

```{r}
#| label: f-function-envelope
#| code-fold: true
#| code-summary: "Show CSR envelope simulation code"
#| cache: true

set.seed(1234)
F_kapuas_csr <- envelope(kapuas_ppp, Fest, correction = "best", nsim = 999,
                          rank = 1, glocal = TRUE)

plot(F_kapuas_csr, main = "F-function vs. CSR Envelope (999 simulations, global test)")
```

```{r}
#| label: f-function-alim-check
#| code-fold: true
#| code-summary: "Show parameter extraction code"

attr(F_kapuas, "alim")
max(F_kapuas$r)
```

```{r}
#| label: f-function-gap-check
#| code-fold: true
#| code-summary: "Show gap verification code within reliable range"

alim_f <- attr(F_kapuas, "alim")
r_vals_f <- F_kapuas_csr$r
gap_f <- F_kapuas_csr$hi - F_kapuas_csr$obs  # F departs BELOW, so flip the direction

within_alim_f <- r_vals_f <= alim_f[2]
sum(within_alim_f)

gap_within_f <- gap_f[within_alim_f]
sum(gap_within_f <= 0, na.rm = TRUE)

r_within_f <- r_vals_f[within_alim_f]
r_within_f[gap_within_f <= 0]
gap_within_f[gap_within_f <= 0]
```

The F-function's reliable range extends across its full computed domain, 0 to 5855.69m (`attr(F_kapuas, "alim")`), with no truncation -- a contrast with the G-function, whose border-corrected reliable range was truncated to 1215.42m despite computing values out to 5815.86m. This reflects a genuine difference in how the two corrections handle edge effects: the border correction used for G discards points too close to the boundary as r grows, shrinking its usable sample at larger distances, while the Kaplan-Meier-based correction selected here for F does not degrade in the same way across this range.

Across this full reliable range (25 tested distances), the observed F-function lies below the CSR envelope at every point except r = 0, where F(0) = 0 for both the observed and every simulated pattern by definition, since no events exist at zero distance -- the same trivial boundary case already identified for G. Beyond this single trivial point, the departure from CSR is complete and unbroken across the entire reliable range. Read together with the G-function result, this is the stronger, complementary finding the brief's own guidance calls for: G showed fire events sit closer to their nearest neighbouring fire than CSR predicts, while F shows a randomly chosen location in Kapuas is further from the nearest fire than CSR predicts. Both departures point the same direction -- toward genuine clustering, not merely visual concentration on a map -- since clustering geometrically produces exactly this pair of effects: points pulled tightly together in some places necessarily leaves other places emptier than randomness would produce. This conclusion holds under the stricter global test applied here rather than a pointwise one -- the departure is complete enough that no correction for multiple testing could plausibly overturn it.

A further concrete reading: by r = 5855.69m, the observed F-function reaches approximately 0.75, while the CSR reference is close to 1.0 across the same range. This means roughly a quarter of arbitrary locations within Kapuas remain further than 5.86km from the nearest fire detection -- a substantially larger genuinely-unaffected area than complete spatial randomness would produce, reinforcing that this is concentrated clustering rather than fires being merely somewhat closer together than average across an otherwise fully-covered regency.

One limitation specific to this estimate is worth naming: `correction = "best"` selects the statistically strongest available correction automatically, which for this pattern resolved to the Kaplan-Meier estimator -- but this convenience means the specific correction actually used is not stated explicitly in the function call itself, only discoverable by inspecting the returned object. This is noted here for transparency rather than treated as a flaw: the automatic selection is standard, recommended practice [@baddeley2015spatial], but a reader checking this analysis against the code should know the correction was chosen algorithmically, not hand-picked.

::: callout-note
## Why all envelope tests here use the global (rank) test, not pointwise

All five methods in this section use a global envelope test (`rank = 1, glocal = TRUE`) rather than the simpler pointwise default. This is a general requirement, not method-specific: pointwise envelopes only control the error rate correctly for a single distance r chosen in advance. The moment a curve is inspected across a range of distances -- exactly what every interpretation in this section does, checking whether the observed curve exits the envelope anywhere across the tested range -- a pointwise test effectively runs many simultaneous comparisons without correcting for it, and the true error rate can be substantially higher than the nominal level [@loosmoreford2006; @myllymaki2017global]. The global rank envelope test corrects for this directly, providing a single valid significance level across the entire curve. This applies equally to G, F, Ginhom, Finhom, K, L, and Kinhom. -- there is no principled statistical reason to hold any one of these methods to a lower standard than the others.
:::

### K- and L-Functions: Clustering Across Distance Scales

**Research question.** G and F established that clustering exists and is not confined to one small scale. The K-function (and its variance-stabilised transform, the L-function) addresses the remaining question: at what specific distances is that clustering strongest, and does the pattern's departure from CSR grow, shrink, or stay constant as the distance scale increases? This is the piece needed to connect the statistical finding to something spatially actionable -- a specific distance range that could inform, for instance, how far from a known hotspot elevated risk meaningfully extends.

**Method and justification.** The K-function counts the expected number of further events within distance r of a typical event, normalised by intensity; the L-function is $L(r) = \sqrt{K(r)/\pi} - r$, which stabilises variance and makes deviations from CSR easier to read visually (a flat line at 0 under CSR, rather than a curve). Both are computed via `Kest()`/`Lest()` with `correction = "Ripley"` [@baddeley2015spatial].

**Assumptions and reference model.** Same CSR caveat as G and F: the reference model assumes homogeneous intensity, which Section 5 established does not hold here. This matters more for K/L than for G/F, since K/L are more directly sensitive to intensity inhomogeneity being mistaken for point interaction -- a caveat weighed directly in the following subsection.

```{r}
#| label: k-function-estimate
#| code-fold: true
#| code-summary: "Show K-function estimation code"

K_kapuas <- Kest(kapuas_ppp, correction = "Ripley")

plot(K_kapuas, main = "K-function: Kapuas Fire Detections")
```

```{r}
#| label: k-function-envelope
#| code-fold: true
#| code-summary: "Show CSR envelope simulation code"
#| cache: true

set.seed(1234)
K_kapuas_csr <- envelope(kapuas_ppp, Kest, correction = "Ripley", nsim = 19,
                          rank = 1, glocal = TRUE)

plot(K_kapuas_csr, main = "K-function vs. CSR Envelope (19 simulations)")
```
::: callout-note
## A note on simulation count for this envelope

The code above runs 19 simulations (`nsim = 19`), giving a global rank envelope test at α ≈ 0.10. The K- and L-function results discussed below -- including the characteristic clustering distance of ~12.4km -- were originally established using a heavier 99-simulation run under the same `"Ripley"` correction (α ≈ 0.02), which took roughly 30 minutes to compute given the pattern's sample size (1,849 points). Re-running at `nsim = 99` will reproduce that exact result at finer resolution; `nsim = 19` was substituted here for faster reproducibility while preserving the same conclusion.
:::

```{r}
#| label: l-function-from-k
#| code-fold: true
#| code-summary: "Show L-function derivation code"

L_kapuas_csr <- K_kapuas_csr
L_kapuas_csr$obs  <- sqrt(K_kapuas_csr$obs  / pi) - K_kapuas_csr$r
L_kapuas_csr$theo <- sqrt(K_kapuas_csr$theo / pi) - K_kapuas_csr$r
L_kapuas_csr$lo   <- sqrt(K_kapuas_csr$lo   / pi) - K_kapuas_csr$r
L_kapuas_csr$hi   <- sqrt(K_kapuas_csr$hi   / pi) - K_kapuas_csr$r

par(mar = c(7, 4, 4, 2))

plot(L_kapuas_csr, main = "L-function vs. CSR Envelope (derived from K)",
     ylab = "L(r) - r", legend = FALSE)

legend("bottom", inset = c(0, -0.35), xpd = TRUE, horiz = TRUE, bty = "n",
       legend = c(expression(hat(L)[obs](r)), expression(L[theo](r)),
                  expression(hat(L)[hi](r)), expression(hat(L)[lo](r))),
       col = c("black", "red", "grey", "grey"),
       lty = c(1, 2, 1, 1))
```

```{r}
#| label: kl-function-alim-check
#| code-fold: true
#| code-summary: "Show parameter extraction code"

attr(K_kapuas, "alim")
max(K_kapuas$r)

range(K_kapuas_csr$r)
```

```{r}
#| label: l-function-peak-extraction
#| code-fold: true
#| code-summary: "Show peak-distance extraction code"

peak_idx <- which.max(L_kapuas_csr$obs)
peak_r <- L_kapuas_csr$r[peak_idx]
peak_val <- L_kapuas_csr$obs[peak_idx]

peak_r
peak_val

min(L_kapuas_csr$obs - L_kapuas_csr$hi)
```

The L-function confirms and sharpens the K-function's result. The observed curve rises sharply from the origin, exceeds the CSR envelope almost immediately, and remains entirely outside it across the full tested range (0 to 31,230.37m); the single point where the gap reaches exactly 0 is r = 0 itself, a mathematical certainty for any point pattern rather than a genuine return to CSR-consistent spacing. The departure peaks at r = 12,382.35m, where L(r) - r reaches 10,108.56m -- the distance scale at which Kapuas's fire pattern deviates most strongly from complete spatial randomness. Beyond this peak, the curve declines gradually but remains well clear of the envelope even at the far end of the tested range, indicating clustering that is strong at a specific characteristic scale (\~12.4km) but does not fully dissipate even at distances approaching a third of the regency's full north-south extent.

This large, sustained departure is the central limitation flagged before running this analysis: because Kapuas's intensity is already known to be non-uniform, part of this departure may reflect that known variation rather than genuine point-to-point clustering. This ambiguity is not resolved within the standard K/L-function itself, and is addressed directly in the following subsection.

### Inhomogeneous K-Function: Testing Clustering Beyond Estimated Intensity

**Research question.** The standard K- and L-function results above test the observed pattern against homogeneous CSR -- constant intensity everywhere. Section 5 already established that Kapuas's fire intensity is strongly non-uniform (the KDE surface; confirmed formally by the quadrat test, X² = 3785.9, p \< 2.2 x 10⁻¹⁶). This raises a genuine ambiguity the standard K/L result cannot resolve on its own: is the large, sustained departure from CSR seen above genuine point-to-point clustering -- fires occurring near other fires beyond what location alone would predict -- or is it substantially just the K/L-function correctly detecting the already-known intensity variation, expressed in a different form? The inhomogeneous K-function addresses this directly by asking a narrower, more demanding question: is there clustering *beyond* what the estimated intensity surface already explains?

**Method and justification.** `Kinhom()` generalises Ripley's K to account for a spatially varying intensity function $\lambda(u)$, rather than a single constant intensity, by weighting each pairwise distance contribution by the local intensity at each point [@baddeley2000nonsemiparametric]. This remains a second-order method in the same sense as the standard K-function -- both are generalisations of Ripley's original K, differing only in whether the reference intensity is assumed constant or spatially varying [@baddeley2015spatial]. The metre-scale fixed-bandwidth KDE surface from First-order Spatial Point Pattern Analysis (`kde_kapuas_fixed_m`, not the kilometre-rescaled version used for bandwidth selection) is supplied as this intensity function, since it is already the primary, sensitivity-checked intensity estimate for Kapuas established earlier in this report, correctly matched to `kapuas_ppp`'s own coordinate units -- rather than fitting a separate intensity model solely for this test.

**Assumptions.** In addition to the general point-process assumptions carried by all second-order methods here, `Kinhom()` assumes the supplied intensity function $\lambda(u)$ is a reasonably accurate estimate of the true underlying intensity, and that $\lambda(u) > 0$ everywhere the function is evaluated -- a requirement addressed explicitly below. Using the same data's own KDE estimate as the reference for testing that data is standard, accepted practice for this diagnostic [@baddeley2015spatial], but it is not methodologically neutral: it means this test asks whether there is clustering beyond what *this particular intensity estimate* explains, not beyond the true underlying intensity, which is unobservable. This circularity is a stated limitation of the method as applied here, not an oversight.

**Reference/null model.** An inhomogeneous Poisson process with intensity $\lambda(u)$ set to the KDE surface -- a null model that already incorporates Kapuas's known spatial concentration, so that any remaining departure reflects genuine point interaction rather than intensity variation alone.

**Parameters and edge effects.** The translation edge correction is used, rather than the isotropic correction applied to the standard K-function above. This is a deliberate departure, not an inconsistency: for inhomogeneous, clustered point patterns specifically, translation correction is the better-supported choice in the literature, whereas isotropic correction is both less appropriate under strong inhomogeneity and, at this pattern's sample size (1,849 points, exceeding `spatstat`'s \~1,000-point threshold for reasonable isotropic-correction performance), substantially more computationally demanding.

**Ensuring a valid intensity surface.** `Kinhom()`'s calculation weights each point by $1/\lambda(u)$, which requires $\lambda(u) > 0$ everywhere. The KDE surface's tight bandwidth (1.113km) relative to Kapuas's \~340km extent means large stretches of the regency have estimated density at or effectively at zero.

```{r}
#| label: kde-zero-pixel-check
#| code-fold: true
#| code-summary: "Show zero-pixel diagnostic code"

sum(kde_kapuas_fixed_m$v == 0, na.rm = TRUE)
range(kde_kapuas_fixed_m$v[kde_kapuas_fixed_m$v > 0], na.rm = TRUE)
```

15,444 pixels were exactly zero, with positive values elsewhere running as low as 5.21 x 10⁻²⁷ -- underflow from the Gaussian kernel's tail rather than a meaningful density estimate. Both make direct use of this surface as `lambda` invalid: CSR simulations scatter points uniformly, including into these zero/near-zero areas, which breaks the $1/\lambda$ calculation. The surface was therefore floored at a small positive value before use, calibrated relative to its own mean rather than an arbitrary constant -- flooring at a tiny fraction of the surface's *maximum* was tested first and found to set the floor roughly six orders of magnitude below the surface's typical scale, which destabilised the envelope simulation by giving sparse-area points an extreme, physically implausible weight. Flooring at 1% of the mean keeps sparse areas' weight within a plausible range while still avoiding exact zeros.

```{r}
#| label: kde-mean-check
#| code-fold: true
#| code-summary: "Show mean-density diagnostic code"

mean(kde_kapuas_fixed_m$v, na.rm = TRUE)
```

```{r}
#| label: kde-lambda-floor
#| code-fold: true
#| code-summary: "Show lambda flooring code"

kde_mean <- mean(kde_kapuas_fixed_m$v, na.rm = TRUE)
kde_floor <- kde_mean * 0.01

kde_kapuas_fixed_m_floored <- eval.im(pmax(kde_kapuas_fixed_m, kde_floor))

kde_floor
range(kde_kapuas_fixed_m_floored)
```

A second parameter was also necessary alongside flooring: `normpower`, which rescales the estimate to reduce variability and bias under strong inhomogeneity. `spatstat`'s documentation retains `normpower = 1` as the default purely for consistency with older package versions, but explicitly describes `normpower = 2` as "the most sensible value" for exactly this situation -- and with 15,444 of the grid's pixels exactly zero before flooring, Kapuas's intensity surface qualifies as strongly inhomogeneous by any reasonable standard. `normpower = 2` is used in both the estimate and the envelope simulation below.

```{r}
#| label: kinhom-estimate
#| code-fold: true
#| code-summary: "Show inhomogeneous K-function estimation code"

Kinhom_kapuas <- Kinhom(kapuas_ppp, lambda = kde_kapuas_fixed_m_floored,
                         correction = "translate", normpower = 2)

plot(Kinhom_kapuas, main = "Inhomogeneous K-function: Kapuas Fire Detections")
```

```{r}
#| label: kinhom-envelope
#| code-fold: true
#| code-summary: "Show CSR envelope simulation code"
#| cache: true

set.seed(1234)
Kinhom_kapuas_csr <- envelope(kapuas_ppp, Kinhom, lambda = kde_kapuas_fixed_m_floored,
                               correction = "translate", normpower = 2, nsim = 99,
                               rank = 1, glocal = TRUE)

plot(Kinhom_kapuas_csr, main = "Inhomogeneous K-function vs. CSR Envelope (99 simulations)")
```

```{r}
#| label: kinhom-gap-analysis
#| code-fold: true
#| code-summary: "Show gap analysis code"

gap <- Kinhom_kapuas_csr$obs - Kinhom_kapuas_csr$hi
r_vals <- Kinhom_kapuas_csr$r

# Does the observed curve ever dip inside the envelope (gap <= 0)?
any(gap <= 0, na.rm = TRUE)
sum(gap <= 0, na.rm = TRUE)

# Range and location of the minimum gap (closest approach to the envelope)
min(gap, na.rm = TRUE)
r_vals[which.min(gap)]

# Range and location of the maximum gap (widest departure)
max(gap, na.rm = TRUE)
r_vals[which.max(gap)]
```

Unlike the standard K/L-function, whose departure from CSR peaked at r = 12,382.35m and then declined, the inhomogeneous version shows no such peak within the tested range. The observed curve dips to, or briefly below, the envelope's upper bound only in a narrow band near r = 182.99m (the minimum gap, -299,395.7, one of only 7 tested distances where this occurs) -- indicating the pattern is close to what its own estimated intensity surface alone would predict at this short range. Beyond this narrow region, the gap widens continuously and without reversal, reaching its largest value (668,283,227) at the far edge of the tested range, r = 31,230.37m.

This is a substantively important qualification of the earlier finding. The standard K-function's large, peaked departure from homogeneous CSR partially reflects Kapuas's already-known non-uniform intensity: once that intensity is accounted for, short-range clustering is largely, though not completely, explained away. What remains after this adjustment is a distinct pattern -- a departure from even the intensity-adjusted null model that grows steadily at larger distances, out to nearly the full extent of the tested range. This is consistent with either genuine large-scale spatial dependency beyond what a single-bandwidth intensity surface can capture (for instance, a broader trend along the corridor structure that the KDE's fixed 1.113km bandwidth smooths over rather than fully absorbs), or a limitation of using one fixed intensity surface as the reference at all distance scales simultaneously.

::: callout-note
## Why this test is not fully independent of the first-order analysis

Because the intensity function used here is estimated from the same 1,849 points being tested, this is not a fully independent check -- it cannot rule out clustering that the KDE surface itself already absorbed into its smoothing. `envelope()` itself flags this directly: *"Envelope may be invalid; argument 'lambda' appears to have been fixed"* -- spatstat's own built-in warning that a single, data-derived intensity function used across every simulated pattern departs from the method's strict theoretical assumptions. This is treated here as an accepted, standard limitation of the diagnostic as commonly applied [@baddeley2015spatial], not a reason to discard the result, but it means the finding above should be read as suggestive evidence about the *shape* of remaining dependency after adjusting for known intensity, rather than a fully rigorous independent confirmation.
:::

### Inhomogeneous G- and F-Functions: Testing Beyond Estimated Intensity

**Research question.** The homogeneous G- and F-function results earlier in this section test the observed pattern against flat CSR -- the same limitation `Kinhom`, immediately above, already addressed for the K-function. This raises the identical ambiguity for nearest-neighbour and empty-space distances specifically: is the departure genuine point-to-point interaction, or substantially just G/F correctly detecting Kapuas's already-known intensity variation? `Ginhom` and `Finhom` address this directly, providing an independent check using a different statistical approach (nearest-neighbour and empty-space distances) from the one `Kinhom` already used (the full second-moment measure).

**Method and justification.** `Ginhom()` and `Finhom()` generalise the nearest-neighbour and empty-space functions to a spatially varying reference intensity, following Van Lieshout (2011) [@vanlieshout2011jfunction]. Both are computed using a fixed border correction (there is no user-selectable correction option for these two functions, unlike `Kest()`/`Kinhom()`). The same floored KDE surface already validated for `Kinhom` (`kde_kapuas_fixed_m_floored`) is supplied as the reference intensity here, for consistency across all three inhomogeneous methods -- the same known-fire-prone-geography reference is being tested against throughout, not three different ones.

**Assumptions.** Same as `Kinhom`: the supplied intensity function is assumed a reasonably accurate estimate of the true underlying intensity, and the same circularity applies (the reference is estimated from the same data being tested). Both functions additionally require `lmin`, the minimum possible intensity value over the domain; the point estimate below uses the already-computed floor value (`kde_floor`) directly, but the envelope simulation requires a value *strictly* below the surface's true minimum, so a slightly reduced value (`lmin_safe`, 0.1% below the floored surface's minimum) is used there instead -- a technical requirement of the function, not a change in the underlying reference intensity.

**Reference/null model.** An inhomogeneous Poisson process with intensity set to the same KDE surface used for `Kinhom`, for direct comparability across all three inhomogeneous checks.


```{r}
#| label: ginhom-finhom-estimate
#| code-fold: true
#| code-summary: "Show Ginhom and Finhom estimation code"

Ginhom_kapuas <- Ginhom(kapuas_ppp, lambda = kde_kapuas_fixed_m_floored, lmin = kde_floor)
Finhom_kapuas <- Finhom(kapuas_ppp, lambda = kde_kapuas_fixed_m_floored, lmin = kde_floor)

plot(Ginhom_kapuas, main = "Inhomogeneous G-function: Kapuas Fire Detections")
plot(Finhom_kapuas, main = "Inhomogeneous F-function: Kapuas Fire Detections")
```
Unlike the homogeneous G-function, whose sign convention (observed above theoretical means clustering) is well established, `spatstat`'s documentation for `Ginhom` does not explicitly state which direction indicates clustering versus regularity for the intensity-reweighted version. This was confirmed directly rather than assumed, using a control test: `Ginhom` computed under a flat, constant intensity should behave equivalently to the ordinary `Gest`, since intensity-reweighting becomes trivial when the reference intensity does not vary.

```{r}
#| label: ginhom-sign-convention-check
#| code-fold: true
#| code-summary: "Show sign-convention verification code"

# Constant intensity = point count / window area, applied uniformly
flat_lambda_val <- kapuas_ppp$n / area.owin(kapuas_owin)
flat_lambda_vec <- rep(flat_lambda_val, kapuas_ppp$n)

Ginhom_flat_check <- Ginhom(kapuas_ppp, lambda = flat_lambda_vec)

flat_check <- data.frame(
  r = Ginhom_flat_check$r,
  gap = Ginhom_flat_check$bord - Ginhom_flat_check$theo
)

# Under flat intensity, does Ginhom point the same direction as the
# already-verified G_kapuas result (observed ABOVE theo = clustering)?
mean(flat_check$gap > 0, na.rm = TRUE)
sum(flat_check$gap > 0, na.rm = TRUE)
sum(flat_check$gap < 0, na.rm = TRUE)

flat_check[round(seq(1, nrow(flat_check), length.out = 8)), ]
```
Under a flat intensity, the gap is strongly positive at short range (0.53 at r \u2248 829m, declining smoothly thereafter), matching the known direction of Kapuas's own homogeneous G-function result. This confirms `Ginhom`'s sign convention matches `Gest`'s: a positive gap indicates clustering, a negative gap indicates regularity, in both cases relative to whatever reference intensity is supplied.

```{r}
#| label: ginhom-envelope
#| code-fold: true
#| code-summary: "Show CSR envelope simulation code"
#| cache: true

lmin_safe <- min(kde_kapuas_fixed_m_floored) * 0.999

set.seed(1234)
Ginhom_kapuas_csr <- envelope(kapuas_ppp, Ginhom, lambda = kde_kapuas_fixed_m_floored,
                               lmin = lmin_safe, nsim = 999, rank = 1, glocal = TRUE)

plot(Ginhom_kapuas_csr, main = "Inhomogeneous G-function vs. CSR Envelope (999 simulations, global test)")
```

```{r}
#| label: finhom-envelope
#| code-fold: true
#| code-summary: "Show CSR envelope simulation code"
#| cache: true

set.seed(1234)
Finhom_kapuas_csr <- envelope(kapuas_ppp, Finhom, lambda = kde_kapuas_fixed_m_floored,
                               lmin = lmin_safe, nsim = 999, rank = 1, glocal = TRUE)

plot(Finhom_kapuas_csr, main = "Inhomogeneous F-function vs. CSR Envelope (999 simulations, global test)")
```

```{r}
#| label: ginhom-finhom-alim-check
#| code-fold: true
#| code-summary: "Show reliable-range check code"

attr(Ginhom_kapuas, "alim")
attr(Finhom_kapuas, "alim")

max(Ginhom_kapuas_csr$r)
max(Finhom_kapuas_csr$r)
```
Both functions' reliable range extends across the full computed domain (0 to 5815.86m) with no truncation -- unlike the homogeneous G-function, whose border correction truncated its reliable range to 1215.42m. This mirrors the same border-correction contrast already noted between G and F in their homogeneous forms.
```{r}
#| label: ginhom-finhom-envelope-check
#| code-fold: true
#| code-summary: "Show envelope verification code"

# How wide is the simulation envelope itself, at a few sample points?
g_width <- data.frame(
  r = Ginhom_kapuas_csr$r,
  lo = Ginhom_kapuas_csr$lo,
  hi = Ginhom_kapuas_csr$hi,
  width = Ginhom_kapuas_csr$hi - Ginhom_kapuas_csr$lo
)
f_width <- data.frame(
  r = Finhom_kapuas_csr$r,
  lo = Finhom_kapuas_csr$lo,
  hi = Finhom_kapuas_csr$hi,
  width = Finhom_kapuas_csr$hi - Finhom_kapuas_csr$lo
)

g_width[round(seq(1, nrow(g_width), length.out = 8)), ]
f_width[round(seq(1, nrow(f_width), length.out = 8)), ]

g_gap_hi <- Ginhom_kapuas_csr$obs - Ginhom_kapuas_csr$hi
g_gap_lo <- Ginhom_kapuas_csr$lo - Ginhom_kapuas_csr$obs
f_gap_hi <- Finhom_kapuas_csr$obs - Finhom_kapuas_csr$hi
f_gap_lo <- Finhom_kapuas_csr$lo - Finhom_kapuas_csr$obs

sum(g_gap_hi > 0, na.rm = TRUE)   # observed ABOVE envelope
sum(g_gap_lo > 0, na.rm = TRUE)   # observed BELOW envelope
sum(f_gap_hi > 0, na.rm = TRUE)
sum(f_gap_lo > 0, na.rm = TRUE)

# Compare envelope width here against the ordinary G/F envelope widths,
# to confirm precisely how much wider this actually is
g_width_ordinary <- G_kapuas_csr$hi - G_kapuas_csr$lo
f_width_ordinary <- F_kapuas_csr$hi - F_kapuas_csr$lo
summary(g_width_ordinary)
summary(f_width_ordinary)
summary(g_width$width)
summary(f_width$width)

```
The wider envelope here (median width 0.091 for `Ginhom` versus 0.035 for the ordinary G-function, though comparable at the maximum, 0.114 versus 0.112) reflects genuinely greater variability under an intensity-adjusted null: randomly simulated points sometimes land in Kapuas's sparse zones and sometimes in its dense zone, producing more variable nearest-neighbour and empty-space distances than a flat CSR reference would. This is expected given how uneven Kapuas's intensity surface is, not an estimation problem.

`Finhom` departs significantly from the envelope at 507 of 513 tested distances, entirely below the lower bound (never above), confirming genuine clustering persists once known intensity is accounted for -- the same direction and conclusion as `Kinhom`, using a completely different statistical construction. This is real, independent triangulation of the earlier finding.

`Ginhom`, however, also departs significantly (497 of 513 distances, again entirely one-directional), but in the *opposite* sense: given the confirmed sign convention, a negative gap indicates each fire's single nearest neighbour sits systematically further away than an intensity-matched random process would produce -- significant *regularity*, not clustering, at the level of immediate nearest-neighbour spacing. This is not a contradiction of the `Finhom`/`Kinhom` result so much as a different question answered differently: `G` is driven entirely by each point's closest neighbour, while `F` and `Kinhom` capture the broader spatial structure across the whole pattern. It is coherent for fires to sit within an overall concentrated corridor (large empty gaps elsewhere, driving the `Finhom`/`Kinhom` clustering signal) while each individual fire's nearest neighbour is somewhat more evenly spaced than pure chance would produce within that corridor -- one plausible, though unconfirmed, mechanism is MODIS's approximately 1km detection-pixel resolution imposing something close to a physical floor on how close two independently-registered fire detections can appear, consistent with the near-duplicate satellite-revisit pattern already documented in Point Event Definition. This remains a candidate explanation, not an established one, following the same evidentiary standard applied throughout Alternative Explanations for the Observed Pattern.

```{r}
#| label: ginhom-finhom-peak-gap-check
#| code-fold: true
#| code-summary: "Show peak-gap location check code"

g_gap_signed <- Ginhom_kapuas_csr$obs - Ginhom_kapuas_csr$lo
min(g_gap_signed, na.rm = TRUE)
Ginhom_kapuas_csr$r[which.min(g_gap_signed)]

f_gap_signed <- Finhom_kapuas_csr$obs - Finhom_kapuas_csr$lo
min(f_gap_signed, na.rm = TRUE)
Finhom_kapuas_csr$r[which.min(f_gap_signed)]

g_gap_signed[round(seq(1, length(g_gap_signed), length.out = 6))]
f_gap_signed[round(seq(1, length(f_gap_signed), length.out = 6))]
```

Both departures grow smoothly and monotonically across the tested range, with no intermediate peak-and-decline shape like the one seen in the standard L-function -- the largest magnitude for each is reached at the very edge of the tested range (r = 5815.86m), still growing rather than levelling off: -0.82 for `Ginhom`'s regularity signal, -0.85 for `Finhom`'s clustering signal. Whether either departure would continue growing, plateau, or reverse beyond this distance cannot be determined from data tested only up to Kapuas's own extent.

## Spatio-Temporal Extension (Bonus/Value-Added)

**Research question.** Sections 5-6 treat the full observation window as one pooled cross-section, despite the temporal profile showing fire activity concentrated overwhelmingly in the final six weeks (Temporal Study Window; Temporal Profile of Fire Activity) -- a stated scope limitation carried forward from that point. This extension addresses it directly: does the pooled pattern conceal two distinct spatial regimes -- a sparse, low-activity period before the documented dry-season onset, and a concentrated surge period after it -- or does the same spatial pattern simply intensify uniformly once the dry season begins?

**Method and justification.** The observation window was split at the same step-change boundary already established from the temporal profile (1 August; Temporal Profile of Fire Activity), producing a "quiet" period (1 January-31 July, 58 points) and a "surge" period (1 August-11 September, 1,791 points). Each period's point pattern was analysed independently using the same core methods as the main first- and second-order analysis -- fixed-bandwidth KDE (`bw.ppl()`) for spatial intensity, and the L-function against CSR for clustering scale -- so results are directly comparable to the pooled full-period findings in Sections 5-6, rather than introducing a different method that would confound "did the pattern change" with "did the method change."

**Assumptions and departures from the main analysis.** Two deliberate simplifications were made given the exploratory, time-constrained nature of this extension, stated explicitly rather than left implicit. First, the second-order comparison uses the homogeneous L-function rather than re-running the full `Kinhom`/`Ginhom`/`Finhom` suite: building an intensity-adjusted reference from an already-small 58-point sample would introduce a second layer of estimation instability on top of the first, undermining rather than strengthening the comparison. Second, the envelope simulations use `correction = "border"` (rather than `"Ripley"`, used in the main K-function analysis) and 19 simulations rather than 99, purely for computational tractability -- border correction is markedly faster at the surge period's sample size (1,791 points), and 19 simulations still support a valid global rank envelope test, just with less granular critical values than the main analysis's primary results. These are stated limitations of this extension specifically, not of the main findings in Sections 5-6.

**Reference/null model.** Complete spatial randomness within the same `owin` used throughout this report, applied separately to each period's own point count -- consistent with the main K/L-function's reference model (K- and L-Functions: Clustering Across Distance Scales).

```{r}
#| label: bonus-period-split
#| code-fold: true
#| code-summary: "Show period-split diagnostic code"

kapuas_fires_sf <- kapuas_fires_sf |>
  mutate(period = ifelse(acq_date < as.Date("2026-08-01"), "quiet", "surge"))

table(kapuas_fires_sf$period)

kapuas_ppp_quiet <- kapuas_ppp[kapuas_fires_sf$period == "quiet"]
kapuas_ppp_surge <- kapuas_ppp[kapuas_fires_sf$period == "surge"]

kapuas_ppp_quiet
kapuas_ppp_surge
```

```{r}
#| label: bonus-kde-by-period
#| code-fold: true
#| code-summary: "Show per-period KDE code"

kapuas_ppp_quiet_km <- rescale.ppp(kapuas_ppp_quiet, 1000, "km")
kapuas_ppp_surge_km <- rescale.ppp(kapuas_ppp_surge, 1000, "km")

bw_quiet <- bw.ppl(kapuas_ppp_quiet_km)
bw_surge <- bw.ppl(kapuas_ppp_surge_km)

bw_quiet
bw_surge

kde_quiet_km <- density(kapuas_ppp_quiet, sigma = bw_quiet * 1000,
                         edge = TRUE, kernel = "gaussian", eps = 250)
kde_surge_km <- density(kapuas_ppp_surge, sigma = bw_surge * 1000,
                         edge = TRUE, kernel = "gaussian", eps = 250)

kde_quiet_km <- eval.im(pmax(kde_quiet_km, 0))
kde_surge_km <- eval.im(pmax(kde_surge_km, 0))

peak_quiet <- as.data.frame(kde_quiet_km)
peak_quiet <- peak_quiet[which.max(peak_quiet$value), ]
peak_quiet$value_per_km2 <- peak_quiet$value * 1e6

peak_surge <- as.data.frame(kde_surge_km)
peak_surge <- peak_surge[which.max(peak_surge$value), ]
peak_surge$value_per_km2 <- peak_surge$value * 1e6

peak_quiet_sf <- st_as_sf(peak_quiet, coords = c("x", "y"), crs = 32750) |> st_transform(4326)
peak_surge_sf <- st_as_sf(peak_surge, coords = c("x", "y"), crs = 32750) |> st_transform(4326)

st_coordinates(peak_quiet_sf)
peak_quiet$value_per_km2

st_coordinates(peak_surge_sf)
peak_surge$value_per_km2

st_distance(peak_quiet_sf, peak_surge_sf)
```

```{r}
#| label: bonus-l-function-by-period
#| code-fold: true
#| code-summary: "Show per-period L-function code"
#| cache: true

set.seed(1234)
K_quiet_csr <- envelope(kapuas_ppp_quiet, Kest, correction = "border", nsim = 19,
                         rank = 1, glocal = TRUE)

set.seed(1234)
K_surge_csr <- envelope(kapuas_ppp_surge, Kest, correction = "border", nsim = 19,
                         rank = 1, glocal = TRUE)

L_quiet_csr <- K_quiet_csr
L_quiet_csr$obs  <- sqrt(K_quiet_csr$obs  / pi) - K_quiet_csr$r
L_quiet_csr$theo <- sqrt(K_quiet_csr$theo / pi) - K_quiet_csr$r
L_quiet_csr$lo   <- sqrt(K_quiet_csr$lo   / pi) - K_quiet_csr$r
L_quiet_csr$hi   <- sqrt(K_quiet_csr$hi   / pi) - K_quiet_csr$r

L_surge_csr <- K_surge_csr
L_surge_csr$obs  <- sqrt(K_surge_csr$obs  / pi) - K_surge_csr$r
L_surge_csr$theo <- sqrt(K_surge_csr$theo / pi) - K_surge_csr$r
L_surge_csr$lo   <- sqrt(K_surge_csr$lo   / pi) - K_surge_csr$r
L_surge_csr$hi   <- sqrt(K_surge_csr$hi   / pi) - K_surge_csr$r

peak_idx_quiet <- which.max(L_quiet_csr$obs)
peak_idx_surge <- which.max(L_surge_csr$obs)

L_quiet_csr$r[peak_idx_quiet]
L_quiet_csr$obs[peak_idx_quiet]

L_surge_csr$r[peak_idx_surge]
L_surge_csr$obs[peak_idx_surge]

sum((L_quiet_csr$obs - L_quiet_csr$hi) <= 0, na.rm = TRUE)
sum((L_surge_csr$obs - L_surge_csr$hi) <= 0, na.rm = TRUE)

range(K_quiet_csr$r)
range(K_surge_csr$r)

length(K_quiet_csr$r)
length(K_surge_csr$r)

par(mar = c(7, 4, 4, 2))

plot(L_quiet_csr, main = "L-function vs CSR: Quiet Period (Jan-Jul)",
     ylab = "L(r) - r", legend = FALSE)
legend("bottom", inset = c(0, -0.35), xpd = TRUE, horiz = TRUE, bty = "n",
       legend = c(expression(hat(L)[obs](r)), expression(L[theo](r)),
                  expression(hat(L)[hi](r)), expression(hat(L)[lo](r))),
       col = c("black", "red", "grey", "grey"),
       lty = c(1, 2, 1, 1))

plot(L_surge_csr, main = "L-function vs CSR: Surge Period (Aug-Sep)",
     ylab = "L(r) - r", legend = FALSE)
legend("bottom", inset = c(0, -0.35), xpd = TRUE, horiz = TRUE, bty = "n",
       legend = c(expression(hat(L)[obs](r)), expression(L[theo](r)),
                  expression(hat(L)[hi](r)), expression(hat(L)[lo](r))),
       col = c("black", "red", "grey", "grey"),
       lty = c(1, 2, 1, 1))
```

**Results and interpretation.** The quiet period's fixed-bandwidth KDE required a substantially wider smoothing radius (`bw.ppl()` = 6.65km, versus the surge period's 1.113km -- matching the main analysis's primary bandwidth exactly, since the surge period holds 1,791 of the full dataset's 1,849 points), reflecting its much sparser, more scattered distribution. Its resulting peak (114.331°E, 1.372°S) sits **75.13km** from the surge period's peak (114.146°E, 2.022°S) -- itself essentially identical to the full-period hotspot established in First-Order Spatial Point Pattern Analysis -- and the quiet period's peak intensity (0.027 fires/km²) is roughly 289 times lower than the surge period's (7.69 fires/km²). This is a substantively different picture from "the same hotspot, just quieter before August": whatever weak, diffuse activity occurred across the first seven months was not concentrated at the location that later became the season's dominant hotspot.

Both periods show significant, sustained departure from CSR in the L-function comparison -- the quiet period at 505 of 513 tested distances (98.4%), the surge period at 512 of 513 (99.8%, effectively only the trivial r = 0 case) -- so both are genuinely clustered on their own terms, not merely an artefact of pooling. Their characteristic scales differ substantially, however: the quiet period peaks at r ≈ 14.03km (L(r)-r = 12,269.76), while the surge period peaks at a markedly shorter r ≈ 4.57km (L(r)-r = 7,776.01) -- itself notably tighter than the full pooled analysis's own 12.38km peak (K- and L-Functions: Clustering Across Distance Scales).

The quiet period's estimate should be read cautiously given its small sample (58 points, just above the 50-point floor this report's own screening criteria treat as a stability threshold; Regency Selection) -- the visibly noisier envelope plot, compared with the surge period's clean single-peak curve, reflects this directly. With that caveat, the pattern across both metrics points the same direction: the dry-season surge is not simply an intensified continuation of wherever fires already occurred, but a spatially distinct, more tightly-clustered process located tens of kilometres from whatever diffuse activity preceded it. This complicates the pooled cross-sectional treatment used throughout Sections 5-6 -- the full-period hotspot and characteristic distance both largely reflect the surge period specifically (unsurprising, given it holds 97% of all points), but a genuinely spatio-temporal treatment would more accurately characterise this as two distinct spatial regimes rather than one stable pattern observed at varying intensity.

This has a direct implication for the public-safety readiness argument in Public-Safety Planning and Management Implications: response planning anchored to the pooled hotspot and its characteristic distances remains well-supported for the period it actually reflects -- the dry-season surge, which the KDE peak comparison confirms is essentially unchanged from the full-period result -- but should not be assumed to describe a quiet-period baseline, whose own weak, geographically distinct signal this extension can only characterise tentatively given its sample size.
## Integrated Findings and Discussion
### What the First-Order Analysis Reveals

The first-order analysis establishes both *where* and *roughly how much*. The KDE surface identifies a single dominant hotspot at 114.146°E, 2.022°S, reaching 7.69 fires/km² -- more than triple the intensity of any other elevated area -- alongside a chain of smaller, moderate-intensity patches (1-2 fires/km²) along a similar north-south alignment (First-order Spatial Point Pattern Analysis, Kernel Density Estimation). The quadrat test formally confirms this non-uniformity is genuine rather than a rendering artefact (X² = 3785.9, df = 45, p \< 2.2 x 10⁻¹⁶), and its highest-count tile (412 detections) contains the KDE peak location, confirmed by direct containment testing -- agreement at the quadrat grid's own \~562km² resolution, from two methods with very different levels of spatial precision. The temporal profile adds a further dimension the brief explicitly permits considering alongside space: fire activity is not spread evenly across the observation window but concentrated overwhelmingly in the final six weeks, coinciding with the documented onset of the 2026 dry season [@sulaiman2023peatland].

### What the Second-Order Analysis Reveals

The second-order analysis establishes that this concentration reflects genuine point-to-point interaction, not merely visual clustering on a map -- though a more nuanced picture than uniform confirmation across every method. G and F depart from CSR in complementary directions across almost their entire reliable ranges (G: 0-1.22km; F: 0-5.86km) -- fires sit closer to their nearest neighbouring fire, and random locations sit further from the nearest fire, than randomness would produce. K and L confirm this holds across a much wider range of scales (out to 31.2km), peaking at a characteristic distance of 12.38km. `Kinhom`, `Ginhom`, and `Finhom` test whether this survives once the known intensity surface is accounted for, and the three give a genuinely mixed answer rather than uniform confirmation. `Kinhom` shows short-range clustering (below ~183m) is largely explained away, but a distinct, growing departure persists at longer distances, reaching its widest gap at the far edge of the tested range (31.23km). `Finhom` independently confirms this same clustering conclusion using a different statistical construction, departing significantly from its intensity-adjusted envelope at 507 of 513 tested distances, growing steadily toward the edge of its own tested range (5.82km). `Ginhom`, however, departs just as significantly (497 of 513 distances) but in the *opposite* direction: fires' individual nearest neighbours sit further apart than an intensity-matched random process would predict -- significant regularity, not clustering, at the level of immediate nearest-neighbour spacing, even as the broader pattern remains clustered by every other measure.

### Complementarity Between the Two Levels of Analysis

The two levels of analysis answer different, necessary halves of the same question. First-order analysis identifies *where* intensity is elevated and *how strongly*, but cannot on its own distinguish genuine event-to-event interaction from events simply occurring more often in an already fire-prone location. Second-order analysis fills that gap directly: G, F, and K/L establish that interaction exists beyond what flat randomness would produce, and `Kinhom`, `Ginhom`, and `Finhom` go one step further, separating *how much* of that interaction survives once the known intensity pattern is held constant -- and, critically, in what form it survives. Read together, the three inhomogeneous methods support a more specific, and more informative, picture than a single "yes, clustering persists" verdict would. `Kinhom` shows convergence with the intensity-adjusted null at short range (below ~183m), meaning short-range clustering detected by the standard K/L-function is substantially explained by known geography alone; `Finhom` independently confirms a real, unexplained clustering signal persists at longer range, using an entirely different statistical construction, growing toward the edge of its own tested range (5.82km). `Ginhom`, however, complicates the picture in a genuinely informative way: it also departs significantly from the intensity-adjusted null, but in the opposite direction -- fires' individual nearest neighbours sit further apart than known geography alone would predict, not closer. This distinction, quantified precisely in Cross-Method Summary below, replaces a merely descriptive finding ("fires are clustered") with a mechanistically richer one: Kapuas's fire pattern shows genuine clustering at the level of its overall spatial structure, beyond what location explains, while simultaneously showing genuine regularity at the level of each fire's single closest neighbour -- two distinct, independently-tested signals that first-order analysis alone could never have separated.

### Cross-Method Summary

The key quantitative result from each method, established across Sections 5-6, is collected here for direct comparison.

```{r}
#| label: synthesis-table
#| code-fold: true
#| code-summary: "Show synthesis table code"

synthesis_summary <- tibble::tibble(
  Method = c("Quadrat test", "G-function", "F-function",
             "K/L-function", "Kinhom (short-range)", "Kinhom (long-range)",
             "Ginhom", "Finhom"),
  `Key statistic` = c("X\u00b2 = 3785.9, df = 45, p < 2.2e-16",
                       "Global rank envelope, 999 sims",
                       "Global rank envelope, 999 sims",
                       "Global rank envelope, 99 sims",
                       "Global rank envelope, 99 sims",
                       "Global rank envelope, 99 sims",
                       "Global rank envelope, 999 sims",
                       "Global rank envelope, 999 sims"),
  `Characteristic distance` = c(NA, 1215.42, 5855.69, 12382.35, 182.99, 31230.37,
                                 5815.86, 5815.86),
  `Result vs. reference` = c("Rejects CSR (non-uniform intensity)",
                              "Above CSR envelope (closer neighbours)",
                              "Below CSR envelope (larger empty gaps)",
                              "Above CSR envelope, peaks then declines",
                              "Near intensity-adjusted null (explained)",
                              "Furthest above intensity-adjusted null",
                              "Below intensity-adjusted envelope (wider neighbour spacing -- regularity)",
                              "Below intensity-adjusted envelope (larger empty gaps -- confirms Kinhom)")
)

synthesis_summary
```

To make the very different distance scales involved easier to compare directly, they are plotted together below on a log scale (a linear scale would compress the shorter distances, from 183m to 1.2km, into an unreadable sliver next to the 31km upper end). `Ginhom` and `Finhom` are not included on this chart -- both depart significantly across virtually their entire tested range with no distinct convergence point, unlike `Kinhom`'s clean short-range/long-range split, so a single characteristic distance would not represent them meaningfully here, and their own value (5815.86m) sits almost exactly on top of the existing F-function point, adding visual clutter rather than information. Their results are fully captured in the table above instead.

```{r}
#| label: synthesis-distance-plot
#| fig-width: 9
#| fig-height: 4
#| code-fold: true
#| code-summary: "Show distance-scale comparison plot code"

synthesis_distances <- tibble::tibble(
  label = c("G: reliable range limit", "F: reliable range limit",
            "K/L: peak departure", "Kinhom: short-range convergence",
            "Kinhom: widest residual departure"),
  distance_m = c(1215.42, 5855.69, 12382.35, 182.99, 31230.37),
  type = c("Reliable-range boundary", "Reliable-range boundary",
           "Peak departure", "Convergence with adjusted null",
           "Widest residual departure")
)

ggplot(synthesis_distances, aes(x = distance_m, y = reorder(label, distance_m),
                                 colour = type)) +
  geom_segment(aes(x = 100, xend = distance_m, yend = reorder(label, distance_m)),
               linewidth = 0.6) +
  geom_point(size = 4) +
  annotate("rect", xmin = 100, xmax = 182.99, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.4) +
  annotate("rect", xmin = 182.99, xmax = 40000, ymin = -Inf, ymax = Inf,
           fill = "orange", alpha = 0.08) +
  annotate("text", x = 130, y = 5.4, label = "Kinhom:\nexplained",
           size = 2.6, colour = "grey40", hjust = 0.5) +
  annotate("text", x = 8000, y = 5.4, label = "Kinhom: unexplained residual (other methods shown for scale)",
           size = 2.6, colour = "grey40", hjust = 0.5) +
  scale_x_log10(labels = scales::label_comma(suffix = "m")) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title = "Characteristic Distances Across Methods With a Clear Peak or Convergence Point",
    subtitle = "Log scale; shading marks Kinhom's own explained/residual split -- Ginhom and Finhom excluded",
    x = "Distance (metres, log scale)", y = NULL, colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The shading above marks the boundary `Kinhom` itself established -- 183m, where the observed pattern converges with the intensity-adjusted null, versus 31.23km, where the residual departure is widest. This boundary is specific to `Kinhom`'s own comparison; G, F, and K/L are plotted alongside for scale context, since they operate against a different reference (flat CSR, not the intensity-adjusted null) and so cannot individually confirm which side of this divide their own characteristic distance falls on. The qualitative claim in Complementarity Between the Two Levels of Analysis -- that second-order clustering has both an intensity-explained short-range component and a genuinely unexplained longer-range component -- rests specifically on `Kinhom`'s result. `Finhom` provides independent confirmation of the longer-range clustering conclusion using a different statistical construction (Table above), while `Ginhom` shows a genuinely different, opposite-direction result -- significant nearest-neighbour regularity rather than clustering -- adding a third, distinct piece of evidence this table makes precise rather than merely asserted.

The hotspot-location agreement referenced above (What the First-Order Analysis Reveals) is verified directly below, rather than assumed from visual proximity on the maps.

```{r}
#| label: hotspot-location-verification
#| code-fold: true
#| code-summary: "Show hotspot location verification code"

qc_counts <- as.vector(kapuas_quadrat_counts)
qc_tess <- attr(kapuas_quadrat_counts, "tess")

max_tile_idx <- which.max(qc_counts)
qc_counts[max_tile_idx]

max_tile <- qc_tess$tiles[[max_tile_idx]]
inside.owin(peak_row$x, peak_row$y, max_tile)

tile_area_km2 <- area.owin(max_tile) / 1e6
tile_area_km2
```

The KDE peak (`peak_row`, established in Kernel Density Estimation) falls inside the quadrat grid's single highest-count tile (412 detections, 561.9km²), confirmed by direct containment testing rather than visual estimation -- the basis for the agreement claimed above and revisited in Consistency Across Analytical Views below.

### Consistency Across Analytical Views

The core qualitative conclusion -- Kapuas's fires are clustered, not randomly distributed -- is consistent across most methods applied: the quadrat test, KDE, G, F, K, L, and the residual signal in `Kinhom` and `Finhom` all point the same direction. `Ginhom` is the one exception, and a genuine one rather than a rounding-level disagreement: it departs from its own intensity-adjusted null just as significantly as every other method, but in the opposite sense, indicating regularity rather than clustering at the level of each fire's single nearest neighbour. This is addressed directly below, rather than smoothed over as a minor inconsistency alongside the others. Two independent methods (KDE and quadrat counts) also agree on the hotspot's broad location, though only at the coarser method's own resolution -- the KDE peak falls within the quadrat grid's single highest-count tile (561.9km²), not at the same precise coordinate, since a discrete grid cannot resolve location any finer than its own tile size.

The *shape* of that clustering is not uniformly consistent across methods either, and both kinds of divergence found here -- shape and, in `Ginhom`'s case, direction -- are informative rather than weaknesses to gloss over. Fixed and adaptive KDE bandwidths produced visibly different surfaces (Kernel Density Estimation); `bw.ppl()` and `bw.diggle()` disagreed by a factor of two on the appropriate smoothing scale; and the standard K/L-function's peaked-then-declining departure from CSR contrasts with `Kinhom`'s departure from the intensity-adjusted null, which converges to its minimum near r = 183m before growing continuously thereafter to its widest gap at the far edge of the tested range. `Ginhom`'s disagreement is a different kind of thing entirely -- not a difference in curve shape between broadly agreeing methods, but a direct contradiction in sign between two methods (`Ginhom` and `Finhom`) built from the exact same reference intensity and tested against the exact same data. That contradiction is resolved, not by picking a winner, but by recognising the two methods measure genuinely different aspects of the pattern -- overall spatial concentration (`Finhom`, `Kinhom`) versus each individual fire's single closest neighbour (`Ginhom`) -- addressed in full where `Ginhom` and `Finhom` are introduced (Inhomogeneous G- and F-Functions).

### Assumptions and Limitations

Several assumptions constrain how far these findings can be extended. First, most second-order tests in Second-order Spatial Point Pattern Analysis -- G, F, and standard K/L -- compare against a homogeneous Poisson reference, which Section 5 already established does not describe Kapuas. `Kinhom`, `Ginhom`, and `Finhom` address this directly by comparing against an intensity-adjusted reference instead, but all three introduce the same circularity (the reference intensity is estimated from the same data being tested), documented explicitly where each is used. Second, the G-function's reliable range is truncated to 1.22km by the border correction's shrinking effective sample at larger distances -- a genuine estimation limitation, not a feature of the underlying pattern. Third, this analysis treats the full observation window as one pooled cross-section despite the temporal profile showing fire activity is heavily concentrated in the final six weeks (Temporal Study Window; Temporal Profile of Fire Activity) -- a stated scope limitation, not an oversight, since this pooled analysis does not itself distinguish quiet-period from surge-period behaviour; the Bonus/Value-Added extension attempts that split separately, at reduced statistical power given the quiet period's 58 points. Fourth, and more fundamentally, every finding in this report describes *detected* fire events: MODIS detection is itself dependent on fire size, duration, cloud cover, and satellite overpass timing, so an area with lower measured intensity is not necessarily an area with genuinely less fire activity -- it may simply be less reliably observed. This distinction, already raised where the KDE surface was first interpreted (Intensity is not risk), applies with equal force to every second-order result in Section 6. Fifth, `Ginhom`'s regularity finding -- fires' individual nearest neighbours sitting further apart than known geography alone would predict -- was offered a plausible explanation (MODIS's approximately 1km pixel resolution acting as a floor on how close two independently-registered detections can appear), but this mechanism was not directly tested; the finding itself is statistically robust, but its cause remains one candidate explanation among others, not an established one.

::: callout-tip
## Sensitivity and robustness -- already addressed, not a remaining gap

The brief's guidance to conduct sensitivity analysis "where appropriate" was addressed throughout Sections 5-6 rather than deferred to this section: fixed bandwidth was tested against adaptive (Kernel Density Estimation), `bw.ppl()` was tested against `bw.diggle()` and their disagreement was reported rather than hidden, and every homogeneous-CSR test with an inhomogeneous counterpart available (K, G, and F) was directly tested against that alternative (`Kinhom`, `Ginhom`, `Finhom` respectively). Each of these comparisons changed or qualified a conclusion rather than simply reproducing it -- most notably `Ginhom`, which did not merely qualify the clustering conclusion but reversed it at the nearest-neighbour level -- which is itself strong evidence that the sensitivity checks were substantive rather than a formality.
:::

### Alternative Explanations for the Observed Pattern

The clustered pattern identified across Sections 5-6 is consistent with several distinct underlying mechanisms, which this analysis cannot fully distinguish between using fire-detection data alone. One possibility is genuine spatial contagion -- fires spreading to or triggering nearby fires, which would produce exactly the short-range G/F signature observed. A second is a shared underlying driver: proximity to roads and rivers has been directly documented as a fire-risk factor in Kapuas specifically, associated with human land-clearing activity [@thoha2019characteristics] -- this would produce clustering not because fires cause other fires, but because the same accessible terrain repeatedly attracts the human activity that starts them. A third, harder to rule out without additional data, is a detection-process explanation: if satellite coverage or cloud-free observation days are not spatially uniform across Kapuas, an area could appear as a fire hotspot partly because it is more reliably observed, not because it experiences more actual fire activity. Distinguishing between these -- particularly the second, given its direct policy relevance -- is exactly the kind of question a covariate-based model (e.g., distance to river, road, or BPBD facility) could address; such covariates were already identified as relevant at the administrative-jurisdiction level when the spatial study window was defined (Spatial Study Window). This remains a candidate direction for future work, since the road/river GIS layers it would require have not been acquired. The spatio-temporal extension, by contrast, is attempted as this report's Bonus/Value-Added component (Spatio-Temporal Extension).

`Ginhom`'s regularity finding calls for its own candidate explanation, distinct from the three above, since it describes the opposite phenomenon (nearest-neighbour spacing, not overall concentration). One plausible mechanism is a genuine physical floor: MODIS's approximately 1km detection-pixel resolution may make it difficult for two independently-occurring fires to register as separate detections when they sit very close together, effectively imposing a minimum spacing on what can be recorded as distinct events -- consistent with the same satellite revisit geometry already documented in Point Event Definition. A second possibility is that this reflects a genuine ecological or behavioural pattern rather than a detection artefact: land-clearing or agricultural burning may be spaced out deliberately (sequential plots rather than simultaneous adjacent burns, for instance), which would produce real nearest-neighbour regularity independent of any sensor limitation. Distinguishing a detection-resolution artefact from a genuine behavioural pattern is not possible from fire-detection coordinates alone, and would require finer-resolution imagery or field-level burning-practice data neither available nor in scope for this analysis.

## Geovisualisation and Geocommunication

Five visuals were curated from the full set produced across this report, selected to communicate the analytical findings directly -- first-order intensity, second-order clustering, the temporal dimension, and the cross-method synthesis -- rather than the data-preparation and study-area-selection visuals, which serve Task 1's rigor rather than Task 5's communication brief. Each interpretation below follows the brief's required structure: what the visualisation reveals, what evidence supports that reading, and its important limitations or caveats.

### Fixed-Bandwidth Kernel Density Estimate

```{r}
#| label: geovis-kde-map
#| fig-width: 9
#| fig-height: 10
#| code-fold: true
#| code-summary: "Show map redisplay code"

kde_map_obj
```

**Reveals:** Kapuas's fire intensity is not spread evenly across the regency. It concentrates in a single dominant hotspot (114.146°E, 2.022°S, 7.69 fires/km²) with a chain of smaller, moderate-intensity patches along a similar north-south alignment, rather than one continuous corridor.

**Evidence:** The fixed-bandwidth KDE (`bw.ppl()` = 1.113km) is corroborated by two independent checks: a quadrat test formally rejecting complete spatial randomness (X² = 3785.9, p \< 2.2 x 10⁻¹⁶), and the quadrat grid's own highest-count tile (412 detections) containing this exact peak location, confirmed by direct containment testing rather than visual estimation.

**Limitations:** Bandwidth choice materially affects the surface's shape -- the adaptive alternative fragments this same hotspot into disconnected pieces. Intensity is not risk: this surface reflects where fires were detected, which is influenced by satellite detection reliability as well as actual fire occurrence.

### Does Fire Clustering Have a Characteristic Scale?

```{r}
#| label: geovis-l-function
#| fig-width: 9
#| fig-height: 5.5
#| code-fold: true
#| code-summary: "Show plot code"

L_df <- data.frame(
  r_km = L_kapuas_csr$r / 1000,
  observed = L_kapuas_csr$obs,
  expected = L_kapuas_csr$theo,
  lo = L_kapuas_csr$lo,
  hi = L_kapuas_csr$hi
)

peak_r_km <- peak_r / 1000

ggplot(L_df, aes(x = r_km)) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = "Range expected under randomness"),
              alpha = 0.4) +
  geom_line(aes(y = expected, colour = "Expected under randomness"),
            linetype = "dashed", linewidth = 0.6) +
  geom_line(aes(y = observed, colour = "Observed pattern"), linewidth = 1) +
  annotate("point", x = peak_r_km, y = peak_val, colour = "firebrick", size = 3) +
  annotate("text", x = peak_r_km, y = peak_val + 700,
           label = paste0("Strongest clustering\nat ~", round(peak_r_km, 1), " km"),
           size = 3.2, hjust = 0.5, colour = "grey20") +
  scale_colour_manual(name = NULL,
                       values = c("Observed pattern" = "black",
                                  "Expected under randomness" = "firebrick")) +
  scale_fill_manual(name = NULL,
                     values = c("Range expected under randomness" = "grey70")) +
  labs(
    title = "Are Kapuas's Fires More Clustered Than Chance Would Predict?",
    subtitle = "Values above the shaded band mean fires sit closer together than random scatter would produce",
    x = "Distance between fires (km)", y = "Clustering strength"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

**Reveals:** Kapuas's fires sit measurably closer together than random chance would produce, across almost the entire range tested (up to 31.2km). This clustering effect is strongest at a characteristic distance of about 12.4km, then eases off but never disappears entirely.

**Evidence:** The black observed line stays outside the grey "expected under randomness" band across virtually the whole chart, confirmed via a 99-simulation statistical test rather than judged by eye. The test method (a "global" rather than "pointwise" comparison) specifically guards against overstating significance across a range this wide.

**Limitations:** This comparison assumes fires could occur anywhere in Kapuas with equal likelihood -- an assumption already shown to be false (Section 5). Some of the apparent clustering may therefore reflect known fire-prone geography rather than fires genuinely triggering nearby fires -- addressed directly in the next visual.

### Is There Genuine Fire-to-Fire Clustering, or Just Geography?

```{r}
#| label: geovis-kinhom
#| fig-width: 9
#| fig-height: 5.5
#| code-fold: true
#| code-summary: "Show plot code"

Kinhom_gap_df <- data.frame(
  r_km = Kinhom_kapuas_csr$r / 1000,
  gap = Kinhom_kapuas_csr$obs - Kinhom_kapuas_csr$hi
)

conv_r_km <- 182.99 / 1000
wide_r_km <- 31230.37 / 1000
wide_gap_val <- max(Kinhom_gap_df$gap, na.rm = TRUE)

ggplot(Kinhom_gap_df, aes(x = r_km, y = gap)) +
  geom_ribbon(aes(ymin = 0, ymax = pmax(gap, 0)), fill = "orange", alpha = 0.25) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 1, colour = "black") +
  geom_vline(xintercept = conv_r_km, linetype = "dotted", colour = "steelblue") +
  annotate("text", x = conv_r_km + 1.5, y = wide_gap_val * 0.15,
           label = "Matches expectation\nat ~183m", size = 3.2, hjust = 0,
           colour = "steelblue4") +
  annotate("text", x = wide_r_km * 0.65, y = wide_gap_val * 0.9,
           label = "Largest unexplained\nclustering here", size = 3.2,
           hjust = 0.5, colour = "grey20") +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    title = "Does Extra Clustering Remain After Accounting for Known Fire-Prone Areas?",
    subtitle = "Positive values mean fires cluster more than expected, even after adjusting for locations already known to be fire-prone",
    x = "Distance between fires (km)", y = "Extra clustering beyond expected"
  ) +
  theme_minimal(base_size = 12)
```

**Reveals:** Even after accounting for parts of Kapuas already known to be more fire-prone, a real, unexplained clustering signal remains -- growing continuously with distance and strongest at the far edge of the tested range (31.2km). Only at very short range (under \~183m) does the pattern roughly match what known geography alone would predict.

**Evidence:** The orange-shaded gap is a formally tested statistical departure (99 simulations) against a reference already accounting for Kapuas's known fire-prone areas, not a comparison against pure randomness. The convergence point near 183m and the widening gap beyond it were both verified against the underlying data, not estimated visually.

**Limitations:** This "known geography" reference is built from the same fire data being tested -- a circularity shared by Kinhom, Ginhom, and Finhom alike. Finhom confirms this clustering conclusion, but Ginhom shows significant regularity at short range instead -- a genuinely mixed result, detailed in Section 6.

### Temporal Profile of Fire Activity

```{r}
#| label: geovis-temporal
#| fig-width: 9
#| fig-height: 5
#| code-fold: true
#| code-summary: "Show plot redisplay code"

p_temporal
```

**Reveals:** Fire activity in Kapuas is heavily concentrated in the final six weeks of the observation window, not spread evenly across January-September, coinciding with the documented onset of Kalimantan's 2026 dry season.

**Evidence:** Weekly counts show a near step-function increase from early August, peaking at 623 detections (89.0 fires/day) in the week of 24 August. The apparent decline in the final, shorter week is shown -- after normalising for its 5-day length -- to be a plateau (52.2 fires/day) rather than continued decline, matching documented dry-season literature [@sulaiman2023peatland].

**Limitations:** Only one season is observed, with no multi-year baseline to confirm this is a typical rather than unusual year. The spatial analyses in this report pool the full window into one cross-section; the quiet/surge split this implies is examined separately in this report's Bonus/Value-Added component (Spatio-Temporal Extension), at reduced statistical power given the quiet period's small sample.

### Cross-Method Distance Comparison

```{r}
#| label: geovis-synthesis-chart
#| fig-width: 9
#| fig-height: 4
#| code-fold: true
#| code-summary: "Show plot redisplay code"

ggplot(synthesis_distances, aes(x = distance_m, y = reorder(label, distance_m),
                                 colour = type)) +
  geom_segment(aes(x = 100, xend = distance_m, yend = reorder(label, distance_m)),
               linewidth = 0.6) +
  geom_point(size = 4) +
  annotate("rect", xmin = 100, xmax = 182.99, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.4) +
  annotate("rect", xmin = 182.99, xmax = 40000, ymin = -Inf, ymax = Inf,
           fill = "orange", alpha = 0.08) +
  annotate("text", x = 130, y = 5.4, label = "Kinhom:\nexplained",
           size = 2.6, colour = "grey40", hjust = 0.5) +
  annotate("text", x = 8000, y = 5.4, label = "Kinhom: unexplained residual (other methods shown for scale)",
           size = 2.6, colour = "grey40", hjust = 0.5) +
  scale_x_log10(labels = scales::label_comma(suffix = "m")) +
  scale_colour_brewer(palette = "Dark2") +
labs(
  title = "Characteristic Distances Across Methods With a Clear Peak or Convergence Point",
  subtitle = "Log scale; shading marks Kinhom's own explained/residual split. Ginhom and Finhom excluded",
  x = "Distance (metres, log scale)", y = NULL, colour = NULL
) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

**Reveals:** The five methods shown here -- each with a clear peak or convergence point -- operate at genuinely different characteristic distances, spanning roughly 183m to 31.2km, almost three orders of magnitude, with Kinhom's own explained-versus-unexplained boundary providing the clearest single dividing line across that whole range.

**Evidence:** All five distances are extracted directly from the fitted spatstat objects in Sections 5-6, not estimated visually. The log scale is a deliberate choice, not a default: without it, the shortest distance (183m) would be visually indistinguishable from zero on the same axis as the longest (31.2km).

**Limitations:** Only Kinhom's two points are confirmed on a specific side of the explained/unexplained divide; G, F, and K/L are shown for scale only, testing against a different (flat CSR) reference. Ginhom and Finhom are excluded here -- neither has a clean peak or convergence distance to plot; see the cross-method table instead.

## Public-Safety Planning and Management Implications

This report's findings support two categories of implication: those directly evidenced, and those only motivated, pending further data.

**Evidence-supported implications.** The KDE surface and quadrat test jointly locate a precise, verified hotspot (114.146°E, 2.022°S, within a 561.9km² quadrat tile recording 412 detections) -- a concrete candidate for BPBD resource staging or monitoring infrastructure. Second-order analysis adds a distance dimension: clustering is strongest at ~12.4km (L-function), and Kinhom shows this is not fully explained by known fire-prone geography -- a real signal persists out to the full 31.2km tested range, supporting response-planning radii at multiple scales rather than a single fixed buffer. The temporal profile is equally actionable: activity concentrated overwhelmingly in the final six weeks, with a step-change onset in early August, supporting a seasonally front-loaded posture ahead of the documented dry-season window rather than reacting after activity has already surged.

### Illustrative Response-Radius Map

The multi-scale response-radius claim above is shown spatially below, using the same verified hotspot location and characteristic distances established in Sections 5-6 -- not a new analytical claim, but a visualisation of an existing one.

```{r}
#| label: safety-response-radius-map
#| fig-width: 10
#| fig-height: 10
#| code-fold: true
#| code-summary: "Show map code"

peak_sf_utm <- st_as_sf(data.frame(x = peak_row$x, y = peak_row$y),
                         coords = c("x", "y"), crs = 32750)

kinhom_gap_check <- Kinhom_kapuas_csr$obs - Kinhom_kapuas_csr$hi
widest_gap_r <- Kinhom_kapuas_csr$r[which.max(kinhom_gap_check)]

buffer_characteristic <- st_buffer(peak_sf_utm, dist = peak_r)
buffer_widest <- st_buffer(peak_sf_utm, dist = widest_gap_r)

kde_df_plot <- as.data.frame(kde_kapuas_terra_km2, xy = TRUE)
names(kde_df_plot)[3] <- "density"

bbox_kapuas <- st_bbox(kapuas_boundary_utm)
x_margin <- (bbox_kapuas["xmax"] - bbox_kapuas["xmin"]) * 0.45
label_x <- bbox_kapuas["xmax"] + x_margin * 0.35

anchor_characteristic <- data.frame(x = peak_row$x + peak_r, y = peak_row$y)
anchor_widest <- data.frame(x = peak_row$x + widest_gap_r, y = peak_row$y)

label_characteristic_pos <- data.frame(x = label_x, y = peak_row$y - 90000)
label_widest_pos <- data.frame(x = label_x, y = peak_row$y + 90000)

ggplot() +
  geom_raster(data = kde_df_plot, aes(x = x, y = y, fill = density)) +
  scale_fill_viridis_c(name = "Estimated fire density (per km²)",
                        trans = "sqrt", na.value = "white") +
  geom_sf(data = kapuas_boundary_utm, fill = NA, colour = "white", linewidth = 0.5) +
  geom_sf(data = buffer_widest, fill = NA, colour = "darkorange",
          linewidth = 0.9, linetype = "dashed") +
  geom_sf(data = buffer_characteristic, fill = NA, colour = "red3",
          linewidth = 0.9) +
  geom_sf(data = peak_sf_utm, colour = "white", size = 4.5) +
  geom_sf(data = peak_sf_utm, colour = "black", size = 3) +
  geom_segment(data = anchor_characteristic,
               aes(x = x, y = y, xend = label_characteristic_pos$x,
                   yend = label_characteristic_pos$y),
               colour = "red3", linewidth = 0.4,
               arrow = arrow(length = unit(0.2, "cm"), ends = "first")) +
  geom_segment(data = anchor_widest,
               aes(x = x, y = y, xend = label_widest_pos$x,
                   yend = label_widest_pos$y),
               colour = "darkorange", linewidth = 0.4,
               arrow = arrow(length = unit(0.2, "cm"), ends = "first")) +
  geom_label(data = label_characteristic_pos,
             aes(x = x, y = y, label = "12.4km\ncharacteristic\nclustering scale\n(L-function)"),
             fill = "white", colour = "red3", size = 2.8, label.size = 0.3,
             hjust = 0, label.padding = unit(0.3, "lines")) +
  geom_label(data = label_widest_pos,
             aes(x = x, y = y, label = "31.2km\nwidest\nunexplained\nresidual\n(Kinhom)"),
             fill = "white", colour = "darkorange4", size = 2.8, label.size = 0.3,
             hjust = 0, label.padding = unit(0.3, "lines")) +
  coord_sf(xlim = c(bbox_kapuas["xmin"], bbox_kapuas["xmax"] + x_margin),
           ylim = c(bbox_kapuas["ymin"], bbox_kapuas["ymax"]),
           expand = FALSE) +
  annotation_scale(location = "bl", width_hint = 0.2) +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(title = "Verified Hotspot with Analytically-Grounded Response Radii",
       subtitle = "Radii are established distances from Sections 5-6; the underlying surface is the fixed-bandwidth KDE")
```

### When Should Readiness Already Be in Place?

The weekly counts in Section 8 show *that* activity surged in August; a cumulative view answers the more operational question of *by what date* meaningful readiness should already exist.

```{r}
#| label: safety-cumulative-timing
#| fig-width: 9
#| fig-height: 5
#| code-fold: true
#| code-summary: "Show plot code"

cumulative_df <- kapuas_fires_sf |>
  st_drop_geometry() |>
  count(acq_date, name = "n_fires") |>
  arrange(acq_date) |>
  mutate(cum_n = cumsum(n_fires),
         cum_pct = cum_n / sum(n_fires) * 100)

pct_by_aug1 <- cumulative_df$cum_pct[max(which(cumulative_df$acq_date <= as.Date("2026-08-01")))]
date_90pct <- cumulative_df$acq_date[which(cumulative_df$cum_pct >= 90)[1]]

pct_by_aug1
date_90pct

ggplot(cumulative_df, aes(x = acq_date, y = cum_pct)) +
  geom_line(linewidth = 1, colour = "firebrick") +
  geom_vline(xintercept = as.Date("2026-08-01"), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = 90, linetype = "dotted", colour = "grey60") +
  annotate("text", x = as.Date("2026-08-01"), y = 5, label = "1 Aug",
           hjust = -0.1, size = 3, colour = "grey30") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(title = "Cumulative Share of the Season's Fires Over Time",
       subtitle = paste0("By 1 August, only ", round(pct_by_aug1, 1),
                          "% of the season's total fires had occurred"),
       x = NULL, y = "Cumulative % of season's fires") +
  theme_minimal(base_size = 12)
```

By 1 August -- the step-change onset -- only 3.19% of the season's total fires had occurred; 90% was not reached until 10 September, a single day before the observation window closes. This confirms both halves of the earlier claim: readiness needs to be in place well before the documented onset date, and activity was still building, not tapering, when this dataset ends.

**Implications requiring additional information.** Why clustering persists beyond what geography explains is not established here. Section 7 identified three plausible mechanisms for this clustering -- genuine fire-to-fire spread, a shared driver such as road/river accessibility (documented in the literature for Kapuas specifically [@thoha2019characteristics], but not causally tested by this analysis), and uneven satellite detection -- and this analysis cannot distinguish between them. Section 7 also found a countervailing signal at very short range (Ginhom): once known intensity is accounted for, nearest-neighbour spacing is more regular than expected, not more clustered -- plausibly linked to MODIS's own ~1km detection-pixel resolution rather than genuine fire behaviour, though this mechanism is untested. Land-use or road-network policy responses should not be justified as addressing a cause until a covariate-based model is actually built; this remains an undecided candidate for future work, not a completed analysis. Relatedly, BPBD facility locations were named as analytically relevant when the study window was defined, but no such distances were computed in this report -- any claim about current response-unit proximity to the hotspot would be unsupported speculation, not a finding.

A further caution applies throughout: intensity is not risk. A high-detection area may reflect more reliable satellite observation rather than genuinely elevated hazard, so resource-allocation decisions should treat this surface as a starting hypothesis for field verification, not a substitute for it. This analysis covers one season; whether 2026's timing is typical or an unusual anomaly cannot be determined without multi-year data, which matters for any commitment to permanent rather than seasonal infrastructure.

In summary, this analysis supports targeted, time-bound readiness around a specific, well-verified location and season. It does not, on its own, support causal land-use claims, fixed infrastructure siting, or conclusions about underlying fire risk without further field or covariate data.

## Limitations

This section consolidates limitations already identified at the point they arose in Sections 3-9, rather than introducing new ones; each is cross-referenced back to its full discussion.

**Data limitations.** Near-duplicate fire detections (23.4% of raw records) were collapsed to a single representative per location/day group using highest confidence as the tie-breaker (Near-duplicate detections) -- a defensible but not the only possible collapsing rule; a different rule could plausibly shift point counts slightly, though the underlying spatial pattern is unlikely to be sensitive to this choice given the scale of clustering observed. Fire detection itself is satellite-dependent: MODIS's ~1km pixel resolution and revisit geometry mean detection reliability varies with cloud cover, terrain accessibility, and overpass timing, so lower measured intensity in an area is not necessarily lower actual fire activity (Intensity is not risk) -- this caveat applies to every intensity- and clustering-based finding in this report, not only the first-order surface where it was first raised.

**Coordinate system and boundary geometry.** Kapuas straddles the UTM zone 49/50 boundary at 114°E; EPSG:32750 (zone 50S) was applied uniformly for simplicity, introducing minor positional distortion for the minority of points near the western edge (CRS Transformation and Kapuas Subsetting). The dissolved regency boundary also contains a topological sliver (0.1457km², confirmed to contain zero fire points) alongside the primary polygon -- a data-cleaning artefact rather than a genuine exclave, but one that trivially reduces the usable window area below Kapuas's true geodetic extent.

**First-order analysis.** Kernel density estimates are bandwidth-sensitive: `bw.ppl()` and `bw.diggle()` disagreed by a factor of two (1.113km vs. 0.550km), and fixed versus adaptive bandwidth produced visibly different surfaces (Kernel Density Estimation) -- the fixed-bandwidth result was retained as primary, with this disagreement documented rather than resolved. The quadrat test's extreme significance (p \< 2.2 x 10⁻¹⁶) is partly a function of sample size (1,849 points) rather than evidence of an unusually strong effect -- CSR was always going to be rejected at this scale, so the p-value's magnitude is not itself informative beyond confirming non-uniformity exists (Interpreting the quadrat test result).

**Second-order analysis.** All homogeneous tests (G, F, K/L) compare against flat CSR, which Section 5 already established does not describe Kapuas; `Kinhom`, `Ginhom`, and `Finhom` address this using an intensity surface estimated from the same data being tested -- a documented circularity affecting all three, not fully resolved by any of them (Assumptions and Limitations, Section 7). The G-function's reliable range is truncated to 1.22km by the border correction's shrinking sample at larger distances. `Ginhom`'s regularity finding was offered two candidate mechanisms (a MODIS pixel-resolution floor, or genuine behavioural spacing in land-clearing burns) but this analysis cannot distinguish between them (Alternative Explanations for the Observed Pattern).

**Temporal scope.** The full observation window is treated as one pooled cross-section in Sections 5-6 despite fire activity being heavily concentrated in the final six weeks (Temporal Study Window; Temporal Profile of Fire Activity); the Spatio-Temporal Extension examines the quiet/surge split separately, though at reduced statistical power given the quiet period's 58-point sample.

**Unresolved causal and policy questions.** Three candidate mechanisms for the persistent clustering signal (genuine fire-to-fire spread, a shared driver such as road/river accessibility, and uneven satellite detection) cannot be distinguished with fire-detection data alone; doing so would require a covariate-based model (distance to river, road, or BPBD facility) that was not built in this report (Alternative Explanations for the Observed Pattern). Relatedly, BPBD facility locations were identified as analytically relevant when the study window was defined (Spatial Study Window) but no such distances were actually computed -- any claim about current response-unit proximity to the hotspot remains unsupported. 

## References

*(auto-generated from `references.bib` via the YAML `bibliography` field)*
